% Glossary entries for the HPC book
% This file is \input from the main document after the preamble

% Glossary stance: categories are maintained clusters, not essences.
\renewcommand*{\glossarypreamble}{%
  \epigraph{The circle of the English language has a well-defined centre but no discernible circumference.}{James Murray, \textit{OED} Vol.~1 (1888), p.~xvii}
  \noindent \small Many of the entries in this glossary are themselves categories and so are defined in terms of mechanisms and diagnostics where relevant; examples illustrate use, not definitions. For page references, see the Subject Index (p.~\ref{idx:subject}).\par\bigskip}

% Micro-structure helpers (quiet, Bringhurst-style)
\newcommand{\glsdef}[1]{#1}
\newcommand{\glsmech}[1]{#1}
\newcommand{\glsdiag}[1]{#1}
\newcommand{\glsnote}[1]{#1}
\newcommand{\glsxsee}[1]{See also: #1}

% ===== Core framework terms =====

\newglossaryentry{compositecoupling}{
  name={composite coupling},
  text={composite coupling},
  description={\glsdef{A form--meaning pairing stabilized as a holistic gestalt by language use.} \glsmech{Maintained by cue redundancy, frequency entrenchment, and normative pressure rather than automatic biophysical constraints.} \glsdiag{Characteristic of linguistic constructions where habit constitutes rather than merely mediates the pairing.} \glsnote{Formerly termed `re-unified coupling'. Replaces `loose coupling' at the top of the zipper hierarchy.} \glsxsee{loose coupling, hard coupling, zipper.}}
}

\newglossaryentry{hardcoupling}{
  name={hard coupling},
  text={hard coupling},
  description={\glsdef{A rigid, highly systematic mapping between form, object, and interpretant.} \glsmech{Maintained by robust, low-level constraints like quantal articulatory regions and perceptual magnets.} \glsdiag{Sharp boundaries, resistance to perturbation, and minimal slippage during interpretation.} \glsnote{Most visible at the phoneme level.} \glsxsee{loose coupling, composite coupling, zipper.}}
}

\newglossaryentry{loosecoupling}{
  name={loose coupling},
  text={loose coupling},
  description={\glsdef{A mapping where form and meaning are connected but partially dissociable.} \glsmech{Sustained by mediating habits, context, and distributed social/cognitive mechanisms rather than brute transmission.} \glsdiag{Tolerates gradience, permits independent drift of the formal/semantic poles, and is field-relative.} \glsnote{Characteristic of grammatical categories like definiteness or lexical class.} \glsxsee{hard coupling, composite coupling, zipper.}}
}

\newglossaryentry{hpc}{
  name={homeostatic property cluster (HPC)},
  text={HPC},
  description={\glsdef{A category whose properties cluster because mechanisms maintain their co-occurrence rather than because of a shared essence.} \glsmech{Maintained by acquisition, entrenchment, alignment, transmission, and functional pressure.} \glsdiag{Passes both projectibility and homeostasis (perturbation sensitivity).} \glsnote{Membership can be graded and internally structured.} \glsxsee{projectibility, homeostasis, stabilizer, two-diagnostic test.}}
}

\newglossaryentry{projectibility}{
  name={projectibility},
  text={projectibility},
  description={\glsdef{A category's capacity to support induction from observed cases to unobserved ones.} \glsmech{It relies on stable correlations enforced by stabilizers.} \glsdiag{Predicts held-out data or generalises across contexts better than chance.} \glsnote{Field-relative: projectible for one purpose may fail for another.} \glsxsee{homeostasis, two-diagnostic test, field-relative projectibility.}}
}

\newglossaryentry{homeostasis}{
  name={homeostasis},
  text={homeostasis},
  description={\glsdef{Active return after perturbation; stability by self-correction rather than mere persistence.} \glsmech{Maintained by alignment, entrenchment, transmission, and other stabilizers across timescales.} \glsdiag{Perturbation sensitivity: weaken a mechanism and the clustering frays.} \glsnote{The ontological anchor for genuine kinds.} \glsxsee{stabilizer, perturbation sensitivity.}}
}

\newglossaryentry{entrenchment}{
  name={entrenchment},
  text={entrenchment},
  description={\glsdef{Strengthening of mental representations through repeated use.} \glsmech{Driven by frequency effects and chunking in memory.} \glsdiag{Resistance to analogical change; faster processing and higher accessibility.} \glsnote{Anchors category cores and stabilizes prototypes.} \glsxsee{chunking, prototype.}}
}

\newglossaryentry{essentialism}{
  name={essentialism},
  text={essentialism},
  description={\glsdef{The view that categories are defined by necessary and sufficient conditions.} \glsmech{It is grounded in definitional criteria rather than stabilizers.} \glsdiag{Breaks down when exceptions proliferate without principled repair.} \glsnote{The maintenance view rejects essentialism in favour of mechanism-based kinds.} \glsxsee{nominalism, prototype.}}
}

\newglossaryentry{stabilizer}{
  name={stabilizer},
  text={stabilizer},
  description={\glsdef{The functional role played by a mechanism in maintaining the clustering of a category.} \glsmech{Instances include acquisition, entrenchment, alignment, transmission, and functional pressure.} \glsdiag{Removing it changes the clustering in predictable ways.} \glsnote{Refers to the maintenance function rather than the causal structure itself \parencite{illari2012}.} \glsxsee{homeostasis, perturbation sensitivity, mechanism.}}
}

\newglossaryentry{alignment}{
  name={interactive alignment},
  text={alignment},
  description={\glsdef{Convergence of interlocutors' linguistic choices during interaction.} \glsmech{Realized through accommodation and repair in real-time conversation.} \glsdiag{Local convergence and resistance to innovations under misalignment.} \glsnote{A fast-timescale stabilizer.} \glsxsee{stabilizer, transmission.}}
}

\newglossaryentry{analogy}{
  name={analogy},
  text={analogy},
  description={\glsdef{Extension of a pattern to new cases based on perceived similarity.} \glsmech{Driven by schema abstraction and entrenchment in acquisition.} \glsdiag{Analogical leveling and productivity for novel forms.} \glsnote{A stabilizer that can also reshape categories.} \glsxsee{entrenchment, transmission.}}
}

\newglossaryentry{errorandrepair}{
  name={error and repair},
  text={error and repair},
  description={\glsdef{Interactive correction of production or interpretation errors in real time.} \glsmech{Self-monitoring, interlocutor feedback, and repair sequences.} \glsdiag{Convergence after misfires; reduced divergence over repeated interactions.} \glsnote{A fast stabilizer operating at discourse timescales.} \glsxsee{alignment, stabilizer.}}
}

\newglossaryentry{processingeconomy}{
  name={processing economy},
  text={processing economy},
  description={\glsdef{Pressure to minimize cognitive and articulatory cost in production and comprehension.} \glsmech{Frequency-driven facilitation and memory constraints.} \glsdiag{Shorter or more predictable forms are preferred and stabilize.} \glsnote{Interacts with social and functional pressures.} \glsxsee{entrenchment, chunking.}}
}

\newglossaryentry{socialindexing}{
  name={social indexing},
  text={social indexing},
  description={\glsdef{Association of linguistic variants with social identities, stances, or registers.} \glsmech{Sociolinguistic signaling reinforced by community norms.} \glsdiag{Variant use correlates with social groupings and style shifts.} \glsnote{A stabilizer that can protect arbitrary variants.} \glsxsee{prestige selection, transmission.}}
}

\newglossaryentry{indexicality}{
  name={indexicality},
  text={indexicality},
  description={\glsdef{The linkage between linguistic forms and social meanings or stances.} \glsmech{Maintained by sociolinguistic association and uptake.} \glsdiag{Variants signal identity, register, or attitude.} \glsnote{Often discussed as social indexing.} \glsxsee{social indexing, prestige selection.}}
}

\newglossaryentry{creativity}{
  name={creativity},
  text={creativity},
  description={\glsdef{The production of novel forms or uses beyond memorized patterns.} \glsmech{Driven by productive schemas and analogical extension.} \glsdiag{Innovations that test and reshape category boundaries.} \glsnote{A source of variation that can become stabilized.} \glsxsee{analogy, transmission.}}
}

\newglossaryentry{standardization}{
  name={standardization},
  text={standardization},
  description={\glsdef{Institutional codification and enforcement of a linguistic norm.} \glsmech{Education, style guides, and prescriptive feedback.} \glsdiag{Reduced variation and convergence toward a prestige standard.} \glsnote{Creates explicit feedback loops outside ordinary transmission.} \glsxsee{social indexing, transmission.}}
}

\newglossaryentry{prototype}{
  name={prototype},
  text={prototype},
  description={\glsdef{The most typical member of a category.} \glsmech{Shaped by entrenchment and frequency effects.} \glsdiag{Typicality effects and graded membership.} \glsnote{Descriptive of structure; requires a mechanism story to explain stability.} \glsxsee{entrenchment, real gradience.}}
}

\newglossaryentry{maintenanceview}{
  name={maintenance view},
  text={maintenance view},
  description={\glsdef{The claim that categories are real because they are maintained by stabilizing mechanisms.} \glsmech{Supported by a braid of mechanisms across timescales.} \glsdiag{Implies the two-diagnostic test: projectibility plus homeostasis.} \glsnote{Rejects both essentialism and nominalism.} \glsxsee{hpc, two-diagnostic test.}}
}

\newglossaryentry{metaoccam}{
  name={meta-Occam},
  text={meta-Occam},
  description={\glsdef{The principle that parsimony applies to mechanisms rather than to the objects they produce.} \glsmech{In evolutionary theory: natural selection is simple; organisms are not. In linguistics: stabilizers are tractable; categories are complex.} \glsdiag{A good framework requires fewer mechanism types, not fewer category types.} \glsnote{Krakauer's term; applied here to the HPC framework.} \glsxsee{stabilizer, homeostasis, hpc.}}
}

\newglossaryentry{class}{
  name={class},
  text={class},
  description={\glsdef{A grouping used for description or convenience without a commitment to shared mechanisms.} \glsmech{No stabilizer story required.} \glsdiag{May be projectible locally but need not pass the two-diagnostic test.} \glsnote{Useful for taxonomy; not necessarily a natural kind.} \glsxsee{category, fat class, thin class, comparative concept.}}
}

\newglossaryentry{category}{
  name={category},
  text={category},
  description={\glsdef{A cluster of properties used to organize linguistic phenomena.} \glsmech{When genuine, maintained by stabilizers; when merely convenient, it is a class.} \glsdiag{A category earns kind-status only if it passes projectibility and homeostasis.} \glsnote{The book's core question is which categories are kinds.} \glsxsee{kind, two-diagnostic test, projectibility, homeostasis.}}
}

\newglossaryentry{kind}{
  name={kind},
  text={kind},
  description={\glsdef{A category that supports induction because mechanisms maintain its clustering.} \glsmech{Stabilizers keep the profile coherent across speakers and time.} \glsdiag{Passes the two-diagnostic test.} \glsnote{In this framework, linguistic kinds are HPC kinds rather than definitional essences.} \glsxsee{hpc, copied kind, two-diagnostic test.}}
}

\newglossaryentry{value}{
  name={value},
  text={value},
  description={\glsdef{A unit's contribution when deployed; what it counts as in the system.} \glsmech{Value is relational, not intrinsic: for phonemes, contrastive identity; for morphemes, conventional meaning; for constructions, interpretive potential.} \glsdiag{Captures what is coupled to form at each level of linguistic structure.} \glsnote{Introduced in Chapter~\ref{ch:what-we-havent-been-asking}; developed in Chapter~\ref{ch:the-category-zipper}.} \glsxsee{coupling, form cluster, definiteness cluster.}}
}

\newglossaryentry{mechanism}{
  name={mechanism},
  text={mechanism},
  description={\glsdef{Entities and activities organized in such a way that they are responsible for a phenomenon.} \glsmech{Mechanisms include acquisition, alignment, entrenchment, and transmission.} \glsdiag{Identified by perturbation: weaken the process and the cluster frays.} \glsnote{Adopts the consensus definition of \textcite{illari2012}.} \glsxsee{stabilizer, homeostasis, perturbation sensitivity.}}
}

\newglossaryentry{habit}{
  name={habit},
  text={habit},
  description={\glsdef{A conditional disposition (would-be) that governs future behaviour; the interpretant a sign generates in its users.} \glsmech{Sustained by mechanisms (acquisition, entrenchment, alignment, transmission) but not reducible to them.} \glsdiag{Reliable projection to unencountered instances; degradation when sustaining mechanisms are removed.} \glsnote{Peirce's term (CP~5.491); extends beyond minds to streams, plants, and populations (CP~5.492). The book's non-reductive position: mechanisms sustain habits, but the habit outstrips any particular mechanism.} \glsxsee{interpretant, would-be, projectibility, stabilizer.}}
}

\newglossaryentry{interpretant}{
  name={interpretant},
  text={interpretant},
  description={\glsdef{The inferential habit a sign generates in its users; the disposition to act that constitutes what the sign means.} \glsmech{Sustained by the mechanisms that maintain the habit: acquisition, entrenchment, alignment.} \glsdiag{Reliable projection from sign to behaviour; different conditions generate different interpretants from the same sign.} \glsnote{Peirce's term (CP~5.491). Not a mental image or a gloss but a readiness to behave. The interpretant mediates between sign and object, making the relation genuinely predictive rather than merely correlational.} \glsxsee{habit, would-be, field-relative projectibility.}}
}

\newglossaryentry{wouldbe}{
  name={would-be},
  text={would-be},
  description={\glsdef{A conditional disposition that outruns any finite set of instances; the content of a habit.} \glsmech{Sustained by the mechanisms that maintain the habit; reliable only while mechanisms operate.} \glsdiag{The disposition holds for instances not yet encountered; removing mechanisms degrades it.} \glsnote{Peirce's term (CP~5.467): \enquote{no agglomeration of actual happenings can ever completely fill up the meaning of a `would-be'.} A category's meaning is what you \emph{would} expect, not what you have observed.} \glsxsee{habit, interpretant, projectibility.}}
}

\newglossaryentry{acquisition}{
  name={acquisition},
  text={acquisition},
  description={\glsdef{The learning process by which speakers internalise linguistic categories from input.} \glsmech{Distributional learning and cue integration across contexts.} \glsdiag{Convergence on shared categories despite variable input.} \glsnote{A core stabilizer and transmission bottleneck.} \glsxsee{transmission, entrenchment, stabilizer.}}
}

\newglossaryentry{transmission}{
  name={transmission},
  text={transmission},
  description={\glsdef{Intergenerational passage of linguistic patterns.} \glsmech{Iterated learning filters for stable, learnable variants.} \glsdiag{Convergence toward compressible systems under repeated learning.} \glsnote{A macro-stabilizer of categories and constructions.} \glsxsee{iterated transmission, acquisition.}}
}

\newglossaryentry{properfunction}{
  name={proper function},
  text={proper function},
  description={\glsdef{In Millikan's framework, the function an item has in virtue of its history of selection.} \glsmech{Secured by reproductive success of uses that perform the function.} \glsdiag{Explains why the form persists across generations.} \glsnote{Historical, not intentional or definitional.} \glsxsee{derived proper function, Normal conditions.}}
}

\newglossaryentry{derivedproperfunction}{
  name={derived proper function},
  text={derived proper function},
  description={\glsdef{A function acquired by recruitment beyond an item's original selected function.} \glsmech{Arises through exploitation of stable side effects of a form or construction.} \glsdiag{Depends on the proper function remaining intact elsewhere.} \glsnote{Parasitic rather than foundational.} \glsxsee{proper function, Normal conditions.}}
}

\newglossaryentry{deitality}{
  name={deitality},
  text={deitality},
  description={\glsdef{The morphosyntactic form cluster associated with English determiners, distinct from semantic definiteness.} \glsmech{Formed by grammaticalization of demonstratives plus distributional constraints.} \glsdiag{\mention{There}-resistance, partitive \mention{of}, and hosting behaviour converge.} \glsnote{Cross-cuts the definiteness cluster.} \glsxsee{definiteness cluster.}}
}

\newglossaryentry{fatclass}{
  name={fat class},
  text={fat class},
  description={\glsdef{A label that lumps distinct causal clusters into a single bin.} \glsmech{Produced by multiple unrelated stabilizers with no shared mechanism.} \glsdiag{Fails projectibility across subtypes.} \glsnote{Useful for pedagogy, not a natural kind.} \glsxsee{thin class, comparative concept.}}
}

\newglossaryentry{adverb}{
  name={adverb},
  text={adverb},
  description={\glsdef{A traditional residual category for modifiers that are not adjectives.} \glsmech{Maintained by multiple unrelated mechanisms across subtypes.} \glsdiag{Fails projectibility across manner, degree, and sentence adverbs.} \glsnote{A paradigmatic fat class; the remedy is decomposition.} \glsxsee{fat class.}}
}

\newglossaryentry{thinclass}{
  name={thin class},
  text={thin class},
  description={\glsdef{A pattern that is weakly maintained or not maintained at all.} \glsmech{Occurs when stabilizers are absent or too weak.} \glsdiag{Fails homeostasis; disappears under perturbation.} \glsnote{Often a byproduct of other mechanisms.} \glsxsee{fat class.}}
}

\newglossaryentry{inflationproblem}{
  name={inflation problem},
  text={inflation problem},
  description={\glsdef{The tendency to treat every stable pattern as an HPC kind.} \glsmech{Fuelled by overgeneralization from local stability or institutional reinforcement.} \glsdiag{Categories appear projectible until the homeostasis test is applied.} \glsnote{A methodological warning against false positives.} \glsxsee{two-diagnostic test, fat class, thin class.}}
}

\newglossaryentry{comparativeconcept}{
  name={comparative concept},
  text={comparative concept},
  description={\glsdef{An analyst's tool constructed for cross-linguistic comparison.} \glsmech{Not guaranteed to be supported by shared stabilizers across languages.} \glsdiag{Does not imply shared homeostasis or projectibility.} \glsnote{Useful without being a kind.} \glsxsee{field-relative projectibility.}}
}

\newglossaryentry{epistemickind}{
  name={epistemic kind},
  text={epistemic kind},
  description={\glsdef{A category that serves epistemic purposes without claiming a shared causal mechanism.} \glsmech{May be institutionally reinforced rather than mechanistically maintained.} \glsdiag{Can be stable without passing the two-diagnostic test.} \glsnote{Khalidi's term for useful but non-kind groupings.} \glsxsee{class, comparative concept.}}
}

\newglossaryentry{madagascarfallacy}{
  name={Madagascar fallacy},
  text={Madagascar fallacy},
  description={\glsdef{The mistake of treating an ecosystem as if it were a single kind.} \glsmech{It bundles mechanisms that maintain components rather than the whole.} \glsdiag{Fails homeostasis at the wrong grain.} \glsnote{A warning about scale in category analysis.} \glsxsee{two-diagnostic test.}}
}

% ===== HPC extensions and stability taxonomies (Ch 4) =====

\newglossaryentry{etiologicalkind}{
  name={etiological kind},
  text={etiological kind},
  description={\glsdef{A kind defined by its history of production or transmission.} \glsmech{Maintained by copying lineages with variation and selection.} \glsdiag{Similarity is explained by lineage rather than by shared essence.} \glsnote{Khalidi's term for history-defined natural kinds.} \glsxsee{copied kind, transmission.}}
}

\newglossaryentry{simplecausaltheory}{
  name={simple causal theory},
  text={simple causal theory},
  description={\glsdef{A view that kinds are anchored by primary properties that causally generate secondary properties.} \glsmech{One-way causal dependence rather than feedback or self-correction.} \glsdiag{Stable property dependencies without homeostatic return.} \glsnote{Contrasts with homeostatic accounts.} \glsxsee{stable property cluster, homeostasis.}}
}

\newglossaryentry{stablepropertycluster}{
  name={stable property cluster},
  text={stable property cluster},
  description={\glsdef{A cluster of co-occurring properties that remains stable over time or context.} \glsmech{Stability can arise from multiple stabilizers, not only homeostasis.} \glsdiag{Cliquish stability with projectible correlations.} \glsnote{Slater's broadened stability framework.} \glsxsee{cliquish stability, homeostasis.}}
}

\newglossaryentry{instancestability}{
  name={instance stability},
  text={instance stability},
  description={\glsdef{Stability of individual members retaining a cluster's properties over time.} \glsmech{Supported by entrenchment, selection, or institutional reinforcement.} \glsdiag{Members resist drift even under perturbation.} \glsnote{Distinct from cliquish stability.} \glsxsee{cliquish stability, entrenchment.}}
}

\newglossaryentry{symmetrycondition}{
  name={symmetry condition},
  text={symmetry condition},
  description={\glsdef{A sign-language constraint: if both hands move, they share handshape and mirror movement.} \glsmech{Motor planning and perceptual constraints on bimanual coordination.} \glsdiag{Cross-linguistic convergence in two-handed signs.} \glsnote{An example of convergent stability without borrowing.} \glsxsee{dominance condition.}}
}

\newglossaryentry{dominancecondition}{
  name={dominance condition},
  text={dominance condition},
  description={\glsdef{A sign-language constraint: if only one hand moves, the other uses an unmarked handshape.} \glsmech{Motor and perceptual constraints on bimanual articulation.} \glsdiag{Restricted non-dominant handshapes across sign languages.} \glsnote{Pairs with the symmetry condition.} \glsxsee{symmetry condition.}}
}

\newglossaryentry{species}{
  name={species},
  text={species},
  description={\glsdef{A biological grouping of populations maintained by overlapping mechanisms rather than a single essence.} \glsmech{Gene flow, shared selection pressures, and reproductive systems.} \glsdiag{Cohesion with boundary cases (ring species, hybrids).} \glsnote{A paradigm case for mechanism-based kinds.} \glsxsee{hpc, etiological kind.}}
}

% ===== Countability (Ch 9) =====

\newglossaryentry{countability}{
  name={countability},
  text={countability},
  description={\glsdef{The coupled system linking semantic individuation and morphosyntactic count marking.} \glsmech{Maintained by bidirectional inference between construal and form.} \glsdiag{Implicational hierarchy of count properties and robust projectibility.} \glsnote{An interface category with two partially distinct clusters.} \glsxsee{individuation cluster, count cluster, bidirectional inference.}}
}

\newglossaryentry{individuationcluster}{
  name={individuation cluster},
  text={individuation cluster},
  description={\glsdef{The semantic profile that enables discrete construal: boundedness, atomicity, enumerability, and homogeneity resistance.} \glsmech{Maintained by perceptual segmentation, object files, and cross-modal integration.} \glsdiag{Supports predictions about count morphosyntax.} \glsnote{Coupled to the count cluster to yield countability.} \glsxsee{count cluster, bidirectional inference.}}
}

\newglossaryentry{boundedness}{
  name={boundedness},
  text={boundedness},
  description={\glsdef{The property of having discrete edges that separate an entity from its environment.} \glsmech{Grounded in perceptual segmentation and boundary detection.} \glsdiag{Supports exact counting and telic construals.} \glsnote{One component of the individuation cluster.} \glsxsee{individuation cluster, atomicity.}}
}

\newglossaryentry{atomicity}{
  name={atomicity},
  text={atomicity},
  description={\glsdef{The availability of entities as discrete units for reference and tracking.} \glsmech{Maintained by object files and unit-based representation.} \glsdiag{Compatibility with numerals and singular reference.} \glsnote{One component of the individuation cluster.} \glsxsee{individuation cluster, boundedness.}}
}

\newglossaryentry{enumerability}{
  name={enumerability},
  text={enumerability},
  description={\glsdef{Compatibility with exact counting and cardinal quantification.} \glsmech{Depends on stable unitization in conceptual structure.} \glsdiag{Accepts numerals and exact quantifiers.} \glsnote{One component of the individuation cluster.} \glsxsee{individuation cluster, count cluster.}}
}

\newglossaryentry{homogeneityresistance}{
  name={homogeneity resistance},
  text={homogeneity resistance},
  description={\glsdef{The property that proper parts are not of the same kind as the whole.} \glsmech{Tied to construal of individuals rather than substance.} \glsdiag{Fails \enquote{half an N} tests for count nouns.} \glsnote{One component of the individuation cluster.} \glsxsee{individuation cluster, boundedness.}}
}

\newglossaryentry{countcluster}{
  name={count cluster},
  text={count cluster},
  description={\glsdef{The morphosyntactic properties that travel together in English count nouns (plural, cardinals, \textit{many}/\textit{few}, agreement).} \glsmech{Maintained by acquisition and entrenchment in count frames, plus alignment in discourse.} \glsdiag{Implicational hierarchy: tight properties imply loose ones.} \glsnote{Coupled to individuation by bidirectional inference.} \glsxsee{individuation cluster, bidirectional inference, functional anchoring.}}
}

\newglossaryentry{bidirectionalinference}{
  name={bidirectional inference},
  text={bidirectional inference},
  description={\glsdef{The mechanism coupling semantic construal to morphosyntactic form in comprehension and production.} \glsmech{Realized by inference from morphosyntax to construal and back.} \glsdiag{Mutual predictability of count properties across constructions.} \glsnote{The core homeostatic link in countability.} \glsxsee{count cluster, individuation cluster.}}
}

\newglossaryentry{functionalanchoring}{
  name={functional anchoring},
  text={functional anchoring},
  description={\glsdef{Lexical alternatives absorb pressure to regularise a quasi-count noun.} \glsmech{Implemented through division of labour in the lexicon.} \glsdiag{Intermediate classes persist without extending tight properties.} \glsnote{Explains stability of \textit{cattle}/\textit{police} and drift of \textit{data}.} \glsxsee{count cluster.}}
}

\newglossaryentry{massword}{
  name={mass-word},
  text={mass-word},
  description={\glsdef{Jespersen's term for nouns that resist counting and take mass quantifiers.} \glsmech{Maintained by mass construal and distributional patterns.} \glsdiag{Rejects numerals but accepts \textit{much}/\textit{little}.} \glsnote{Highlights grammatical packaging over ontology.} \glsxsee{countability, count cluster.}}
}

\newglossaryentry{quasicountnoun}{
  name={quasi-count noun},
  text={quasi-count noun},
  description={\glsdef{A noun that allows some count diagnostics but resists tight ones.} \glsmech{Stabilized by functional anchoring and lexeme competition.} \glsdiag{Accepts \textit{many} but resists low numerals and articles.} \glsnote{A stable intermediate in the countability system.} \glsxsee{functional anchoring, singulative.}}
}

\newglossaryentry{singulative}{
  name={singulative},
  text={singulative},
  description={\glsdef{A morphological marker deriving a single unit from a collective or mass base.} \glsmech{Grammaticalizes individuation within the count system.} \glsdiag{Enables numerals and tight count diagnostics.} \glsnote{Common in Welsh and Arabic.} \glsxsee{countability, quasi-count noun.}}
}

\newglossaryentry{classifier}{
  name={classifier},
  text={classifier},
  description={\glsdef{A morpheme that mediates numerals and nouns in classifier languages.} \glsmech{Encodes individuation or shape properties in the numeral phrase.} \glsdiag{Obligatory in counting contexts in languages like Mandarin.} \glsnote{Relocates countability coupling to the classifier system.} \glsxsee{countability, count cluster.}}
}

\newglossaryentry{chunking}{
  name={chunking},
  text={chunking},
  description={\glsdef{Storage of high-frequency sequences as units, bypassing compositional assembly.} \glsmech{Driven by frequency-based memory consolidation.} \glsdiag{Fluency and acceptability differences in entrenched frames.} \glsnote{Supports entrenchment and stability at the cluster core.} \glsxsee{entrenchment.}}
}

\newglossaryentry{basin}{
  name={basin (attractor)},
  text={basin},
  description={\glsdef{A dynamical-systems metaphor: an attractor region toward which category states are drawn.} \glsmech{Created by stabilizer-driven pull toward the cluster centre.} \glsdiag{Stable cores with tolerated peripheries.} \glsnote{Explains graded membership with sharp boundaries.} \glsxsee{dynamic discreteness, real gradience.}}
}

% ===== Definiteness (Ch 10) =====

\newglossaryentry{definiteness}{
  name={definiteness},
  text={definiteness},
  description={\glsdef{A semantic category of identifiability, familiarity, and uniqueness.} \glsmech{Maintained by discourse tracking, domain restriction, and Theory of Mind.} \glsdiag{Predicts anaphoric recoverability and discourse behaviour.} \glsnote{Distinct from the morphosyntactic form cluster.} \glsxsee{definiteness cluster, deitality, weak definite.}}
}

\newglossaryentry{definitenesscluster}{
  name={definiteness cluster},
  text={definiteness cluster},
  description={\glsdef{The semantic cluster of familiarity, uniqueness, and identifiability.} \glsmech{Maintained by discourse tracking, domain restriction, and Theory of Mind.} \glsdiag{Anaphoric recoverability and discourse behaviour.} \glsnote{Distinct from the morphosyntactic deitality cluster.} \glsxsee{deitality, Normal conditions.}}
}

\newglossaryentry{familiarity}{
  name={familiarity},
  text={familiarity},
  description={\glsdef{The property of being discourse-old or previously introduced.} \glsmech{Maintained by discourse tracking and common ground updates.} \glsdiag{Supports anaphoric uptake and given/new contrasts.} \glsnote{A core component of the definiteness cluster.} \glsxsee{definiteness cluster, anaphoric recoverability.}}
}

\newglossaryentry{uniqueness}{
  name={uniqueness},
  text={uniqueness},
  description={\glsdef{The property of having exactly one salient candidate in a domain.} \glsmech{Maintained by domain restriction and contextual narrowing.} \glsdiag{Infelicity when multiple candidates are equally salient.} \glsnote{A core component of the definiteness cluster.} \glsxsee{definiteness cluster, identifiability.}}
}

\newglossaryentry{identifiability}{
  name={identifiability},
  text={identifiability},
  description={\glsdef{The property that a hearer can pick out the intended referent.} \glsmech{Maintained by Theory of Mind and discourse modeling.} \glsdiag{First-mention definites are licensed when the description suffices for retrieval.} \glsnote{A core component of the definiteness cluster.} \glsxsee{definiteness cluster, familiarity.}}
}

\newglossaryentry{anaphoricrecoverability}{
  name={anaphoric recoverability},
  text={anaphoric recoverability},
  description={\glsdef{The capacity of a referent to be taken up by pronouns or anaphora.} \glsmech{Maintained by discourse coherence and tracking mechanisms.} \glsdiag{Natural antecedent status in subsequent discourse.} \glsnote{A downstream affordance of definiteness.} \glsxsee{familiarity, definiteness cluster.}}
}

\newglossaryentry{prestigeselection}{
  name={prestige selection},
  text={prestige selection},
  description={\glsdef{Spread of variants due to association with high-status speakers.} \glsmech{Social evaluation and identity signaling reinforce the variant.} \glsdiag{Prestige forms persist despite competing regularization pressures.} \glsnote{A social stabilizer for arbitrary variants.} \glsxsee{social indexing, transmission.}}
}

\newglossaryentry{formcluster}{
  name={form cluster},
  text={form cluster},
  description={\glsdef{The morphosyntactic bundle associated with English determiners (\mention{there}-resistance, partitive \mention{of}, hosting).} \glsmech{Grammaticalization of demonstratives and stabilized distributional constraints.} \glsdiag{Convergence of structural diagnostics across determiners.} \glsnote{Also called the deitality cluster.} \glsxsee{deitality, definiteness.}}
}

\newglossaryentry{weakdefinite}{
  name={weak definite},
  text={weak definite},
  description={\glsdef{A definite article use that lacks a unique or familiar referent (institutional/role frames).} \glsmech{A derived function exploiting stable form-cluster properties.} \glsdiag{Form-cluster behaviour without definiteness-cluster satisfaction.} \glsnote{Productive in stereotyped activity frames.} \glsxsee{deitality, definiteness, form cluster.}}
}

\newglossaryentry{normalconditions}{
  name={Normal conditions},
  text={Normal conditions},
  description={\glsdef{The circumstances under which a device performs its proper function.} \glsmech{Sustained by shared context and recoverable referents.} \glsdiag{Breakdown produces miscoordination; derived uses exploit stable side effects.} \glsnote{Frames proper vs. derived function.} \glsxsee{proper function, derived proper function.}}
}

% ===== General framework =====

\newglossaryentry{fieldrelativeprojectibility}{
  name={field-relative projectibility},
  text={field-relative projectibility},
  description={\glsdef{Projectibility indexed to analytic purpose or domain.} \glsmech{Depends on task-specific stabilizers.} \glsdiag{A category may project for one purpose and fail for another.} \glsnote{Explains cross-cutting categories.} \glsxsee{projectibility, crosscutting kinds.}}
}

\newglossaryentry{twodiagnostictest}{
  name={two-diagnostic test},
  text={two-diagnostic test},
  description={\glsdef{The criterion for genuine kinds: projectibility plus homeostasis.} \glsmech{Implemented by pairing the projectibility and homeostasis diagnostics.} \glsdiag{Pass both to count as an HPC; fail one to diagnose thin/fat/negative.} \glsnote{Operational rather than definitional.} \glsxsee{projectibility, homeostasis, perturbation sensitivity.}}
}

\newglossaryentry{noun}{
  name={noun},
  text={noun},
  description={\glsdef{A lexical category associated with nominals and argument structure.} \glsmech{Maintained by distributional cues, morphology, and acquisition biases.} \glsdiag{Cliquish stability across multiple diagnostics (determiners, number, agreement).} \glsnote{Cross-linguistic comparability requires mechanism mapping.} \glsxsee{verb, subject, copied kind.}}
}

\newglossaryentry{propernoun}{
  name={proper noun},
  text={proper noun},
  description={\glsdef{A syntactic category of nominals with distinctive distributional behaviour.} \glsmech{Maintained by morphosyntactic cues such as article resistance and agreement patterns.} \glsdiag{Distributional diagnostics cluster in specific constructions.} \glsnote{Distinct from the semantic category of proper names.} \glsxsee{proper name, noun, field-relative projectibility.}}
}

\newglossaryentry{propername}{
  name={proper name},
  text={proper name},
  description={\glsdef{A semantic category of expressions that directly refer to individuals.} \glsmech{Maintained by discourse tracking and stable reference to individuals.} \glsdiag{Rigid designation and referential opacity effects.} \glsnote{Cross-cuts the syntactic category of proper nouns.} \glsxsee{proper noun, definiteness cluster, field-relative projectibility.}}
}

\newglossaryentry{verb}{
  name={verb},
  text={verb},
  description={\glsdef{A lexical category associated with predication and event structure.} \glsmech{Maintained by distributional frames, inflectional paradigms, and acquisition.} \glsdiag{Convergent diagnostics (tense/aspect marking, argument structure).} \glsnote{Stability varies by language and construction.} \glsxsee{noun, aspect.}}
}

\newglossaryentry{adjective}{
  name={adjective},
  text={adjective},
  description={\glsdef{A lexical category associated with property attribution.} \glsmech{Maintained by distributional position, morphology, and modification patterns.} \glsdiag{Compatibility with attributive and predicative positions.} \glsnote{Cross-linguistic status is often comparative rather than universal.} \glsxsee{noun, verb, comparative concept.}}
}

\newglossaryentry{auxiliary}{
  name={auxiliary},
  text={auxiliary},
  description={\glsdef{A verbal category that contributes tense, aspect, or modality without full lexical content.} \glsmech{Typically arises through grammaticalization and distributional entrenchment.} \glsdiag{Inversion, \mention{do}-support, or restricted non-finite behaviour.} \glsnote{Boundary cases are common in diachronic change.} \glsxsee{verb, grammaticalization.}}
}

\newglossaryentry{subject}{
  name={subject},
  text={subject},
  description={\glsdef{A syntactic function cluster defined by converging behavioural and coding properties.} \glsmech{Agreement, word order, case marking, and discourse roles.} \glsdiag{Convergence of multiple diagnostics rather than a single definition.} \glsnote{Cross-linguistic identity is not guaranteed.} \glsxsee{comparative concept, category.}}
}

\newglossaryentry{voice}{
  name={voice},
  text={voice},
  description={\glsdef{A morphosyntactic system that maps argument roles to grammatical functions.} \glsmech{Maintained by constructional alternations and discourse pressures.} \glsdiag{Voice morphology and alternations in argument realization.} \glsnote{Cross-linguistic mechanisms vary widely.} \glsxsee{subject, category.}}
}

\newglossaryentry{anaphor}{
  name={anaphor},
  text={anaphor},
  description={\glsdef{An expression that requires a local antecedent for interpretation.} \glsmech{Maintained by binding constraints and acquisition of dependency patterns.} \glsdiag{Obligatory local binding in its domain.} \glsnote{A classic category in binding theory.} \glsxsee{subject, category.}}
}

\newglossaryentry{interrogativephrase}{
  name={interrogative phrase},
  text={interrogative phrase},
  description={\glsdef{A phrase that introduces a question variable (e.g., wh-phrases).} \glsmech{Licensed by interrogative constructions and dependency formation.} \glsdiag{Participates in wh-dependencies and question formation.} \glsnote{Structural diagnostics vary across languages.} \glsxsee{bounding node, category.}}
}

\newglossaryentry{boundingnode}{
  name={bounding node},
  text={bounding node},
  description={\glsdef{A structural node posited to delimit extraction domains in early generative syntax.} \glsmech{Constrains long-distance dependencies by structural configuration.} \glsdiag{Island effects and extraction failures beyond the node.} \glsnote{A theory-internal construct rather than a mechanism.} \glsxsee{interrogative phrase, anaphor.}}
}

\newglossaryentry{canonicalclause}{
  name={canonical clause},
  text={canonical clause},
  description={\glsdef{An idealized benchmark clause used for descriptive comparison.} \glsmech{A regimentation tool rather than a maintained kind.} \glsdiag{Core diagnostics converge most cleanly on canonical cases.} \glsnote{Useful for exposition, not ontologically privileged.} \glsxsee{class, comparative concept.}}
}

\newglossaryentry{animacy}{
  name={animacy},
  text={animacy},
  description={\glsdef{A semantic distinction between animate and inanimate entities.} \glsmech{Grounded in cognitive salience and agency detection.} \glsdiag{Grammatical contrasts like \textit{who} vs.\ \textit{what}.} \glsnote{Often cross-cuts lexical similarity classes.} \glsxsee{personhood cluster.}}
}

\newglossaryentry{factivity}{
  name={factivity},
  text={factivity},
  description={\glsdef{The property of predicates that presuppose the truth of their complements.} \glsmech{Lexical semantics and pragmatic inference patterns.} \glsdiag{Presupposition survives negation and questions.} \glsnote{An inferential grouping that cross-cuts similarity.} \glsxsee{category.}}
}

\newglossaryentry{copiedkind}{
  name={copied kind},
  text={copied kind},
  description={\glsdef{A category whose members are produced from each other or a common template.} \glsmech{Maintained by a transmission lineage with variation.} \glsdiag{Similarity explained by copying rather than essences.} \glsnote{Grammatical categories are copied kinds par excellence.} \glsxsee{replicator, interactor.}}
}

\newglossaryentry{unicept}{
  name={unicept},
  text={unicept},
  description={\glsdef{A concept that tracks sameness through multiple fallible methods (Millikan).} \glsmech{Sustained by convergent cues with no privileged diagnostic.} \glsdiag{Cliquish stability: some cues reliably indicate the cluster.} \glsnote{Matches multi-diagnostic category tracking.} \glsxsee{cliquish stability, two-diagnostic test.}}
}

\newglossaryentry{iteratedtransmission}{
  name={iterated transmission},
  text={iterated transmission},
  description={\glsdef{Structure emerging from multi-generational learning bottlenecks.} \glsmech{Produced by filtering for learnable variants.} \glsdiag{Convergence toward compressible/compositional systems.} \glsnote{A macro-stabilizer of linguistic structure.} \glsxsee{copied kind, entrenchment.}}
}

\newglossaryentry{perturbationsensitivity}{
  name={perturbation sensitivity},
  text={perturbation sensitivity},
  description={\glsdef{The diagnostic that a category frays when a stabilizer is weakened.} \glsmech{Observed when interventions target acquisition, alignment, or transmission.} \glsdiag{Predictable degradation patterns under perturbation.} \glsnote{Operationalizes homeostasis.} \glsxsee{homeostasis, stabilizer.}}
}

\newglossaryentry{cliquishstability}{
  name={cliquish stability},
  text={cliquish stability},
  description={\glsdef{Property-correlation: some cues reliably indicate the whole cluster.} \glsmech{Maintained by stabilizers that keep correlations intact despite drift.} \glsdiag{Knowing one property predicts others.} \glsnote{Supports induction without perfect instance stability.} \glsxsee{projectibility.}}
}

% ===== Discreteness and gradience (Ch 5) =====

\newglossaryentry{dynamicdiscreteness}{
  name={dynamic discreteness},
  text={dynamic discreteness},
  description={\glsdef{Sharp boundaries produced by mechanisms rather than definitions.} \glsmech{Maintained by stabilizers that keep tolerance thresholds over time.} \glsdiag{Sharp judgments within speakers, variance across populations.} \glsnote{Explains discreteness without essences.} \glsxsee{relative tolerance, real gradience.}}
}

\newglossaryentry{realgradience}{
  name={real gradience},
  text={real gradience},
  description={\glsdef{Gradient structure as meaningful distance from category centres.} \glsmech{Shaped by variable stabilizer strength across the basin.} \glsdiag{Typicality and marginality effects.} \glsnote{Gradience is signal, not noise.} \glsxsee{basin, dynamic discreteness.}}
}

\newglossaryentry{relativetolerance}{
  name={relative tolerance},
  text={relative tolerance},
  description={\glsdef{Membership depends on scale-relative tolerance of change.} \glsmech{Determined by decision thresholds indexed to context and magnitude.} \glsdiag{Small changes acceptable at one scale but not another.} \glsnote{Explains tolerance intuitions at category edges.} \glsxsee{dynamic discreteness.}}
}

% ===== Projectibility mechanisms (Ch 6) =====

\newglossaryentry{mechanisticdrift}{
  name={mechanistic drift},
  text={mechanistic drift},
  description={\glsdef{Mechanisms that maintain a category can differ from those that originally produced it.} \glsmech{Driven by diachronic reconfiguration of stabilizers.} \glsdiag{Mismatch between current cues and historical rationale.} \glsnote{Explains semantic drift under stable labels.} \glsxsee{grammaticalization.}}
}

\newglossaryentry{aspect}{
  name={aspect},
  text={aspect},
  description={\glsdef{A grammatical category describing how events are temporally and structurally construed.} \glsmech{Maintained by distributional cue structures and morphological marking.} \glsdiag{Projectible patterns may diverge from textbook semantic definitions.} \glsnote{A comparative concept whose language-specific mechanisms can drift.} \glsxsee{projectibility, mechanistic drift.}}
}

\newglossaryentry{npi}{
  name={negative polarity item (NPI)},
  text={NPI},
  description={\glsdef{An expression whose distribution is restricted to negative or downward-entailing contexts.} \glsmech{Multiple licensing mechanisms yield heterogeneous subtypes.} \glsdiag{Restricted distribution with distinct licensing profiles.} \glsnote{Often a distributional class rather than a single kind.} \glsxsee{class, category.}}
}

\newglossaryentry{perfective}{
  name={perfective},
  text={perfective},
  description={\glsdef{An aspectual category presenting events as bounded or completed.} \glsmech{Maintained by morphological marking and cue–outcome associations.} \glsdiag{Distributional asymmetries across tense frames.} \glsnote{Highlights the gap between semantic definition and usage.} \glsxsee{aspect, boundedness.}}
}

\newglossaryentry{grueproblem}{
  name={grue problem},
  text={grue problem},
  description={\glsdef{Goodman's riddle: why do we project \enquote{green} but not \enquote{grue}?} \glsmech{Explained by the fact that only mechanism-grounded properties are projectible.} \glsdiag{Gerrymandered predicates fail induction.} \glsnote{Applied to language, pseudo-categories lack mechanisms.} \glsxsee{projectibility.}}
}

\newglossaryentry{crosscuttingkinds}{
  name={crosscutting kinds},
  text={crosscutting kinds},
  description={\glsdef{Overlapping causal networks where the same entity belongs to multiple kinds.} \glsmech{Maintained by distinct stabilizers for each network.} \glsdiag{Different inductions for different purposes.} \glsnote{Cross-cutting is expected under field-relative projectibility.} \glsxsee{field-relative projectibility.}}
}

% ===== Essentialism and its alternatives (Ch 1–3) =====

\newglossaryentry{nominalism}{
  name={nominalism},
  text={nominalism},
  description={\glsdef{The view that categories are convenient labels without commitment to natural kinds.} \glsmech{Requires no stabilizer story.} \glsdiag{No homeostasis or projectibility demanded.} \glsnote{Rejected by the maintenance view.} \glsxsee{essentialism, comparative concept.}}
}

\newglossaryentry{grammaticalization}{
  name={grammaticalization},
  text={grammaticalization},
  description={\glsdef{Diachronic shift from lexical item to grammatical marker.} \glsmech{Driven by frequency, reduction, reanalysis, and semantic bleaching.} \glsdiag{Layering and distributional expansion over time.} \glsnote{Creates form–meaning decouplings that later stabilize.} \glsxsee{mechanistic drift.}}
}

\newglossaryentry{replicator}{
  name={replicator},
  text={replicator},
  description={\glsdef{An entity that passes copies of itself through time with variation (Hull).} \glsmech{Sustained by transmission with differential survival.} \glsdiag{Selection on variants across generations.} \glsnote{Supports evolutionary framing of language change.} \glsxsee{interactor, copied kind.}}
}

\newglossaryentry{interactor}{
  name={interactor},
  text={interactor},
  description={\glsdef{An entity that interacts with its environment in ways that cause differential replication (Hull).} \glsmech{Realized in communicative success and failure.} \glsdiag{Selection pressure visible in usage patterns.} \glsnote{Pairs with replicator in an evolutionary account.} \glsxsee{replicator.}}
}

\newglossaryentry{verblessclause}{
  name={verbless clause},
  text={verbless clause},
  description={\glsdef{A clause-like construction with subject–predicate structure but no verb in the predicate.} \glsmech{Arises under competing diagnostics (predication vs. VP-headedness).} \glsdiag{Shows decoupling of form and function.} \glsnote{Illustrates pressure against essentialist definitions.} \glsxsee{two-diagnostic test.}}
}

\newglossaryentry{nonfiniteclause}{
  name={non-finite clause},
  text={non-finite clause},
  description={\glsdef{A residual category for clauses lacking primary tense inflection.} \glsmech{No single shared mechanism maintains the set.} \glsdiag{Fails both projectibility and homeostasis across subtypes.} \glsnote{A negative class; best replaced by positive constructions.} \glsxsee{fat class, category.}}
}

\newglossaryentry{opentexture}{
  name={open texture},
  text={open texture},
  description={\glsdef{Hart's term for the penumbral region where rules underdetermine outcomes.} \glsmech{Emerges when stabilizers are weak or in competition.} \glsdiag{Persistent variability and institutional repair.} \glsnote{Grammar has no Supreme Court; resolution is emergent.} \glsxsee{dynamic discreteness, mechanistic drift.}}
}

% ===== Information structure and lexical categories (Ch 11) =====

\newglossaryentry{bci}{
  name={Backgrounded Constituent Infelicity},
  text={Backgrounded Constituent Infelicity},
  description={\glsdef{The constraint that backgrounded constituents resist focus-sensitive operations.} \glsmech{Information-structural requirements on focus particles.} \glsdiag{Infelicity when a focus operator targets backgrounded material.} \glsnote{A diagnostic from focus-particle research.} \glsxsee{category.}}
}

% ===== Gender and pro-forms (Ch 12) =====

\newglossaryentry{gender}{
  name={gender},
  text={gender},
  description={\glsdef{A system of grammatically relevant contrasts over referents or nouns.} \glsmech{Maintained by agreement systems or pro-form inventories.} \glsdiag{Consistent selection of gender-marked items in anaphora or agreement.} \glsnote{Can be referential or noun-class based.} \glsxsee{personhood cluster, pro-form inventory.}}
}

\newglossaryentry{epicene}{
  name={epicene},
  text={epicene},
  description={\glsdef{A gender category unmarked for sex within the personal domain.} \glsmech{Maintained by personhood-based construal rather than sex marking.} \glsdiag{Compatible with any sexed referent.} \glsnote{Common in pronominal systems.} \glsxsee{gender, personhood cluster.}}
}

\newglossaryentry{designatum}{
  name={designatum},
  text={designatum},
  description={\glsdef{The entity as conceptualized by the speaker in context.} \glsmech{Constraining inference from meaning to form in reference.} \glsdiag{Pro-form choice tracks construal over antecedent form.} \glsnote{Distinct from antecedent and referent.} \glsxsee{gender, pro-form inventory.}}
}

\newglossaryentry{semantictransparency}{
  name={semantic transparency},
  text={semantic transparency},
  description={\glsdef{The predictability of form–meaning mappings in a system.} \glsmech{Stabilized by consistent coupling between semantic and morphosyntactic cues.} \glsdiag{Low mismatch rates and easier acquisition.} \glsnote{Supports tight coupling between clusters.} \glsxsee{pro-form inventory, countability.}}
}

\newglossaryentry{personhoodcluster}{
  name={personhood cluster},
  text={personhood cluster},
  description={\glsdef{The semantic properties that make a referent construable as a person.} \glsmech{Grounded in Theory of Mind and social cognition.} \glsdiag{Triggers personal pro-forms (\textit{who}, \textit{he/she}) over non-personal ones.} \glsnote{A semantic anchor for gender systems.} \glsxsee{gender, animacy.}}
}

\newglossaryentry{proforminventory}{
  name={pro-form inventory},
  text={pro-form inventory},
  description={\glsdef{The set of pro-forms and their distributional constraints in a language.} \glsmech{Maintained by acquisition and usage-based entrenchment.} \glsdiag{Stable patterns in wh-forms and pronominal selection.} \glsnote{A lexico-grammatical cluster in pro-form gender.} \glsxsee{gender, designatum.}}
}

% ===== Grammaticality (Ch 14) =====

\newglossaryentry{grammaticality}{
  name={grammaticality},
  text={grammaticality},
  description={\glsdef{The maintained coupling between morphosyntactic form and structural meaning.} \glsmech{Maintained by entrenchment, acquisition, alignment, and the feeling of ungrammaticality.} \glsdiag{A sentence is grammatical when it sits within the basin of attraction for its form--value pairing.} \glsnote{Not a binary switch but an HPC in its own right.} \glsxsee{acceptability, structural meaning, zipper.}}
}

\newglossaryentry{acceptability}{
  name={acceptability},
  text={acceptability},
  description={\glsdef{What the detector reports; the gradient feeling that a sentence is or is not well-formed.} \glsmech{Shaped by probability, processing ease, and entrenchment.} \glsdiag{Noisy readout of grammaticality; can mismatch the underlying coupling.} \glsnote{Gradient even when grammaticality is discrete.} \glsxsee{grammaticality, grammaticality illusion.}}
}

\newglossaryentry{structuralmeaning}{
  name={structural meaning},
  text={structural meaning},
  description={\glsdef{The contribution of grammar before lexical content fills the slots: modification, argument structure, tense, clause type, information structure.} \glsmech{Maintained by the same stabilizers as grammaticality.} \glsdiag{Coupled to morphosyntactic form in the grammaticality zipper.} \glsnote{Subcompositional; the instruction manual, not the assembled product.} \glsxsee{grammaticality, value.}}
}

\newglossaryentry{grammaticalityillusion}{
  name={grammaticality illusion},
  text={grammaticality illusion},
  description={\glsdef{A case where the feeling of (un)grammaticality mismatches the underlying structure.} \glsmech{Produced by processing ease (garden paths) or semantic distractors (Escher sentences).} \glsdiag{Can dissociate detector output from form--value coupling.} \glsnote{Parallel to perceptual illusions.} \glsxsee{acceptability, grammaticality.}}
}

\newglossaryentry{zipper}{
  name={zipper},
  text={zipper},
  description={\glsdef{The Chapter 14 metaphor for form--value coupling: morphosyntax and structural meaning as interlocking teeth.} \glsmech{Maintained by acquisition, entrenchment, alignment, and repair.} \glsdiag{Grammatical well-formedness corresponds to successful closure; mismatch corresponds to failed meshing.} \glsnote{A relational metaphor for coupling, not a separate mechanism.} \glsxsee{grammaticality, structural meaning, detector.}}
}

\newglossaryentry{detector}{
  name={detector},
  text={detector},
  description={\glsdef{The metacognitive readout that reports felt well-formedness.} \glsmech{Shaped by probabilistic expectations, processing ease, and entrenchment.} \glsdiag{Tracks grammaticality imperfectly and supports both accurate judgments and illusions.} \glsnote{In Chapter 14 this is the source of acceptability judgments.} \glsxsee{acceptability, grammaticality illusion, selective fallibility.}}
}

\newglossaryentry{comparativeillusion}{
  name={comparative illusion},
  text={comparative illusion},
  description={\glsdef{An acceptability illusion where locally plausible comparative chunks mask a global compositional failure.} \glsmech{Arises from chunkwise parsing and shallow interpretation under processing pressure.} \glsdiag{Sentences feel acceptable despite lacking a coherent comparison interpretation.} \glsnote{Often discussed with \enquote{Escher sentences}.} \glsxsee{Escher sentence, grammaticality illusion.}}
}

\newglossaryentry{eschersentence}{
  name={Escher sentence},
  text={Escher sentence},
  description={\glsdef{A sentence that seems locally grammatical but resists globally coherent interpretation.} \glsmech{Produced by locally licit structure plus global interpretive incompatibility.} \glsdiag{Strong acceptability with downstream interpretation failure.} \glsnote{A hallmark case of comparative illusion.} \glsxsee{comparative illusion, grammaticality illusion.}}
}

\newglossaryentry{selectivefallibility}{
  name={selective fallibility},
  text={selective fallibility},
  description={\glsdef{Systematic asymmetry in parser reliability: robust for some constraints, vulnerable for others.} \glsmech{Driven by cue timing, retrieval limits, and processing heuristics.} \glsdiag{Stable error profiles such as agreement attraction and garden-path susceptibility.} \glsnote{Supports treating grammaticality judgments as detector outputs rather than oracle verdicts.} \glsxsee{detector, grammaticality illusion.}}
}

% ===== Social stabilization (Ch 15) =====

\newglossaryentry{doublecopula}{
  name={double copula},
  text={double copula},
  description={\glsdef{The construction with repeated copula material (\eg\ \mention{The thing is is ...}).} \glsmech{Maintained in specific discourse contexts by planning dynamics, alignment, and local norming.} \glsdiag{Stable distribution in spontaneous speech but strong penalty in formal edited registers.} \glsnote{Used in Chapter 15 as a running case for socially conditioned grammatical stability.} \glsxsee{register, conditioned probability, mixed bin.}}
}

\newglossaryentry{culturalattractor}{
  name={cultural attractor},
  text={cultural attractor},
  description={\glsdef{A recurrent form toward which cultural variants converge despite noisy transmission.} \glsmech{Biases in production, uptake, and evaluation prune variation over repeated iterations.} \glsdiag{Robust recurrence of specific variants under changing token-level realizations.} \glsnote{From \textcite{sperber1996}; applied in Chapter 15 to linguistic stability.} \glsxsee{transmission, iterated transmission, maintained kind.}}
}

\newglossaryentry{mixedbin}{
  name={mixed bin},
  text={mixed bin},
  description={\glsdef{The modeling error created by pooling outputs from distinct maintenance regimes into one aggregate distribution.} \glsmech{Inflates apparent variance by averaging across latent subpopulations.} \glsdiag{Loose-looking marginal distributions that resolve into tight conditioned clusters.} \glsnote{Chapter 15's bolt-factory metaphor for sociolinguistic mixture structure.} \glsxsee{conditioned probability, unconditioned probability, dialect, register.}}
}

\newglossaryentry{conditionedprobability}{
  name={conditioned probability},
  text={conditioned probability},
  description={\glsdef{A probability distribution computed relative to source variables (\eg\ dialect, register, community).} \glsmech{Captures structure regenerated within a specific maintenance regime.} \glsdiag{Yields tighter, more projectible patterns than pooled marginals.} \glsnote{Central to the Chapter 15 argument that social variables are conditioning structure, not noise.} \glsxsee{unconditioned probability, mixed bin.}}
}

\newglossaryentry{unconditionedprobability}{
  name={unconditioned probability},
  text={unconditioned probability},
  description={\glsdef{A pooled probability distribution marginalizing over latent source variables.} \glsmech{Combines outputs from distinct stabilized systems into one aggregate.} \glsdiag{Can obscure homeostatic structure by blending multiple tight regimes.} \glsnote{Useful descriptively, but often misleading for mechanism diagnosis.} \glsxsee{conditioned probability, mixed bin.}}
}

\newglossaryentry{maintainedkind}{
  name={maintained kind},
  text={maintained kind},
  description={\glsdef{A kind whose stability is actively regenerated by cognitive and social mechanisms.} \glsmech{Sustained by interacting stabilizers across speakers, contexts, and timescales.} \glsdiag{Persists under perturbation and supports scoped induction because maintenance work continues.} \glsnote{Chapter 15 emphasizes the social layer of maintenance.} \glsxsee{hpc, stabilizer, maintenance view.}}
}

\newglossaryentry{dialect}{
  name={dialect},
  text={dialect},
  description={\glsdef{A relatively stable speaker-linked variety (\enquote{variety according to user}).} \glsmech{Maintained by network transmission, identity signaling, and community norming.} \glsdiag{Durable distributional profiles across contexts for a speech community or lineage.} \glsnote{In Chapter 15, dialect functions as a persistent conditioning variable.} \glsxsee{register, discourse community, conditioned probability.}}
}

\newglossaryentry{register}{
  name={register},
  text={register},
  description={\glsdef{A situation-linked variety (\enquote{variety according to use}) realized as reweighting of linguistic options.} \glsmech{Maintained by gatekeeping, genre expectations, and recurrent context-specific feedback.} \glsdiag{Systematic shifts in lexical/syntactic distributions across activity types.} \glsnote{Chapter 15 treats register as a mode switch rather than a separate grammar.} \glsxsee{field, tenor, mode, dialect.}}
}

\newglossaryentry{field}{
  name={field},
  text={field},
  description={\glsdef{Halliday's register dimension for what activity or topic is being enacted.} \glsmech{Shapes lexical and structural expectations through task demands.} \glsdiag{Topic- and activity-sensitive clustering of forms.} \glsnote{One component of field--tenor--mode register description.} \glsxsee{register, tenor, mode.}}
}

\newglossaryentry{tenor}{
  name={tenor},
  text={tenor},
  description={\glsdef{Halliday's register dimension for participant relations and social stance.} \glsmech{Maintained by role expectations and interpersonal norming.} \glsdiag{Systematic modulation of politeness, formality, and alignment signals.} \glsnote{One component of field--tenor--mode register description.} \glsxsee{register, field, mode.}}
}

\newglossaryentry{mode}{
  name={mode},
  text={mode},
  description={\glsdef{Halliday's register dimension for the communicative channel and discourse organization.} \glsmech{Constrains form choices via medium-specific production/comprehension affordances.} \glsdiag{Predictable contrasts across spoken, written, and mixed channels.} \glsnote{One component of field--tenor--mode register description.} \glsxsee{register, field, tenor.}}
}

\newglossaryentry{discoursecommunity}{
  name={discourse community},
  text={discourse community},
  description={\glsdef{A community defined by shared communicative goals, genres, and evaluative norms.} \glsmech{Maintained by apprenticeship, gatekeeping, citation/review practice, and reward structure.} \glsdiag{Stable, source-attributable conventions despite membership turnover.} \glsnote{In Chapter 15 this is modeled as a latent mixture component.} \glsxsee{dialect, register, interactive kind.}}
}

\newglossaryentry{communicativesituation}{
  name={Communicative Situation (com-sit)},
  text={com-sit},
  description={\glsdef{A recurrent interactional setting pairing specific interlocutors, contexts, and expectations.} \glsmech{Children acquire language by tracking such situation-bound distributions before abstracting larger variety labels.} \glsdiag{Early learning is partitioned by source-conditioned routines, then generalized upward.} \glsnote{From \textcite{wiese2023}; used in Chapter 15 for acquisition as source inference.} \glsxsee{conditioned probability, dialect, register.}}
}

\newglossaryentry{generalcategoryfallacy}{
  name={general category fallacy},
  text={general category fallacy},
  description={\glsdef{The mistake of treating broad comparative labels as if they were single language-internal categories.} \glsmech{Arises when analysts ignore conditioning structure and mechanism heterogeneity.} \glsdiag{Overgeneralized claims that collapse distinct stabilized systems into one.} \glsnote{Used in Chapter 15 after \textcite{haspelmath2018}.} \glsxsee{comparative concept, mixed bin, field-relative projectibility.}}
}

\newglossaryentry{loopingkind}{
  name={looping kind},
  text={looping kind},
  description={\glsdef{A kind whose classification feeds back on members' behavior, thereby reshaping the kind itself.} \glsmech{Labeling, uptake, sanction, and self-adjustment create reflexive stabilization loops.} \glsdiag{Category labels alter trajectories of the very items/people classified.} \glsnote{From \textcite{hacking1999}; central to Chapter 15's social feedback model.} \glsxsee{interactive kind, discourse community, social meaning.}}
}

\newglossaryentry{interactivekind}{
  name={interactive kind},
  text={interactive kind},
  description={\glsdef{A socially maintained kind whose members adjust under awareness of the category and its norms.} \glsmech{Sustained by feedback from evaluation, correction, and coordination incentives.} \glsdiag{Conformity pressures and norm-sensitive adaptation keep the cluster coherent.} \glsnote{A linguistic extension of looping-kind dynamics.} \glsxsee{looping kind, maintained kind.}}
}

\newglossaryentry{socialmeaning}{
  name={social meaning},
  text={social meaning},
  description={\glsdef{The inferred social signal of a linguistic form (stance, identity, affiliation, context).} \glsmech{Computed by inverse inference from form to likely source conditions.} \glsdiag{Stable form-to-social inferences within communities and registers.} \glsnote{Chapter 15 formalizes this as the inverse of grammar's conditioning function.} \glsxsee{indexicality, indexical field, register.}}
}

\newglossaryentry{indexicalfield}{
  name={indexical field},
  text={indexical field},
  description={\glsdef{A structured distribution of socially interpretable stances associated with a form.} \glsmech{Maintained by repeated community uptake and context-sensitive interpretation.} \glsdiag{Forms cue a range (not a single point) of socially meaningful readings.} \glsnote{Following \textcite{eckert2008indexical}, used in Chapter 15.} \glsxsee{social meaning, indexicality.}}
}

\newglossaryentry{thickterm}{
  name={thick term},
  text={thick term},
  description={\glsdef{A term that combines descriptive categorization with evaluative stance.} \glsmech{Its usage is stabilized by both referential practice and norm enforcement.} \glsdiag{Meaning shifts track social evaluation as well as denotation.} \glsnote{Chapter 15 uses this for slur-label dynamics in public discourse.} \glsxsee{social meaning, looping kind.}}
}

% ===== Coverage pass: additional chapter terms =====

\newglossaryentry{agentbasedmodel}{
  name={agent-based model},
  text={agent-based model},
  description={\glsdef{A simulation framework where many simple agents interact so mechanism-level assumptions generate macro-patterns.} \glsmech{Stability or drift emerges from explicit local update rules rather than top-down forcing.} \glsdiag{Useful for perturbation and ablation tests of a maintenance hypothesis.} \glsnote{In this book, ABMs are mechanism sketches, not calibrated empirical models.} \glsxsee{homeostasis, perturbation sensitivity, mechanism.}}
}

\newglossaryentry{particle}{
  name={particle},
  text={particle},
  description={\glsdef{A residual lexical class for small function-like items that resist cleaner category assignments in a language-specific grammar.} \glsmech{Often maintained by construction-specific distributions and lexical idiosyncrasy rather than a single unified profile.} \glsdiag{Typically weak projectibility outside narrow constructional contexts.} \glsnote{Frequently a descriptive class rather than a mechanism-backed category.} \glsxsee{class, category, construction.}}
}

\newglossaryentry{infinitivalcomplement}{
  name={infinitival complement},
  text={infinitival complement},
  description={\glsdef{A non-finite clause functioning as the complement of a selecting head.} \glsmech{Licensed by head-specific selectional and structural constraints.} \glsdiag{Distribution tracks complement-taking profiles of verbs/adjectives/nouns.} \glsnote{A constructional slot with language-specific restrictions.} \glsxsee{non-finite clause, construction, argument structure.}}
}

\newglossaryentry{perfectconstruction}{
  name={perfect construction},
  text={perfect construction},
  description={\glsdef{A construction encoding a prior event's current relevance from a reference point.} \glsmech{Maintained by stable form--value coupling between perfect morphology and discourse-temporal interpretation.} \glsdiag{Supports recurrent inferences about result-state, experience, or continuation depending on language and context.} \glsnote{Distinct from simple past and from progressive aspect.} \glsxsee{aspect, progressive construction, compositionality.}}
}

\newglossaryentry{progressiveconstruction}{
  name={progressive construction},
  text={progressive construction},
  description={\glsdef{A construction marking an event as ongoing, unfolding, or internally viewed.} \glsmech{Sustained by recurrent mappings between progressive morphology and viewpoint-aspect interpretation.} \glsdiag{Contrasts systematically with non-progressive aspect choices in matched contexts.} \glsnote{Language-specific profiles can include futurate and habitual extensions.} \glsxsee{aspect, perfect construction, construction.}}
}

\newglossaryentry{lexicalcategory}{
  name={lexical category},
  text={lexical category},
  description={\glsdef{A part-of-speech-level cluster defined by distributional, morphological, and semantic tendencies maintained within a grammar.} \glsmech{Stabilized by acquisition, entrenchment, and constructional reuse.} \glsdiag{Projectible behavior across slots and constructions, with expected boundary cases.} \glsnote{Core cases include noun, verb, adjective, and adposition where supported.} \glsxsee{noun, verb, adjective, category.}}
}

\newglossaryentry{proform}{
  name={pro-form},
  text={pro-form},
  description={\glsdef{An expression interpreted through recovery from context or discourse rather than full lexical specification in place.} \glsmech{Maintained by anaphoric tracking, construal routines, and distributional constraints.} \glsdiag{Interpretation depends on recoverability from antecedent, designatum, or situation model.} \glsnote{Includes pronominal and related substitution-like forms without requiring a single category essence.} \glsxsee{anaphor, antecedent, designatum, pro-form inventory.}}
}

\newglossaryentry{irewords}{
  name={IRE words},
  text={IRE words},
  description={\glsdef{The interrogative/relative/exclamative lexical family treated as a cross-cutting functional cluster.} \glsmech{Maintained by partially shared paradigms and distributional alignments across clause types.} \glsdiag{Substantial overlap of forms and constraints across interrogative, relative, and exclamative uses.} \glsnote{A broader family than colloquial `question words'.} \glsxsee{interrogative, interrogative phrase, lexical category.}}
}

\newglossaryentry{referent}{
  name={referent},
  text={referent},
  description={\glsdef{The entity in the world or discourse model that an expression is used to pick out.} \glsmech{Resolved by lexical cues, grammar, context, and interactional inference.} \glsdiag{Successful reference yields coordinated uptake across interlocutors.} \glsnote{Distinguished from antecedent form and from designatum construal.} \glsxsee{antecedent, designatum, social meaning.}}
}

\newglossaryentry{antecedent}{
  name={antecedent},
  text={antecedent},
  description={\glsdef{A prior linguistic expression that supplies features or recoverability for a dependent form.} \glsmech{Maintained by anaphoric linking constraints and discourse memory.} \glsdiag{Constrains agreement, binding, and recoverability in dependent expressions.} \glsnote{Need not be identical to the intended designatum.} \glsxsee{anaphor, referent, designatum.}}
}

\newglossaryentry{adposition}{
  name={adposition},
  text={adposition},
  description={\glsdef{A lexical category covering prepositions, postpositions, and related relator items.} \glsmech{Maintained by stable distribution in phrase structure and recurrent relational semantics.} \glsdiag{Systematic behavior in complement selection, placement, and syntactic function.} \glsnote{A cross-linguistic comparative label requiring language-specific diagnostics.} \glsxsee{comparative concept, lexical category, category.}}
}

\newglossaryentry{grammaticalbasin}{
  name={grammatical basin},
  text={grammatical basin},
  description={\glsdef{The attractor region where form--value pairings are treated as grammatical by a speech community.} \glsmech{Regenerated by acquisition, entrenchment, alignment, and repair.} \glsdiag{Small perturbations are corrected back toward stable well-formed patterns.} \glsnote{A Chapter 14 extension of basin language to grammaticality itself.} \glsxsee{grammaticality, basin, homeostasis.}}
}

\newglossaryentry{maintenancemechanism}{
  name={maintenance mechanism},
  text={maintenance mechanism},
  description={\glsdef{Any causal process that actively regenerates a category's clustering after perturbation.} \glsmech{Includes feedback loops such as correction, alignment, gatekeeping, and transmission filtering.} \glsdiag{Disabling it weakens the cluster or changes its attractor structure.} \glsnote{Mechanism role is defined functionally: what keeps the profile stable over time.} \glsxsee{stabilizer, homeostasis, maintained kind.}}
}

\newglossaryentry{feelingofungrammaticality}{
  name={feeling of ungrammaticality},
  text={feeling of ungrammaticality},
  description={\glsdef{The metacognitive signal that a form--value coupling has failed or is at risk of failure.} \glsmech{Computed from structural expectations, probability, and processing strain.} \glsdiag{Drives hesitation, repair, and rejection judgments, but can misfire.} \glsnote{A detector output, not an infallible oracle of structure.} \glsxsee{detector, acceptability, grammaticality illusion.}}
}

\newglossaryentry{agreementattraction}{
  name={agreement attraction},
  text={agreement attraction},
  description={\glsdef{A processing illusion where a nearby feature competitor induces apparent agreement with the wrong controller.} \glsmech{Triggered by memory retrieval interference under real-time parsing.} \glsdiag{Ungrammatical strings can receive inflated acceptability when distractor features are salient.} \glsnote{Canonical evidence for selective fallibility in grammaticality judgments.} \glsxsee{selective fallibility, grammaticality illusion, detector.}}
}

\newglossaryentry{fluency}{
  name={fluency},
  text={fluency},
  description={\glsdef{Experienced ease of production or comprehension during linguistic processing.} \glsmech{Shaped by familiarity, frequency, and efficient retrieval pathways.} \glsdiag{Higher fluency tends to increase felt well-formedness even when structure is unchanged.} \glsnote{A major contributor to acceptability signals.} \glsxsee{acceptability, detector, processing economy.}}
}

\newglossaryentry{tipofthetongue}{
  name={tip-of-the-tongue},
  text={tip-of-the-tongue},
  description={\glsdef{A metacognitive state of partial access in which intended meaning is available but target form retrieval fails.} \glsmech{Arises from temporary retrieval bottlenecks in lexical-phonological access.} \glsdiag{Strong feeling of imminent recall with incomplete or fragmentary form output.} \glsnote{Shows dissociability of content access from form retrieval.} \glsxsee{fluency, detector, processing economy.}}
}

\newglossaryentry{passive}{
  name={passive},
  text={passive},
  description={\glsdef{A construction that backgrounds or demotes a canonical agent while promoting another participant to subject-like prominence.} \glsmech{Maintained by recurrent discourse-functional utility and constructional entrenchment.} \glsdiag{Systematic alternation with active clauses under predictable information-structural conditions.} \glsnote{A construction-level form--value pairing in the chapter's zipper framing.} \glsxsee{voice, construction, structural meaning.}}
}

\newglossaryentry{declarative}{
  name={declarative},
  text={declarative},
  description={\glsdef{A clause-type profile conventionally associated with assertion.} \glsmech{Stabilized by repeated use for information-giving and compatible prosodic/discourse routines.} \glsdiag{Predictable interaction with interrogative and imperative alternatives in clause-typing systems.} \glsnote{Clause type and speech act can diverge in marked contexts.} \glsxsee{interrogative, imperative, structural meaning.}}
}

\newglossaryentry{interrogative}{
  name={interrogative},
  text={interrogative},
  description={\glsdef{A clause-type profile conventionally associated with questioning and information-seeking force.} \glsmech{Maintained by recurring mappings among morphosyntax, prosody, and discourse practice.} \glsdiag{Stable formal contrasts with declarative and imperative clause types.} \glsnote{Includes but is not limited to interrogative phrases.} \glsxsee{interrogative phrase, declarative, imperative.}}
}

\newglossaryentry{imperative}{
  name={imperative},
  text={imperative},
  description={\glsdef{A clause-type profile conventionally associated with directive force.} \glsmech{Sustained by recurrent command/request uses and associated morphosyntactic marking.} \glsdiag{Regular contrasts with declarative and interrogative profiles in the same grammar.} \glsnote{Directive force can also be realized by non-imperative forms.} \glsxsee{declarative, interrogative, structural meaning.}}
}

\newglossaryentry{compositionality}{
  name={compositionality},
  text={compositionality},
  description={\glsdef{The principle that complex meanings are systematically built from component meanings plus mode of combination.} \glsmech{Maintained by stable constructional mappings and reusable combinatorial routines.} \glsdiag{Interpretive predictions extend productively to novel, well-formed combinations.} \glsnote{Distinct from grammaticality: a string may be grammatical yet semantically odd.} \glsxsee{structural meaning, grammaticality, inference.}}
}

\newglossaryentry{inference}{
  name={inference},
  text={inference},
  description={\glsdef{Interpretive reasoning that derives implications from linguistic signal plus contextual knowledge.} \glsmech{Combines grammatical cues with pragmatic and world-knowledge processes.} \glsdiag{Systematic enrichment beyond literal structural contribution.} \glsnote{Not all inference is grammatical; multiple channels contribute.} \glsxsee{structural meaning, compositionality, social meaning.}}
}

\newglossaryentry{appropriateness}{
  name={appropriateness},
  text={appropriateness},
  description={\glsdef{Fit between a linguistic form and the full bundle of values relevant in context, including structural, pragmatic, and social values.} \glsmech{Maintained by interactional feedback, norming, and register-sensitive expectations.} \glsdiag{Judgments can diverge from grammaticality when social or discourse values dominate.} \glsnote{Broader than grammatical well-formedness.} \glsxsee{grammaticality, social meaning, register.}}
}

\newglossaryentry{standingwave}{
  name={standing wave},
  text={standing wave},
  description={\glsdef{A Chapter 14 metaphor for dynamic equilibrium: a stable pattern sustained by ongoing underlying activity.} \glsmech{Persistence requires continuous maintenance work, not static storage.} \glsdiag{Pattern survives local fluctuation while preserving global profile.} \glsnote{Used to frame grammaticality as maintained equilibrium rather than fixed law.} \glsxsee{homeostasis, maintained kind, spinning top.}}
}

\newglossaryentry{spinningtop}{
  name={spinning top},
  text={spinning top},
  description={\glsdef{A Chapter 14 metaphor for a pattern that remains upright only while maintenance dynamics continue.} \glsmech{Ongoing corrective force prevents collapse into alternative states.} \glsdiag{Removing maintenance rapidly destabilizes the observed structure.} \glsnote{Highlights grammaticality's dual role as pattern and maintaining force.} \glsxsee{standing wave, homeostasis, stabilizer.}}
}

\newglossaryentry{syntacticsatiation}{
  name={syntactic satiation},
  text={syntactic satiation},
  description={\glsdef{Improved acceptability judgments after repeated exposure to some initially marginal sentence types.} \glsmech{Likely driven by expectation updating, task adaptation, and selective detector recalibration.} \glsdiag{Exposure effects are constraint-specific rather than uniform across structures.} \glsnote{Distinguishes flexible margins from robustly maintained categorical gaps.} \glsxsee{acceptability, selective fallibility, schema drift.}}
}

\newglossaryentry{schemadrift}{
  name={schema drift},
  text={schema drift},
  description={\glsdef{Shift in an abstract category schema under repeated exposure to atypical but recurrent exemplars.} \glsmech{Reweighting of feature priorities during learning and retrieval.} \glsdiag{The category profile itself changes, not just tolerance for noisy tokens.} \glsnote{Contrasts with one-off repair or mere response bias.} \glsxsee{rerepresentation, syntactic satiation, entrenchment.}}
}

\newglossaryentry{rerepresentation}{
  name={rerepresentation},
  text={rerepresentation},
  description={\glsdef{Re-encoding an input to fit an existing category without substantially altering the category schema.} \glsmech{Strategic reinterpretation and feature remapping at token level.} \glsdiag{Apparent accommodation with little long-run change in core category structure.} \glsnote{Useful contrast with schema drift.} \glsxsee{schema drift, detector, inference.}}
}

\newglossaryentry{conspicuousabsence}{
  name={conspicuous absence},
  text={conspicuous absence},
  description={\glsdef{A systematically missing pattern whose absence is informative because neighboring possibilities are robustly attested.} \glsmech{Maintained by entrenched alternatives and inhibitory feedback against the absent option.} \glsdiag{Persistent non-occurrence despite strong opportunities for production.} \glsnote{Used in Chapter 14 for unsatiatable structural gaps.} \glsxsee{workarounds, syntactic satiation, homeostasis.}}
}

\newglossaryentry{gardenpathsentence}{
  name={garden-path sentence},
  text={garden-path sentence},
  description={\glsdef{A sentence that induces an initially plausible parse that must be revised when later material arrives.} \glsmech{Sequential parsing heuristics favor local commitments that can later fail.} \glsdiag{Transient unacceptability or slowdown despite eventual grammatical interpretation.} \glsnote{A central source of detector/structure mismatch.} \glsxsee{grammaticality illusion, selective fallibility, detector.}}
}

\newglossaryentry{conditionedsystem}{
  name={conditioned system},
  text={conditioned system},
  description={\glsdef{A locally stabilized grammar-like regime defined relative to a source condition such as community, register, or interactional situation.} \glsmech{Maintained by source-specific learning, alignment, and evaluation loops.} \glsdiag{High internal projectibility within condition, weaker pooled projectibility across conditions.} \glsnote{Chapter 15 treats languages as mixtures of such regimes.} \glsxsee{conditioned probability, mixed bin, communicative situation.}}
}

\newglossaryentry{orderlyheterogeneity}{
  name={orderly heterogeneity},
  text={orderly heterogeneity},
  description={\glsdef{Structured sociolinguistic variation governed by stable constraints rather than random noise.} \glsmech{Regenerated by community transmission, social indexing, and interactional alignment.} \glsdiag{Variation patterns are predictable from social and contextual variables.} \glsnote{A key sociolinguistic observation integrated into the HPC maintenance view.} \glsxsee{dialect, register, conditioned probability.}}
}

\newglossaryentry{projectable}{
  name={projectable},
  text={projectable},
  description={\glsdef{Able to support reliable induction from observed to unobserved cases within a specified analytical purpose.} \glsmech{Depends on stable maintenance of relevant correlations.} \glsdiag{Predictions hold on held-out or novel cases at the chosen granularity.} \glsnote{Adjectival form of projectibility.} \glsxsee{projectibility, field-relative projectibility, category.}}
}

\newglossaryentry{academicregister}{
  name={academic register},
  text={Academic Register},
  description={\glsdef{A socially maintained register characterized by explicit norming around evidence display, stance marking, and discourse packaging in scholarly contexts.} \glsmech{Stabilized by institutional gatekeeping, training, review, and editorial correction.} \glsdiag{Systematic lexical-syntactic preferences distinct from casual conversation.} \glsnote{A paradigm case of register as maintenance regime.} \glsxsee{register, discourse community, conditioned system.}}
}

\newglossaryentry{basinrecognition}{
  name={basin recognition},
  text={basin recognition},
  description={\glsdef{The analytical task of identifying recurrent attractor structures across superficially different systems.} \glsmech{Works by detecting shared maintenance pressures that regenerate similar clusters.} \glsdiag{Comparative concepts track recurring solution-regions rather than fixed essences.} \glsnote{Chapter 15 reframes cross-linguistic comparison in these terms.} \glsxsee{comparative concept, basin, field-relative projectibility.}}
}

\newglossaryentry{mixture}{
  name={mixture},
  text={mixture},
  description={\glsdef{A population-level distribution formed by combining outputs from multiple latent conditioned systems.} \glsmech{Aggregation over distinct maintenance regimes produces blended marginals.} \glsdiag{Apparent looseness at aggregate level resolves into tighter clusters when conditioned.} \glsnote{Mixture structure is central to Chapter 15's social stabilization account.} \glsxsee{mixed bin, conditioned probability, conditioned system.}}
}

\newglossaryentry{realgeneral}{
  name={real general},
  text={real general},
  description={\glsdef{A law-like habit that is real by virtue of the regularity of its effects, not by being a concrete token.} \glsmech{Maintained through repeated causal reproduction of the same inferential pattern.} \glsdiag{Supports stable would-be inferences across new cases.} \glsnote{Peircean framing used in Chapter 15 for category realism.} \glsxsee{habit, would-be, projectibility.}}
}

\newglossaryentry{languageacquisitiondevice}{
  name={Language Acquisition Device},
  text={Language Acquisition Device},
  description={\glsdef{A historical label for a hypothesized innate language-learning module.} \glsmech{In Chapter 15 it functions as a foil to source-conditioned acquisition accounts.} \glsdiag{Contrasted with learning trajectories grounded in communicative situations and mixture structure.} \glsnote{Included for terminological orientation rather than endorsement.} \glsxsee{acquisition, communicative situation, conditioned system.}}
}

\newglossaryentry{construction}{
  name={construction},
  text={construction},
  description={\glsdef{A conventionalized form--value pairing at any level of schematicity.} \glsmech{Maintained by entrenchment, reuse, and community-level norming across repeated interactions.} \glsdiag{Supports productive extension to novel tokens while preserving structured constraints.} \glsnote{Core unit in usage-based and constructionist analyses.} \glsxsee{composite coupling, structural meaning, grammar.}}
}

\newglossaryentry{subcompositional}{
  name={subcompositional},
  text={subcompositional},
  description={\glsdef{Describing a level of interpretation below full propositional composition, where structural instructions are specified without complete lexical content.} \glsmech{Maintained by stable mappings from morphosyntactic form to abstract interpretive scaffolding.} \glsdiag{Captures clause-type and argument-structure contributions even when lexical semantics is odd.} \glsnote{Used in Chapter 14 for structural meaning.} \glsxsee{structural meaning, compositionality, grammaticality.}}
}

\newglossaryentry{ungrammatical}{
  name={ungrammatical},
  text={ungrammatical},
  description={\glsdef{Outside the maintained basin for a grammar's form--value coupling.} \glsmech{Detected through mismatch signals, though felt judgments can be noisy.} \glsdiag{Persistent failure under competent speaker judgments beyond transient processing confounds.} \glsnote{Distinguished from mere unacceptability caused by processing or contextual factors.} \glsxsee{grammaticality, acceptability, grammatical basin.}}
}

\newglossaryentry{wiringdiagram}{
  name={wiring diagram},
  text={wiring diagram},
  description={\glsdef{A Chapter 14 metaphor for structural meaning as an instruction-level map of how interpretation is assembled.} \glsmech{Stability comes from recurrent form-to-structure mappings independent of lexical plausibility.} \glsdiag{Well-formed structure can remain intact even when lexical combinations are semantically bizarre.} \glsnote{Contrasts with content-level inference or world-knowledge coherence.} \glsxsee{structural meaning, subcompositional, zipper.}}
}

\newglossaryentry{immunesystem}{
  name={immune system},
  text={immune system},
  description={\glsdef{A Chapter 14 metaphor for the corrective maintenance layer that restores grammatical couplings after local deviations.} \glsmech{Realized in repair, self-monitoring, correction, and alignment feedback loops.} \glsdiag{Small violations trigger stabilization responses rather than unrestricted drift.} \glsnote{Metaphor for maintenance dynamics, not a literal biological mechanism in grammar.} \glsxsee{maintenance mechanism, stabilizer, feeling of ungrammaticality.}}
}

\newglossaryentry{phenomenology}{
  name={phenomenology},
  text={phenomenology},
  description={\glsdef{The experienced feel of a cognitive state as it presents to awareness.} \glsmech{In Chapter 14, shaped by metacognitive detector outputs rather than direct access to grammatical structure.} \glsdiag{Can track grammatical coupling imperfectly, enabling stable illusion patterns.} \glsnote{Used to separate lived judgment experience from underlying mechanism.} \glsxsee{detector, acceptability, grammaticality illusion.}}
}

\newglossaryentry{contrast}{
  name={contrast},
  text={contrast},
  description={\glsdef{A functional opposition where differences in form support differences in linguistic value.} \glsmech{Maintained by communicative pressure to preserve discriminable alternatives.} \glsdiag{Merger or neutralization changes predictability and can trigger repair or reorganization.} \glsnote{Central in phonology, but relevant to other form--value systems.} \glsxsee{value, hard coupling, compositionality.}}
}

\newglossaryentry{workarounds}{
  name={workarounds},
  text={workarounds},
  description={\glsdef{Alternative constructions that speakers preferentially use when a target configuration is disfavored or absent.} \glsmech{Stabilized by competition, retrieval strength, and communicative success under existing norms.} \glsdiag{Persistent substitution patterns in corpora and judgment tasks around constrained zones.} \glsnote{In Chapter 14, dense workaround ecology helps explain conspicuous absences.} \glsxsee{conspicuous absence, syntactic satiation, construction.}}
}

\newglossaryentry{contingent}{
  name={contingent},
  text={contingent},
  description={\glsdef{Dependent on historically evolved maintenance conditions rather than fixed by necessity.} \glsmech{Current equilibria persist because stabilizers currently support them, not because alternatives are impossible.} \glsdiag{Shift in social or learning conditions can move the system to a different stable basin.} \glsnote{Key claim in Chapter 14's account of grammatical change potential.} \glsxsee{homeostasis, basin, maintenance mechanism.}}
}

\newglossaryentry{grammar}{
  name={grammar},
  text={grammar},
  description={\glsdef{A maintained system of conventionalized form--value couplings used for production and interpretation.} \glsmech{Regenerated by acquisition, interactional alignment, transmission, and social norming.} \glsdiag{Supports projectible inferences within a conditioned system while allowing mixture at population scale.} \glsnote{In Chapter 15, often modeled as a family of conditioned systems rather than a single homogeneous object.} \glsxsee{construction, conditioned system, mixture.}}
}

\newglossaryentry{conditioned}{
  name={conditioned},
  text={conditioned},
  description={\glsdef{Indexed to specific source variables (speaker, community, register, setting) rather than pooled across them.} \glsmech{Captures the level where maintenance processes operate and regenerate structure.} \glsdiag{Conditioned analyses typically reveal tighter and more projectible distributions than unconditioned aggregates.} \glsnote{Chapter 15 treats this as the correct scale for mechanism diagnosis.} \glsxsee{conditioned probability, conditioned system, mixed bin.}}
}

\newglossaryentry{communicativesituations}{
  name={Communicative Situations},
  text={Communicative Situations},
  description={\glsdef{Plural label for recurrent com-sits: interactional settings that bundle interlocutors, contexts, and expectations.} \glsmech{Learning proceeds by tracking these source-partitioned routines before broad abstractions emerge.} \glsdiag{Early acquisition patterns cluster by situation before language-level labels stabilize.} \glsnote{Alias entry to match chapter terminology.} \glsxsee{communicative situation, conditioned system, acquisition.}}
}

\newglossaryentry{slur}{
  name={slur},
  text={slur},
  description={\glsdef{A socially policed lexical label combining denotation with derogatory stance.} \glsmech{Maintained and transformed through public norming, sanction, and shifting community boundaries.} \glsdiag{Usage judgments and meaning profiles are tightly indexical to discourse community norms.} \glsnote{Chapter 15 uses slur history to illustrate looping social maintenance.} \glsxsee{thick term, looping kind, social meaning.}}
}
