\chapter{Diagnostic methodology}
\label{app:diagnostics}

% This appendix contains methodological details for the projectibility and homeostasis diagnostics
% discussed in Chapter~\ref{ch:the-category-zipper}.

\section{A compact formalization}
\label{sec:app:compact-formalization}

The main argument treats an HPC category as a three-part object: a property profile, a maintenance story, and a projectibility payoff \parencite{boyd1991}. The formalization below makes those parts explicit and keeps them distinct. This is regimentation, not an essence-style definition: the goal is to state success conditions transparently, not to legislate category membership by notation.

Why put this here at all? Because it makes the success conditions checkable: you can see exactly what counts as profile evidence, maintenance evidence, and predictive lift, and what would count against each. The prose argument in the main chapters is primary; the notation is an audit trail.
Read the clauses as design constraints, not turn-key estimators: the hard work remains operationalization, sampling, and perturbation design.

For a candidate category $C$ and analytical field
\[
F=(Q_F,I_F,L_F,D_F,\Sigma_F),
\]
define
\[
\mathrm{HPC}_F(C)=\left\langle \mathcal{P}_F(C),\mathcal{M}(C),\Pi_F(C)\right\rangle .
\]
Here $Q_F$ is a query family, $I_F$ an intervention family, $L_F$ a loss function, $D_F$ a declared diagnostic/measurement regime, and $\Sigma_F$ a scope declaration (population, register, time slice, grain). $\mathcal{P}_F(C)$ is the profile layer, $\mathcal{M}(C)$ is the maintenance layer, and $\Pi_F(C)$ is the field-indexed projectibility layer.

Toy reading with concrete diagnostics: take $C$ as English count nouns in edited prose, and let diagnostics $A/B/C$ in $D_F$ be numeral compatibility, plural morphology, and agreement behaviour. Let $Q_F$ ask whether unseen tokens pattern as count-like, let $\Sigma_F$ fix scope to a declared register/time slice, and let $L_F$ be held-out prediction loss. Then $S_F(C)$ asks whether $A/B/C$ co-vary above baseline, $\tau_{\delta}(a)$ asks whether that profile returns after perturbation, and $\Pi_F(C)$ asks whether modelling $C$ beats a baseline on new data.

\subsection*{Profile layer: dependence structure}

Let $X^F=(X_1,\dots,X_p)$ be the pre-registered diagnostic family from $D_F$ for tokens assigned to $C$. A compact dependence score is
\[
S_F(C)=\frac{1}{|E_C|}\sum_{(i,j)\in E_C} I(X_i;X_j)
      -\frac{1}{|E_0|}\sum_{(i,j)\in E_0} I(X_i;X_j),
\]
where $I(\cdot;\cdot)$ is mutual information, $E_C$ is a pre-declared set of diagnostic-variable index pairs $(i,j)$ (all pairs or graph edges), and $E_0$ is a matched baseline pair set. $S_F(C)$ denotes a bias-controlled dependence summary (mutual information is the default implementation, but other dependence functionals are admissible). This restates the co-occurrence intuition in information-theoretic terms: a profile exists when $S_F(C)>0$ robustly across resamples. The statistic is a summary, not a claim that pairwise dependence is sufficient in principle; higher-order dependence can be substituted where theory or data demands it.

\subsection*{Maintenance layer: dynamical mechanism}

Let $Z_t$ be a latent maintenance state encoding stabilizers (acquisition, entrenchment, alignment, transmission, functional pressure). Write
\[
Z_{t+1}=f_{\theta}(Z_t,U_t)+\eta_t,\qquad X_t=g_{\phi}(Z_t)+\epsilon_t .
\]
Interventions are represented in Pearl-style notation \parencite{pearl2009}:
\[
Z_{t+1}^{(a)}=f_{\theta}\!\left(Z_t^{(a)},U_t,\mathrm{do}(A_t=a)\right)+\eta_t .
\]
Graph and network implementations are special cases of this layer \parencite{khalidi2015three,borsboom2017}.

Homeostasis\ixsq{homeostasis!defined} is one especially strong maintenance pattern: a return-to-basin (of attraction) claim. Let $\mathcal{B}_C^{(a)}$ be the post-perturbation operational proxy for $C$'s basin in diagnostic space and let distance-to-set be defined as $d(x,\mathcal{B})=\inf_{b\in\mathcal{B}} d(x,b)$. Define
\[
\tau_{\delta}(a)=\inf\left\{h\geq 0 : \mathbb{E}\!\left[d\!\left(X_{t+h}^{(a)},\mathcal{B}_C^{(a)}\right)\right]\leq \delta\right\}.
\]
For local perturbations, $\mathcal{B}_C^{(a)}=\mathcal{B}_C$; for structural perturbations, basin geometry may shift. For admissible perturbations $a$, homeostasis holds when $\tau_{\delta}(a)$ is finite for a pre-registered tolerance $\delta$. This operationalizes homeostasis proper (active corrective return toward the maintained basin, with exact-state return as a limiting case). But the book's maintenance picture is broader than homeostasis proper. Metastability, causal isolation, frozen accidents, and related stability sources remain part of $\mathcal{M}(C)$ and can satisfy the maintenance layer when independently evidenced, even if they do not yield finite return times under the relevant perturbations.

\subsection*{Projectibility layer: field-relative lift}

Given $F=(Q_F,I_F,L_F,D_F,\Sigma_F)$, projectibility is expected lift over a baseline:
\[
\Pi_F(C)=\mathbb{E}_{(x,y)\sim D_F}\!\left[L_F\!\left(\hat{y}_{\mathrm{base}}(x),y\right)-L_F\!\left(\hat{y}_{C}(x),y\right)\right].
\]
Projectibility\ixsq{projectibility!defined} for field $F$ holds when $\Pi_F(C)>0$ robustly out of sample.

This is where interest-relativity belongs. Craver's and Onishi--Serpico's challenge is real: explanatory interests change variable sets, scope choices, and evaluation targets \parencite{craver2009,onishi2022}. In this formalization, that dependence is explicit in $D_F$ and $\Sigma_F$ as well as $Q_F$ and $L_F$, while maintenance claims remain empirical via perturbation and recovery. This avoids a mechanism-free account that is too thin and avoids treating any one graph as uniquely privileged \parencite{lemeire2018}.

This structure also links directly to existing formal threads. Martínez's informationally connected property-cluster proposal maps to the profile score $S_F(C)$ via mutual information \parencite{martinez2015informationally}. Rehder's causal-model theory instantiates the maintenance--projectibility bridge with a generative Bayesian network over features \parencite{rehder2003causal}. Van Rooij and Schulz's natural-kind analysis in Pearl-style causal models maps to the intervention and ablation clauses \parencite{vanrooijschulz2019}.

\subsection*{Operational decision rule}

For a declared field $F$, treat $C$ as an HPC category only if all four checks pass:
\begin{enumerate}
    \item Profile: $S_F(C)$ exceeds a pre-registered null margin under the declared $D_F,\Sigma_F$ regime.
    \item Maintenance: either homeostasis proper is supported by $\tau_{\delta}(a)<\infty$ across the admissible perturbation set, or broader maintenance evidence identifies a stabilizing route robust enough to explain persistence under the declared $D_F,\Sigma_F$ regime.
    \item Projectibility: $\Pi_F(C)>0$ on held-out data.
    \item Ablation/perturbation: removing or weakening a candidate stabilizer worsens at least one target by a pre-registered margin: $S_F(C)$ decreases, $\Pi_F(C)$ decreases, or, where homeostasis proper is at issue, $\tau_{\delta}(a)$ increases (possibly to $\infty$).
\end{enumerate}

\subsection*{Failure-mode mapping}

The Part~II and Part~III failure modes map directly onto these quantities:
\begin{description}
    \item[Too thin.] Low $S_F(C)$, weak or absent recovery, and near-zero $\Pi_F(C)$.
    \item[Too fat.] Non-trivial $S_F(C)$ but no stable ablation signature and poor transfer across fields $F$.
    \item[Merely negative.] Complement-defined grouping with no coherent maintenance dynamics and no predictive lift.
\end{description}

\section{Operationalizing the diagnostics}

The two criteria aren't operationalized symmetrically.

This section compresses the audit-template workflow from Chapter~\ref{ch:failure-modes}: declare scope and diagnostics, test profile and projectibility, probe maintenance by perturbation and ablation, and make falsifiers explicit.

Projectibility\ixsq{projectibility!defined} is tested directly: the question is whether patterns learned from one sample generalize to another~-- can we infer properties of new instances from old ones? This is tested via out-of-sample prediction: cross-validated classification accuracy, held-out F1 scores, cross-corpus transfer. Concrete thresholds can be declared in advance: ROC-AUC $\geq 0.70$, PR-AUC $\geq 0.70$, macro-F1 $\geq 0.35$, held-out cohesion $\geq 0.30$. These thresholds are somewhat arbitrary conventions, but that's the point: declaring them in advance disciplines the inference.

Homeostasis\ixsq{homeostasis!defined} is inferred: the question is whether we can identify stabilizers that maintain the property cluster and find their predicted signatures in the data. We name candidate mechanisms (articulatory constraints, frequency entrenchment, normative enforcement), derive their predicted signatures (scaling curves, perturbation sensitivity, cue covariance), and check whether the data is consistent with those predictions. The inference is defeasible~-- the signatures could arise from other sources~-- but it's grounded in mechanism\ixsq{mechanism!defined}--signature pairings rather than asserted by narrative alone.

The relationship between mechanism\ixsq{mechanism!defined} and projection is asymmetric. Figure~\ref{fig:app:asymmetry} shows the structure: mechanisms causally support covariance, which enables projection; projection provides evidence for kindhood only when accompanied by a credible mechanism story.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=2cm,
    box/.style={rectangle, draw, minimum width=2.5cm, minimum height=1cm, align=center},
    arrow/.style={->, >=stealth, thick}
]
\node[box] (mech) {Mechanisms};
\node[box, right=of mech] (cov) {Covariance};
\node[box, right=of cov] (proj) {Projection};

\draw[arrow] (mech) -- node[above] {maintain} (cov);
\draw[arrow] (cov) -- node[above] {supports} (proj);

\draw[arrow, dashed] (proj.south) -- ++(0,-0.8) -| node[below, pos=0.25] {Test projection} (cov.south);
\draw[arrow, dashed] (cov.south) -- ++(0,-0.5) -| node[below, pos=0.25] {Check mechanisms} (mech.south);
\end{tikzpicture}
\caption{The causal arrow runs from mechanisms to projection; the evidential arrows run in reverse. Both diagnostics must succeed independently for kindhood to be warranted.}
\label{fig:app:asymmetry}
\end{figure}

\section{Failure modes}

A category might project without identifiable stabilizers (possibly spurious overfitting). A category might have plausible stabilizers that don't produce stable clustering (broken homeostasis\ixsq{homeostasis!perturbation test}). Both diagnostics have to pass for kindhood to be warranted. The independence matters: it means the framework can fail in two directions.

Three failure types recur across linguistic levels:

\begin{description}
    \item[Too thin.] Nonce coinages, campaign-season blends, idiolectal \enquote{style phonemes}, and child-only overregularizations within adult Standard English\ixgq{English!Standard} don't pass the projectibility\ixsq{projectibility!induction from} test in the relevant population--time slice. Held-out prediction collapses; the stabilizers one would expect to bind properties (frequency, entrenchment, community norms) are absent or act in the opposite direction. These cases are explananda for learning or diffusion, but not kinds.
    
    \item[Too fat.] Cross-linguistic umbrellas such as \enquote{resultative} or \enquote{ditransitive} pool patterns maintained by different morphosyntactic resources, cue reliabilities, and norming regimes. The pooled set can look impressive descriptively, but the mechanism story is disjunctive: dispersion in one language, selectional licensing in another, constructional analogy in a third \ixnq{Croft, William}\ixnq{Haspelmath, Martin}\parencite{croft2001,haspelmath2010}. Cross-corpus prediction drifts toward family- or area-specific quirks, ablations fail to show stable redundancy, and effects wash out under lineage-pruning. The right move is to localize the ontology: retain language-internal equilibria as kinds and treat the global umbrella as an interest-relative taxonomy.
    
    \item[Merely negative.] Complement classes~-- \enquote{ungrammatical strings}, \enquote{all exceptions to rule R}~-- are defined by what they're not rather than what they are. They don't project (there's no stable covariance to learn) and they don't admit a non-accidental mechanism story. A caveat: structured subsets within a complement class can be HPC\ixsq{homeostatic property cluster!membership} candidates in their own right. L2 transfer clusters (e.g., persistent article-drop errors among speakers of articleless L1s) may show stable covariance, predictable error profiles, and mechanism--signature pairings that the grab-bag \enquote{all L2 errors} lacks.
\end{description}

\section{Falsification conditions}

The diagnostics generate clear disconfirmers at each level:

\begin{enumerate}
    \item Phonemes: If the /y/ scaling effect collapses under lineage-pruning (one language per subfamily), the claim that marked segments scale with inventory size fails. If mergers showed no relationship to cue reliability~-- occurring where acoustic cues are robust rather than degraded~-- the homeostasis\ixsq{homeostasis!perturbation test} story would be wrong.

    \item Words: If semantic drift produced no cohesion loss~-- if high-drift adjectives retained the same distributional neighbourhoods as low-drift controls~-- the framework would have nothing to explain. If regularization probability showed no relationship to frequency~-- if high-frequency irregulars like \mention{went}\ixlq{went} were just as vulnerable as low-frequency ones like \mention{clove}\ixlq{clove}~-- the entrenchment stabilizer would be decorative.

    \item Constructions: If cross-corpus transfer performed at chance, or if ablating formal cues (parallelism, licensing context) produced no interpretable degradation pattern, the claim that constructions are HPCs would lack evidential support.
\end{enumerate}
