\chapter{Grammaticality itself}
\label{ch:grammaticality-itself}
\epigraph{it's impossible to get rid of the illusion in spite of our better knowledge.}{~-- Hermann von Helmholtz, \textit{Treatise on Physiological Optics} (trans. J. P. C. Southall).}

The asterisk is the most common symbol in syntax, and one of the least likely to come with operating instructions. But what does the asterisk diagnose?

Chapter~\ref{ch:what-we-havent-been-asking} asked a question almost no one asks: is grammaticality\ixsq{grammaticality} itself a category\ixsq{kind!linguistic kinds as}? Not whether this sentence is grammatical, but what kind of thing we're probing when we ask. If linguistic categories are homeostatic property clusters, then grammaticality itself is an HPC\ixsq{homeostatic property cluster!grammaticality}~-- a projectible, maintained coupling between form and value\ixsq{value}. And the feeling of ungrammaticality\ixsq{feeling of ungrammaticality} that guides our judgments is not an oracle; it's a noisy detector\ixsq{detector}, shaped by entrenchment and processing ease, capable of illusions.

Chapter~\ref{ch:dynamic-discreteness} introduced a two-layer picture: grammaticality\ixsq{grammaticality} that is dynamically discrete~-- maintained sharp by the same mechanisms that could let it soften~-- filtered through processing noise to yield gradient acceptability\ixsq{acceptability}. What this chapter adds is the claim that grammaticality\ixsq{grammaticality} is a full HPC in the book's sense, and that the feeling of ungrammaticality is the stabilizer that maintains it. The sections that follow develop the coupling, identify the detector's characteristic failure modes, and examine the account's empirical commitments.


\section{The HPC claim}
\label{sec:15:hpc-claim}

Grammaticality\ixsq{grammaticality} is one form--value\ixsq{value} pairing among many. The \enquote{value\ixsq{value}} side is most often a kind of meaning~-- specifically, \term{structural meaning\ixsq{structural meaning}}, though later sections will note cases where the value side is not meaning at all.

\term{Structural meaning\ixsq{structural meaning}} is what grammar\ixsq{grammar} contributes before lexical content fills the slots. It includes dependency relations, argument structure, tense--aspect anchoring, clause type, and information structure\ixsq{information structure}. It is \term{subcompositional\ixsq{subcompositional meaning}}: \mention{Colorless green ideas sleep furiously} has impeccable structural meaning\ixsq{structural meaning} and absurd lexical content, which is why it feels strange but not \term{ungrammatical\ixsq{ungrammatical}}. Structural meaning\ixsq{structural meaning} is not semantic coherence, contextual felicity, or whatever the hearer infers overall. (A full operational definition~-- with annotation criteria for borderline cases~-- is a task for the research programme this chapter inaugurates, not a prerequisite. The worked examples below are the chapter's answer to where the boundary falls.)

The HPC\ixsq{homeostatic property cluster!grammaticality} claim, then, is that grammaticality\ixsq{grammaticality} is, prototypically, the maintained, projectible coupling between morphosyntactic form and structural meaning\ixsq{value} at the relevant grain. One such pairing, once mastered, extends to close analogues and novel instances. This is the \term{zipper\ixsq{zipper}}: form, value\ixsq{value}, and the interpretant\ixsq{interpretant} that links them~-- the projective habit\ixsq{habit} that makes the pairing count as a pairing (Chapter~\ref{ch:the-category-zipper}). Ungrammaticality\ixsq{ungrammatical} is what it feels like when the zipper\ixsq{zipper} won't close~-- when form and value fail to mesh under the interpretant that should bind them.

A distinction matters here. \term{Licensing} is what the grammar permits~-- where \enquote{the grammar} is the network of entrenched constructions and their inheritance links, not an autonomous rule system. Licensing covers the structural slots, the selectional frames, the combinatoric possibilities. \term{Coupling} is what a community has entrenched in those slots. Semantic compatibility gates what \emph{can} enter a coupling; entrenchment determines what \emph{stays}. The zipper's teeth are cut by the grammar~-- form on one side, value on the other~-- but the slider that draws them together is the interpretant: the community's entrenched habit of treating this form as delivering this value. Age expressions will make this concrete: the structural slot for \mention{have} + NP exists in English, but the community hasn't filled it with age (§\ref{sec:15:detector}).

The rest of this chapter asks what the zipper metaphor actually commits us to~-- what it predicts, how the detector works, and where it breaks down.


\subsection{Projectibility}
\label{sec:15:projectibility}
None of what follows will surprise a working linguist. But spelling it out matters, because the obvious cases are exactly the evidence that grammaticality projects. If you know a form--value\ixsq{value} pairing is grammatical\ixsq{grammaticality}, you can do something with it. This is the inductive licence that makes grammaticality\ixsq{grammaticality} a category in the book's sense, not just a property of individual sentences. To use the vocabulary of Chapter~\ref{ch:stabilizers}, grammaticality\ixsq{grammaticality} sustains a good bet\ixsq{good bet}: encounter this pattern and you can trust it in new territory. The bet is \term{defeasible\ixsq{defeasibility}}~-- it can be defeated by competing entrenchment, processing overload, or community divergence~-- and that defeasibility is a feature of the framework, not a failure of the data. Essentialist grammaticality predicts exceptionless judgments; HPC grammaticality predicts defeasible ones. The illusions, satiation effects, and individual differences that follow are all instances of defeasibility in action.

The bets are concrete. A speaker who has mastered the double-object construction\ixsq{construction} (\mention{gave her the book}\ixlq{give}) deploys it with verbs never heard in that frame: \mention{texted her the address}, \mention{Slacked him the link}. The would-be\ixsq{would-be} is: encounter this form and you \emph{would} get transfer plus affected recipient. The same projection drives analogical extension\ixsq{analogy}~-- the progressive-aspect coupling (\mention{I'm writing}) extends to \mention{I'm Googling}, \mention{I'm Zooming} as new verbs enter the language. In both cases, new lexical content slots into an existing form--value\ixsq{value} pairing. The speaker hasn't memorized these sentences; she's projected from the coupling.

Repair provides a third line of evidence. You can't fix the wiring unless you know what the wiring should deliver~-- and every correction presupposes projection\ixsq{projectibility!induction from} from the known coupling. A speaker reaching for \mention{adaptiveability} catches the error mid-word and self-corrects to \mention{adaptability}\ixlq{adaptability}, because the would-be\ixsq{would-be} fires before production completes. Post-hoc correction works the same way, since both parties share the would-be\ixsq{would-be}. But these projections aren't all equally strong. Tight couplings project further: subject--verb agreement extends across novel combinations. Loose couplings project less: island constraints\ixsq{island constraint} show construction\ixsq{construction}-specific limits, and speakers disagree about the edges.

Compare this to mere acceptability\ixsq{acceptability}. Acceptability tells you how the detector\ixsq{detector} responded to one token on one occasion. Grammaticality\ixsq{grammaticality} tells you what to expect across tokens~-- across speakers, across novel instances, across contexts~-- because the category projects where any single detector\ixsq{detector} reading doesn't. §\ref{sec:15:value-every-grain} will show this coupling operating at every level of the grammar\ixsq{grammar}, from phonemic contrast to clause type.

Table~\ref{tab:15:architecture} maps the analytical structure that the rest of the chapter develops. Each row isolates a different component of the account; the columns locate its timescale, its key distinction from what it's most easily confused with, and the section where it's tested.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.16\textwidth}>{\raggedright\arraybackslash}p{0.20\textwidth}>{\raggedright\arraybackslash}p{0.28\textwidth}>{\raggedright\arraybackslash}p{0.20\textwidth}@{}}
\toprule
\textbf{Component} & \textbf{Timescale} & \textbf{Key distinction} & \textbf{Developed in} \\
\midrule
Coupling (form--value--interpretant) & Years--generations & Coupling $\neq$ licensing: licensing is what the grammar permits; coupling is what a community has entrenched & §\ref{sec:15:hpc-claim} \\
Detector (feeling of ungrammaticality) & 100 ms--seconds & Grammaticality $\neq$ acceptability: the category $\neq$ the readout & §\ref{sec:15:detector} \\
Negative space & -- & Grammaticality $\neq$ other channels: morphosyntactic coupling $\neq$ register, accent, pragmatic fit & §\ref{sec:15:negative-space} \\
Breakdown & Months--generations & Detector misfire $\neq$ coupling breakdown: illusions are readout errors; language change is maintenance loss & §\ref{sec:15:breakdown} \\
Dual role & System-level & Category $\neq$ mechanism: grammaticality is both a projectable kind and a stabiliser for other kinds & §\ref{sec:15:dual-role} \\
\bottomrule
\end{tabular}
\caption{Analytical architecture of the grammaticality account. Each row is a component of the HPC claim; the key-distinction column marks what it is most easily confused with.}
\label{tab:15:architecture}
\end{table}


\section{The detector}
\ixsq{stabilizer!defined}
\label{sec:15:detector}

Perhaps the most important proximate \term{stabilizer\ixsq{stabilizer!as mechanism role}}~-- the thing that corrects deviations in real time~-- is a \term{feeling of ungrammaticality\ixsq{feeling of ungrammaticality}}. When something's wrong with the form--value\ixsq{value} coupling, speakers notice. They hesitate, repair, rephrase. But the feeling is a \term{detector\ixsq{detector}}, not an oracle~-- and like all detectors, it's noisy. It's informed by probability: a novel sentence is weighted by its likelihood under the distribution of attested forms~-- not just distance from the nearest exemplar, but probability given the entire pattern. Low-probability forms feel suspect even when grammatically licensed.

But probability relative to what? Raw frequency is the wrong baseline. What matters is the difference between a \emph{thin} coupling and an \emph{absent} one~-- between a form that's licensed but rarely deployed and a form that's never licensed at all.

Independent relative \mention{whose}\ixlq{whose} (§\ref{sec:6:independent-whose}) is vanishingly rare in corpora. But the contexts that license it~-- contrastive parallelism, deictic anchoring~-- are themselves rare. Conditioned on opportunity, the construction may not be infrequent at all. What matters for the detector\ixsq{detector} is not how often a form appears in the corpus but how often it appears \emph{given a context where it could}. A thinly coupled form feels marginal because the prior is thin, even though the coupling is intact. An unlicensed form feels ungrammatical because no conditioning rescues it. In Bayesian terms: low posterior versus zero likelihood.

The distribution the detector\ixsq{detector} tracks is built by entrenchment\ixsq{entrenchment!frequency effects}: exposure tunes the prior. This is why \term{acceptability\ixsq{acceptability}} and \term{grammaticality\ixsq{grammaticality}} come apart. Acceptability\ixsq{acceptability} is what the detector\ixsq{detector} reports; grammaticality\ixsq{grammaticality} is the category\ixsq{category!as HPC kind}~-- the HPC\ixsq{homeostatic property cluster!grammaticality} whose core is the form--value\ixsq{value} coupling. The two usually align, but they don't always.

Coupling shapes processing; processing shapes detector\ixsq{detector} output; detector\ixsq{detector} output crosses threshold; threshold triggers repair; repair reshapes coupling. To keep this concrete, we can write the loop with five moving parts:
\begin{quote}
\textbf{Mechanism\ixsq{mechanism!defined} sketch (detector\ixsq{detector} and maintenance loop).}
\begin{enumerate}
    \item \textbf{Coupling state}: how strongly a construction\ixsq{construction} currently links morphosyntactic form to structural meaning\ixsq{structural meaning} for a speaker in a community.
    \item \textbf{Processing channel}: token-time load~-- surprisal, memory pressure, reanalysis cost.
    \item \textbf{Detector\ixsq{detector} output}: a graded mismatch signal combining coupling state and processing load, plus noise.
    \item \textbf{Decision threshold}: above some level, speakers report \enquote{ungrammatical\ixsq{ungrammatical}} and/or initiate repair.
    \item \textbf{Entrenchment\ixsq{entrenchment!defined} update}: exposure and correction histories move both coupling strength and threshold over time.
\end{enumerate}
This keeps the levels separate: token-time parsing (\textasciitilde100 ms--seconds), individual entrenchment (days--years), and community norm dynamics (years--generations). The loop's consequence is to keep the coupling in a state that supports projection: perturb it, and the system restores not just the co-occurrence but the inferential licence it underwrites.
\end{quote}

The sketch has a diagnostic consequence: if we can prise apart the three timescales, we can tell which component has moved. That requires separating processing's two roles in the sketch. As a \emph{filter}, it constrains what patterns the mechanisms can maintain: acquisition bottlenecks and transmission fidelity limits screen out patterns that exceed processing or memory capacity, shaping the space of possible grammars. As a \emph{modulator}, it affects how licensed patterns feel during comprehension: a grammatical structure can feel unacceptable when processing load is high, without its licensing status changing. The filter role is diachronic, while the modulator role is token-time.

This sketch commits us to an identification strategy. If processing load rises while coupling is stable, we should see transient detector\ixsq{detector} spikes without durable shifts in repair norms~-- a centre-embedded sentence feels worse when you're tired, but no one starts repairing it differently the next morning. If coupling weakens at matched load, we should see durable shifts in repair preferences and cohort trajectories~-- the retreat of \mention{whom}\ixlq{whom} tracks cohort, not fatigue; younger speakers don't just tolerate \mention{who}\ixlq{who} in object position, they stop correcting to \mention{whom} altogether. If thresholds move while coupling remains stable, ratings should move before production and repair distributions do~-- a speaker told that split infinitives are wrong may rate them lower without changing how she talks. So ratings alone can't settle the ontology. A sentence that sounds worse after a long day isn't evidence of grammatical change; a sentence that an entire cohort starts repairing differently is. The evidential standard is convergence across online processing, repair behavior, and community-distributional data (Chapters~\ref{ch:failure-modes} and~\ref{ch:what-changes}).

But here's the puzzle: if grammaticality\ixsq{grammaticality} \emph{is} the maintained coupling, and entrenchment\ixsq{entrenchment!resistance to change} is what maintains it, then \enquote{tuning the detector\ixsq{detector}} and \enquote{shaping what counts as grammatical} can't be fully separated over historical time. At a single time-slice, they \emph{are} separable~-- detector\ixsq{detector} output can misfire against a currently entrenched state, as illusions show. Across time, though, detector\ixsq{detector}-guided repair and uptake reshape the coupling itself. The coupling tunes the detector\ixsq{detector}; the detector\ixsq{detector}, over time, tunes the coupling. That's not a vicious circle; it's a reflexive equilibrium, and the age-expression data show what it looks like in practice. The explanatory arrow goes through history: today's coupling exists because past speakers' corrections maintained it~-- past detectors fired, past repairs succeeded, and the resulting entrenchment was transmitted to new speakers who inherited both the coupling and the disposition to correct deviations from it. Remove that history and the present state is unexplained; invoke it and the apparent circularity dissolves into a diachronic spiral. The coupling at any moment is sustained by a community whose members acquired it from predecessors whose own detectors kept it stable.

This reflexive dynamic is characteristic of social HPCs generally. Prescriptive enforcement, metalinguistic awareness, and explicit teaching are maintenance pathways that operate \emph{through} classification~-- what \textcite{hacking1999} calls a \term{looping effect}: the category shapes the behaviour that sustains it. Biological HPCs can loop through other channels~-- niche construction is a loop~-- but not through classification, because a beaver doesn't change because you call it a beaver. Grammaticality loops through classification, which is what makes it a social HPC.

Consider age expressions: in French\ixgq{French} and Spanish\ixgq{Spanish}, age is expressed with \mention{avoir}\ixlgqs{avoir}{avoir}{French}{have}/\mention{tener}\ixlgqs{tener}{tener}{Spanish}{have}: \mention{J'ai vingt ans} (`I have twenty years'). In English\ixgq{English}, \ungram\mention{I have twenty years} is ungrammatical\ixsq{ungrammatical} as a statement of age~-- though perfectly fine as a statement of time (until retirement, on the job, etc.). A superficially similar \mention{have}~+ NP frame is available, but the entrenched constructional mapping differs: the English \mention{have}~+ NP construction couples to possession or experience, not age-predication. The coupling is real but its contents are conventional.

We couldn't correct grammatical errors if we couldn't derive the intended meaning without the grammar. When someone says \ungram\mention{I have twenty years} for their age, we understand what they mean~-- and then we notice the mismatch. The feeling of ungrammaticality\ixsq{feeling of ungrammaticality} arises from comparison: the structural meaning\ixsq{structural meaning} the coupling delivers~-- possession or experience of a duration~-- versus the meaning the speaker intends~-- age-predication. We need both to detect the error. Comprehension is partly independent of grammaticality\ixsq{grammaticality}~-- and grammaticality\ixsq{grammaticality} is partly a matter of checking the grammar's signal against other inference channels.

\subsection{Illusions}
\ixsq{grammaticality illusion}
\label{sec:15:illusions}

The strongest evidence that the feeling is dissociable from the structure comes from \term{grammaticality\ixsq{grammaticality} illusions}~-- cases where the detector\ixsq{detector} misfires. The classic case is \textcite{Bever1970}'s garden-path sentence\ixsq{garden-path sentence}:

\ea[]{\label{ex:15:garden-path}
\mention{The horse raced past the barn fell.}\ixlq{horse}}
\z

This is grammatical~-- a reduced relative clause (\mention{the horse} [that was] \mention{raced past the barn fell}\ixlq{horse}). But the processing system commits to the main-clause parse (\mention{the horse raced past the barn}\ixlq{horse}) and crashes when \mention{fell}\ixlq{fell} appears. The detector\ixsq{detector} reports ungrammaticality, but the structure is fine.

The converse is the \term{comparative illusion\ixsq{comparative illusion}}~-- sometimes called the \term{Escher sentence\ixsq{Escher sentence}} \citep{Wellwood2018}:

\ea[]{\label{ex:15:comparative-illusion}
\mention{More people have been to Russia than I have.}\ixlq{people}}
\z

Each local chunk parses. \mention{More people have been to Russia}\ixlq{people}~-- fine, and it sets up an expectation for a \mention{than}-PP complement. \mention{Than I have}~-- fine as a \mention{than}-PP with a comparative clause. But the whole doesn't compose cleanly: \mention{more} sets up a quantity comparison, but \mention{than I have} doesn't supply a competing quantity. \mention{You've been to more countries than I have} works because both sides supply a number of countries; the operator binds a variable on each side. Here, the two sides of the comparison aren't commensurable~-- no well-formed operator-variable relation links them \citep{Wellwood2018}. For many speakers, the sentence feels both meaningful and acceptable~-- but the meaning dissolves under inspection. The illusion arises because local constraints are satisfied even though the global comparison is ill-formed.

These illusions are exactly analogous to visual illusions. The Müller-Lyer lines are the same length, but the perceptual system reports otherwise. The illusion shows perception is a mechanism with characteristic failure modes. \ixnq{Phillips, Colin}\ixnq{Wagers, Matthew W.}\ixnq{Lau, Ellen F.}\textcite{phillips2011grammatical} develop the point systematically, showing comprehension exhibits \term{selective fallibility\ixsq{selective fallibility}}~-- impressive accuracy for some constraints, systematic vulnerability for others. The pattern reveals the architecture: a retrieval system that's structure-sensitive when structural cues have temporal priority, and error-prone when they don't.


If the detector has characteristic failure modes, it should also have characteristic registers~-- layers that dissociate in predictable ways. One useful decomposition distinguishes three registers in ordinary grammatical experience.

\subsection{Three registers}
\label{sec:15:three-registers}

The core phenomenological datum is pre-reflective disruption. Around that core, two downstream layers matter for behavior and stabilization. Together, these three registers map onto the triadic architecture of §\ref{sec:14:mediation}. In Peircean terms, they align with emotional, energetic, and logical interpretants, respectively.

First, a qualitative jolt: a pre-reflective sense that something's off, the way a sour note feels sour before you know which note it is. This is the sign-vehicle doing its work~-- the perceivable form registers as wrong before you can say why.

Second, reactive resistance: the hesitation, the repair, the sense that you \emph{can't} say it that way. This is the object\ixsq{object} pushing back~-- the brute constraint on what the form can mean. The felt result is blockage, not mere surprise.

Third, normative correction: the stabilized habit\ixsq{habit} that makes the correction count as a correction. The fact that \mention{I'm twenty years old} is the right repair for \ungram\mention{I have twenty years} is not arbitrary; it's the interpretant\ixsq{interpretant}~-- the would-be\ixsq{would-be}~-- reasserting itself. Encounter this construction and you \emph{would} get this structural meaning\ixsq{structural meaning}. When the would-be\ixsq{would-be} misfires, the habit\ixsq{habit} demands its due. The correction counts as a correction because both speaker and hearer are projecting from a shared form--value\ixsq{value} coupling (§\ref{sec:15:projectibility}).

When the coupling fails, these registers often co-activate, but the readout can be partial and dissociable: jolt without stable correction, correction without strong jolt, or persistent blockage with weak explicit norm talk.

The illusions introduced in §\ref{sec:15:illusions} make the dissociation concrete. \mention{The horse raced past the barn fell} triggers a hard jolt~-- processing crashes~-- but no normative correction, because the sentence is grammatical; there's nothing to fix, so only the first register fires. The comparative illusion (\mention{More people have been to Russia than I have}) passes local fluency checks without a jolt, yet resists stable interpretation; speakers sense something is off only when pressed, and can't converge on a repair, so correction stays silent. Now compare \ungram\mention{The dogs runs}: jolt, blockage, and immediate convergent repair (\mention{The dogs run})~-- all three registers fire. The pattern of which registers activate and which stay silent is the empirical signature.

Table~\ref{tab:15:three-register-operational} maps these dissociations to timescales, observable markers, confounds, and falsifiers. If the three-register readout is right, we should expect patterned misfires rather than random noise.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.13\textwidth}>{\raggedright\arraybackslash}p{0.12\textwidth}>{\raggedright\arraybackslash}p{0.21\textwidth}>{\raggedright\arraybackslash}p{0.17\textwidth}>{\raggedright\arraybackslash}p{0.20\textwidth}@{}}
\toprule
\textbf{Register} & \textbf{Timescale} & \textbf{Observable marker} & \textbf{Main confound} & \textbf{Falsifier} \\
\midrule
Qualitative jolt & 100--800 ms & Early disruption signatures (RT, first-pass regressions, EEG perturbation) & High surprisal without structural mismatch & No early disruption above lexical baseline \\
Reactive resistance & 0.5--5 s & Hesitation, restart, and self-repair under mismatch manipulations & General disfluency and task pressure & Repair and hesitation do not increase under controlled mismatch manipulations \\
Normative correction & Turns / years & Preferred repair choices (interactional) and cohort- or register-structured correction norms (sociohistorical) & Schooling or prescriptive monitoring independent of entrenchment & No stable cohort or register structure once exposure is controlled \\
\bottomrule
\end{tabular}
\caption{Operational readout for the three-register proposal.}
\label{tab:15:three-register-operational}
\end{table}

The dissociations in Table~\ref{tab:15:three-register-operational} are predictions, not established findings; the triadic mapping is speculative architecture~-- a framework for organising observations that the chapter does not claim to have confirmed. The table specifies what evidence would confirm or falsify it.

Peirce called the felt disequilibrium that forces inquiry \enquote{doubt}~-- not sceptical doubt, but an irritation that demands repair \ixnq{Peirce, Charles Sanders}\citep[CP~5.374]{peirce1931cp}. Grammatical doubt works the same way. The would-be\ixsq{would-be} has misfired; the habit\ixsq{habit} is unsatisfied; the feeling is what unsatisfied habits feel like from the inside. It forces repair because the system won't settle until the coupling closes. That specificity~-- morphosyntactic coupling, not general appropriateness~-- is what the next section develops.


\section{What doesn't count}
\label{sec:15:negative-space}

The form--value\ixsq{value} coupling that grammaticality\ixsq{grammaticality} mostly tracks is morphosyntactic, and this predicts what won't trigger ungrammaticality. A phonetic error ([\ipa{fɪlɔzəfi}] for \mention{philosophy}\ixlq{philosophy}), a lexical error (saying \mention{airport}\ixlq{airport} when you meant \mention{train station}), and a semantic anomaly (\mention{Colorless green ideas sleep furiously}) can each misfire without being ungrammatical\ixsq{ungrammatical}.

These violations trigger different responses. Phonetic errors trigger \mention{what?} because the phoneme\ixsq{phoneme}'s core value\ixsq{value} is contrast\ixsq{contrast}, and when contrast breaks, comprehension stalls. Lexical errors trigger correction or confusion, and semantic anomaly triggers puzzlement. Even when the clash feels semantic, what matters is whether it's inside the morphosyntactic coupling. \ungram\mention{I have twenty years} jolts because the structural meaning the coupling delivers~-- possession of a duration~-- conflicts with the intended age-predication; the mismatch is morphosyntactic, not lexical. \mention{Colorless green ideas sleep furiously} doesn't jolt because the morphosyntactic coupling is intact; the absurdity is lexical. Prototypically, non-morphosyntactic violations don't trigger the specific feeling of ungrammaticality\ixsq{feeling of ungrammaticality}~-- the sense the wiring is broken.

The allomorphy of the indefinite article confirms the boundary by straddling it. English speakers reject \ungram\mention{a apple} and \ungram\mention{an banana} with the same jolt they bring to syntactic violations. The conditioning is phonological~-- the following sound determines the choice~-- but the selection rule is morphosyntactic, entrenched as part of the zipper\ixsq{zipper}. A mispronunciation of \mention{banana}\ixlq{banana} triggers \mention{what?}; distributing the wrong allomorph triggers the mismatch signal.

\subsection{Value at every grain}
\ixsq{value}
\label{sec:15:value-every-grain}

What's \enquote{value\ixsq{value}} at each level of linguistic structure? The zipper\ixsq{zipper} closes differently depending on what you're zipping (Table~\ref{tab:form-value-grains}).

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}l>{\raggedright\arraybackslash}p{0.3\textwidth}>{\raggedright\arraybackslash}p{0.45\textwidth}@{}}
\toprule
\textbf{Grain} & \textbf{Form} & \textbf{Value} \\
\midrule
Phoneme & Sound segment & Contrast (minimal-pair distinctiveness) \\
Morpheme & Affix, stem & Grammatical or lexical contribution (tense, polarity, category change) \\
Word & Phonological shape & Lexical meaning + category membership \\
Construction & Syntactic pattern & Structural meaning (participant roles, information packaging) \\
Clause type & Structural configuration & Illocutionary force (statement, question, command) \\
\bottomrule
\end{tabular}
\caption{Form--value coupling at different grains of linguistic structure.}
\label{tab:form-value-grains}
\end{table}

At the sub-word levels, value\ixsq{value} reduces to its simplest forms. For phonemes, it's nothing \emph{but} contrast\ixsq{contrast}: the /p/ in \mention{pat}\ixlq{pat} contributes no semantic content but distinguishes the word from \mention{bat}\ixlq{bat}, \mention{cat}\ixlq{cat}, \mention{sat}\ixlq{sat}. When contrast\ixsq{contrast} fails~-- when you can't tell which phoneme\ixsq{phoneme} was intended~-- you say \mention{what?} For morphemes, value\ixsq{value} is the contribution to word meaning or grammatical function: the \mention{-ed}\ixlq{-ed} in \mention{walked}\ixlq{walked} signals past tense, and the \mention{un-}\ixlq{un-} in \mention{unhappy}\ixlq{unhappy} reverses polarity. Break the coupling~-- use \mention{-ed}\ixlq{-ed} on a noun, or \mention{un-}\ixlq{un-} on a verb\ixsq{verb!distributional profile} that doesn't license it~-- and ungrammaticality results.

At the word level, value\ixsq{value} is lexical meaning plus category\ixsq{category!membership graded} membership: \mention{dog}\ixlq{dog} contributes dog-ness \emph{and} nounhood\ixsq{nounhood}, so using a word in the wrong category slot produces ungrammaticality rather than mere confusion. At the constructional level, value\ixsq{value} is the structural meaning\ixsq{structural meaning} the pattern contributes~-- the double-object construction\ixsq{construction} (\mention{gave her the book}) signals transfer plus affected recipient; the \term{passive} (\mention{was eaten}) demotes the agent~-- and breaking the construction\ixsq{construction}'s slot-filling requirements is what stops the zipper\ixsq{zipper} from closing. The English age construction (\mention{be} + \mention{years old}) is another: form maps to \enquote{predicate age,} and substituting the French \mention{avoir} pattern breaks the coupling. At the clause-type level, value\ixsq{value} is illocutionary force: \term{declarative}\ixsq{declarative} structure signals assertion, \term{interrogative} structure signals question, and \term{imperative}\ixsq{imperative} structure signals command. Rising intonation on a declarative to ask a question~-- \mention{You're coming?}~-- isn't a mismatch; it's a licensed coupling where intonation and word order both carry structural meaning and the system resolves the tension.

This table helps explain why grammaticality\ixsq{grammaticality} is primarily diagnosed at the morphosyntactic grain~-- the level where form--value\ixsq{value} couplings are tight, obligatory, and enforced.

The tightness reflects what morphosyntactic contrasts are like. They're predominantly discrete~-- maintained that way by paradigmatic acquisition, agreement systems, and deep entrenchment~-- so violations register as kind-mismatches, sharp and salient. Where morphosyntactic contrasts are less entrenched, the signal is correspondingly gradient: \mention{Who do you think that left?} and \mention{She donated the hospital the money} produce graded acceptability, not the crisp rejection that \ungram\mention{The dogs runs} does. And phonological violations can be just as sharp when comparably entrenched: try starting an English word with /\ipa{ŋ}/, a gap maintained across every token in the language. The sharpness tracks the maintenance profile, not the domain, which means gradient judgments are evidence of the coupling under strain.

Phonotactics and lexical semantics are parallel HPC kinds, maintained by the same mechanism types as grammaticality~-- acquisition, alignment, consolidation, prestige selection, transmission~-- under different constraint regimes. What changes isn't the mechanism but the constraint profile: articulatory-perceptual for phonotactics, conceptual compatibility for lexical semantics, morphosyntactic for grammaticality\ixsq{grammaticality}. This is why interventions don't cross regimes: careful articulation ameliorates phonotactic oddness but doesn't license \ungram\mention{Which did you buy car?}; structural priming improves grammatical acceptability\ixsq{acceptability} but doesn't fix /bn/. And it predicts porosity at the interfaces: because the same mechanism types operate across regimes simultaneously, interactions where regimes share territory are expected. The \mention{a}/\mention{an} allomorphy~-- phonotactic pressure within a grammatical obligation~-- is one such footprint.

\subsection{Parallel couplings}
\ixsq{grammaticality}
\label{sec:15:appropriateness}

Speakers simultaneously maintain couplings between phonetic realization and social identity (accent), between lexical choice and situational stance (register\ixsq{register}), between utterance and communicative fit (pragmatic felicity). Grammaticality\ixsq{grammaticality} is one such coupling~-- prototypically, the morphosyntactic one between form and structural meaning\ixsq{structural meaning}~-- semi-independent of the others, with its own mechanisms, its own detector\ixsq{detector}, and its own failure phenomenology (Table~\ref{tab:appropriateness-channels}).

Operationally, a failure counts as ungrammatical\ixsq{grammaticality} only if two tests pass. First, the minimal repair that preserves intended message requires a morphosyntactic change, not just lexical substitution, accent shift, or register adjustment: in standard English, \ungram\mention{She don't like it} demands \mention{doesn't}~-- a morphosyntactic swap~-- so it's grammaticality; \mention{He utilized the restroom} demands \mention{used}~-- a lexical swap~-- so it's register. Second, speakers in the relevant discourse community\ixsq{discourse community} show convergent repair preferences across contexts: English speakers converge on \mention{doesn't} for the first case; reactions to \mention{utilized} scatter by register and taste, with no stable repair norm. If either test fails, classify the case under another channel.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Domain} & \textbf{Form (prototypical)} & \textbf{Value (prototypical)} \\
\midrule
Grammaticality & Morphosyntax & Structural meaning \\
Register fit & Lexical/syntactic choice & Situational stance \\
Accent & Phonetic realization & Social identity \\
Lexical precision & Word choice & Domain membership, exactness \\
Pragmatic felicity & Utterance in context & Communicative fit \\
\bottomrule
\end{tabular}
\caption{Parallel form--value couplings at different grains.}
\label{tab:appropriateness-channels}
\end{table}

The other channels have different phenomenologies. Register\ixsq{register} mismatch triggers the raised eyebrow, the red ink on contractions, the sense that something's \enquote{off} without being \emph{ungrammatical\ixsq{ungrammatical}}. Accent mismatch triggers identity attribution: \enquote{that's not how we say it here.} Lexical inappropriateness triggers precision failure: \enquote{that's not quite the right word.}

The channels aren't orthogonal. A single utterance is evaluated simultaneously on all of them. And they interact: a register\ixsq{register}-inappropriate form can \emph{feel} ungrammatical\ixsq{ungrammatical} (through prescriptive training or hypercorrection) even when the morphosyntax is impeccable. But they're semi-independent: you can have perfect grammaticality\ixsq{grammaticality} with terrible register\ixsq{register} fit (\mention{I would like to utilize the restroom facilities} is grammatical but register\ixsq{register}-weird in casual speech), and native speakers can have strong grammaticality\ixsq{grammaticality} intuitions with weak register\ixsq{register} intuitions in unfamiliar domains.

This is where the entrenchment story connects to the social, because entrenchment is always entrenchment \enquote{within a discourse community\ixsq{discourse community}}. The grammar\ixsq{grammar} isn't floating free; it's tethered to the people who use it. When the community shifts~-- through contact, migration, register change~-- the entrenchment shifts with it. What feels ungrammatical\ixsq{ungrammatical} in one community may be unremarkable in another. And speakers flow between communities~-- sometimes instantaneously. \mention{Do you lift?} licenses intransitive \mention{lift}\ixlq{lift} for weightlifters but not for warehouse workers; the grammaticality\ixsq{grammaticality} of the construction depends on which community you're invoking, and you can switch in mid-conversation.

Code-switching within a clause~-- \mention{Vamos a hacer shopping}~-- is governed grammar, not error, but only within the discourse community of the bilinguals who maintain it. The detector\ixsq{detector} isn't calibrated once and fixed; it recalibrates to the current interlocutor and the community they invoke.

Chapter~\ref{ch:social-stabilization} develops the social mechanism proper~-- how coordination benefits, prestige asymmetries, and differential uptake aggregate local corrections into community-wide convergence~-- and shows how dialect\ixsq{dialect}, register\ixsq{register}, and discourse community\ixsq{discourse community} act as conditioning variables that select which couplings apply in a given context.


\section{Where the coupling breaks down}
\label{sec:15:breakdown}

Illusions (§\ref{sec:15:illusions}) are detector malfunctions: the coupling is intact but the readout misfires. Coupling breakdown is a different phenomenon. The maintenance itself weakens, and the form--value--interpretant\ixsq{interpretant} relation degrades. The two have different timescales, different signatures, and different consequences. Three patterns of breakdown recur, each with a distinct diagnostic profile:
\begin{itemize}
\item \textbf{Coupling thinning}~-- entrenchment weakens over generations; the form may persist with shifted indexical value (\mention{whom}).
\item \textbf{Community divergence}~-- the same form--value space is carved differently by different maintenance regimes; no coupling has broken, because the couplings were never shared (embedded inversion, double modals).
\item \textbf{Contact interference}~-- competing maintenance regimes coexist in one speaker; the detector fires for one community's norm while another community's coupling is intact (age expressions in bilinguals).
\end{itemize}

The clearest signature of the first pattern is cohort-structured change. §\ref{sec:15:detector} introduced the \mention{whom}\ixlq{whom} retreat as a diagnostic case: younger speakers don't just tolerate \mention{who}\ixlq{who} in object position; they stop correcting to \mention{whom} altogether. But the story isn't simply that the coupling thins and dies. As the grammatical coupling weakens~-- object case on the relative/interrogative pronoun, maintained through correction and uptake~-- \mention{whom} accumulates indexical value on a different channel: formality, education, prescriptive compliance. The form sits at the intersection of two couplings from Table~\ref{tab:appropriateness-channels}: grammaticality (case marking) and register\ixsq{register} (social stance). The retreat is asymmetric across them. The case coupling thins; the register coupling thickens. A speaker who uses \mention{whom} in casual speech now signals something about herself that she wouldn't have signalled a century ago, when the form was unmarked. The grammatical would-be\ixsq{would-be} (\enquote{encounter this form and you would get object case}) is fading; the indexical would-be (\enquote{encounter this form and you would get educated formality}) is strengthening. The loss is also asymmetric across constructions: \mention{whom} in subject position (\ungram\mention{Whom left?}) was never entrenched and has no coupling to lose, while \mention{whom} in prepositional complements (\mention{to whom}) retains stronger maintenance through formulaic chunks. The coupling doesn't evaporate uniformly; it frays where maintenance is thinnest, and it persists where a second channel supplies independent support~-- which is what HPC predicts and essentialism doesn't.

Community divergence produces a different pattern. The embedded inversion of §\ref{sec:15:looking-forward} (\mention{I don't know what is it}) isn't a coupling weakening~-- it's a coupling that was never established in some communities and is robustly maintained in others. The same form--value space is carved differently by different maintenance regimes. Double modals show the same structure: \mention{might could} is productive and entrenched for some Southern US speakers, flatly ungrammatical for others. The coupling hasn't broken; it was built differently.

A third pattern is interference under contact. The age expression (\ungram\mention{I have twenty years}) illustrates this for individual speakers: a French-dominant bilingual has an entrenched \mention{avoir}-age coupling that competes with the English \mention{be}-age coupling. The two maintenance regimes coexist in the same speaker, and the detector\ixsq{detector} can fire for one community's norm while the other community's coupling is fully intact. This is defeasibility\ixsq{defeasibility} at the level of the whole system: the projections that grammaticality supports can be defeated not only by processing overload (illusions) but by competing maintenance regimes (contact), thinning entrenchment (language change), or community divergence (dialect formation).

What unites these cases is that the breakdown is informative. Where a coupling frays, the pattern of fraying reveals the maintenance structure~-- which strands bore the load, which were parasitic on others, which had independent support. The failure modes are diagnostic, not noise. This is why §\ref{sec:15:dual-role} can treat grammaticality as both a category and a mechanism: the same coupling that projects also breaks down in structured ways, and the structure of the breakdown is evidence for the maintenance account.


\section{Grammaticality as mechanism and category}
\ixsq{grammaticality}\ixsq{mechanism!vs.\ essences}\ixsq{category!as HPC kind}
\label{sec:15:dual-role}

Part~III showed that individual categories~-- countability, definiteness, lexical categories, gender~-- are HPC kinds maintained by identifiable mechanisms. But grammaticality as a property of the \emph{system} is a further claim. System-level properties~-- that the same locality constraints govern extraction across construction types, that combining violations produces penalties exceeding the sum of each~-- aren't recoverable from any single construction description.

A concrete case: \marg\mention{Who did you wonder whether left?} violates an island constraint; \ungram\mention{The dogs runs} violates agreement. Each alone produces a predictable penalty. Combine them~-- \ungram\mention{Who did you wonder whether runs?}~-- and impressionistic reports suggest the result feels disproportionately worse~-- as if the system's error-correction resources have been simultaneously overwhelmed at two independent points. If the super-additive interaction holds up under controlled testing, it's a system-level property, not a construction-level one. If the reader struggled to parse the combined violation, the struggle is the datum.

This system-level HPC\ixsq{homeostatic property cluster!grammaticality} plays a dual role. As a \term{category\ixsq{category!as HPC kind}}, it sustains a would-be\ixsq{would-be}: speakers deploy form--value\ixsq{value} pairings in novel sentences, extend them by analogy\ixsq{analogy} to new verbs, and repair deviations by projecting from the known coupling to the correct form (§\ref{sec:15:projectibility}). Grammaticality\ixsq{grammaticality} is a category\ixsq{category!as HPC kind} \emph{because} it projects, not merely because it's maintained.

As a \term{mechanism\ixsq{mechanism!defined}}, grammaticality judgments and repair routines are a major stabilizer\ixsq{stabilizer!as mechanism role} for other HPCs. The category \term{noun} is maintained partly because using a word in a noun slot, with noun morphology, in noun constructions, triggers no mismatch signal. The detector\ixsq{detector} patrols the couplings that constitute other categories~-- but the detector is not the coupling itself. And the dual role is historically grounded in the same way as §\ref{sec:15:detector}'s reflexive loop: the detector patrols couplings that exist because past patrolling maintained them. Grammaticality\ixsq{grammaticality} as a system-level maintained property is one thing; the judgment and repair routines that help maintain it (and other categories) are another.

The bootstrapping question~-- how does the detector develop before the coupling exists?~-- has the same shape in every HPC domain. Non-grammatical anchors (distributional frequency, prosody, semantic plausibility) provide initial traction; the coupling and the detector co-develop as entrenchment deepens. The circularity is developmental, not logical.

No Platonism required; just a self-maintaining target state, which is what \term{homeostatic} means. This changes the evidential status of the asterisk: if grammaticality\ixsq{grammaticality} is a maintained coupling and acceptability\ixsq{acceptability} is a noisy detector\ixsq{detector}, the asterisk is a signal from a measuring instrument~-- informative, but fallible, and calibrated to the community in which it was trained.


\section{Predictions and evidence}
\label{sec:15:predictions}

If grammaticality is a maintained coupling that supports projection, the maintenance account generates predictions that pure mechanism stories don't. The predictions below vary in evidential status: some rest on existing experimental work; others are forward-facing research commitments. Three stand out: that satiation should be construction-specific, tracking entrenchment depth rather than general familiarity; that productivity asymmetries should track coupling strength, with tight couplings projecting further than loose ones; and that repair should fire pre-emptively, before production completes, when the would-be\ixsq{would-be} conflicts with the form under construction. There is meaningful evidence bearing on the first two. The third is documented in §\ref{sec:15:projectibility} and awaits systematic testing.

\subsection{Satiation}
\label{sec:15:satiation}

\term{Syntactic satiation\ixsq{syntactic satiation}} is the finding that repeated exposure to a marginally acceptable sentence can improve its perceived acceptability\ixsq{acceptability} \ixnq{Snyder, William}\parencite{snyder2000grammaticality}. But the effect is selective. Some constraints satiate readily; others resist entirely. In current English judgment datasets, left-branch extraction (\ungram\mention{Which did you buy car?}\ixlq{car}) remains near-categorically rejected even under exposure manipulations \ixnq{Snyder, William}\parencite{Snyder2022}. The intended meaning is easily grasped~-- \mention{Which car did you buy?}\ixlq{car}~-- and the structure is not unusually complex. The resistance has appeared across multiple tasks and speaker samples, though not uniformly in every paradigm; a meta-analysis by \textcite{lufrankdegen2024} provides independent confirmation.

In the Bayesian terms of §\ref{sec:15:detector}, satiation is prior-updating. Each repeated exemplar shifts the posterior upward; the mismatch signal weakens; the sentence feels better. Anyone who has marked student essays in bulk has watched their own standards quietly erode by page thirty. The updating is easiest when the form fills a value niche that no entrenched competitor already occupies~-- when there's a gap in the distribution rather than a crowded field. An analogy from a neighbouring domain suggests the updating can go deeper than the distribution widening. \ixnq{Vagnino, Richard}\ixnq{Walker, Caren}\textcite{vagninoWalker2026}, studying abstract schema drift in relational concepts under repeated comparison (not syntax), show that repeated exposure to diverse exemplars causes the abstract schema itself to drift, accommodating new forms. They term this \term{schema drift\ixsq{schema drift}}, contrasting it with \term{rerepresentation\ixsq{rerepresentation}} (changing the input to fit the category\ixsq{category!boundary cases}): \enquote{instead of changing the square peg to better fit the round hole... the hole might also be altered to better fit the peg.} Whether syntactic entrenchment shows the same schema-level plasticity is an open question, but the candidate mechanism~-- category structure accommodating exemplars, not just tallying them~-- is the right shape for a satiation account.

But ten repetitions can't shift the posterior far when entrenched competitors dominate the distribution. Left-branch extraction is a \term{conspicuous absence\ixsq{conspicuous absence}}: the likelihood isn't zero~-- fronting is productive in English, and the \emph{potential} for extraction is visible~-- but the community prior is actively low, and entrenched \term{workarounds\ixsq{workaround}} hold high posteriors: pied-piping (\mention{whose book}), fused-head constructions (\mention{whose is this}), the \mention{big mess} construction (\mention{how good a teacher}). The detector\ixsq{detector} isn't saying \enquote{this doesn't happen}; it's saying \enquote{we could, but we don't.} \ixnq{Reynolds, Brett}\textcite{reynolds2026lbe} (author's unpublished manuscript) argues that this density of alternatives makes the gap resistant to short-run satiation.

The age construction offers a test (Table~\ref{tab:15:satiation-prediction}).

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lll@{}}
\toprule
 & \textbf{Left-branch extraction} & \textbf{Age construction} \\
 & \ungram\mention{Which did you buy car?} & \ungram\mention{I have twenty years} \\
\midrule
English monolingual & Resists satiation & Resists satiation \\
French-dominant bilingual & Resists satiation & Satiates more readily \\
\bottomrule
\end{tabular}
\caption{Predicted satiation asymmetry. Both constructions are conspicuous absences for English monolinguals. For French-dominant bilinguals, the age construction has a competing entrenched coupling (\mention{avoir}); left-branch extraction does not.}
\label{tab:15:satiation-prediction}
\end{table}

For English monolinguals, \ungram\mention{I have twenty years} is a conspicuous absence: the competitor posterior (\mention{I'm twenty years old}) is high, so exposure barely shifts the ranking. For French-dominant bilinguals, the prediction reverses~-- their prior is tuned to the \mention{avoir} coupling, so the English \mention{have} form should satiate more readily. The asymmetry is a prediction about prior distributions, not about \enquote{knowing the rule.}

\subsection{Productivity asymmetries}
\label{sec:15:productivity-asymmetries}

The second prediction is that tight couplings should project further than loose ones. Subject--verb agreement extends productively to novel combinations: any new verb entering English immediately takes \mention{-s} in the third-person singular present. The coupling is tight~-- paradigmatic, obligatory, and entrenched across virtually every utterance~-- and the projection is correspondingly robust. Contrast island constraints\ixsq{island constraint}: extraction from relative clauses (\marg\mention{What did you meet the person who fixed?}) is licensed for some speakers in some constructions but not others, and judgments scatter. The coupling is loose~-- construction-specific, variable across communities, sensitive to lexical and pragmatic conditioning~-- and the projection is correspondingly narrow.

The prediction is that this asymmetry should be systematic, not accidental. Coupling strength (measured by entrenchment depth, cross-construction consistency, and community convergence) should predict the range of novel instances to which a form--value pairing extends. Tight couplings generate strong would-bes\ixsq{would-be} that extend to unfamiliar territory; loose couplings generate weak would-bes that falter at the edges. §\ref{sec:15:projectibility} documented this informally; the claim here is that it should hold quantitatively, with coupling strength predicting productivity range across construction types.

\subsection{Individual differences}
\label{sec:15:individual-differences}

A further prediction: if projectibility is indexed to a community's entrenchment, speakers with different exposure histories should project differently~-- not just accept different sentences, but generate different analogical extensions and different repair preferences.

Speakers differ systematically in their grammaticality\ixsq{grammaticality} thresholds. The limiting case is trivial: we have no feeling of grammaticality whatsoever for languages we don't speak. A sentence in Zulu\ixgq{Zulu} or Finnish produces no mismatch signal for me because the prior is zero~-- no entrenched pairings, no distribution to compute a posterior from. Foreign language learners have thin priors: their posteriors are unstable, their detector\ixsq{detector} output unreliable, slow to fire, and easily fooled by surface plausibility. Only with sufficient exposure does the prior thicken and the feeling sharpen.

Within a single language community, the variation persists. \ixnq{Dabrowska, Ewa}\textcite{dabrowska2012} shows that grammatical knowledge is not uniform across the adult population: speakers with less formal education, less exposure to complex syntactic constructions, and less metalinguistic training produce reliably different acceptability\ixsq{acceptability} judgments~-- not just noise, but systematic divergence. \textcite{dabrowska2018} links the differences to exposure, education, IQ, and aptitude, showing that what looks like a uniform endpoint of acquisition is in fact shaped by experience throughout the lifespan.

That's the prediction from community-indexed entrenchment\ixsq{entrenchment!frequency effects}. If grammaticality\ixsq{grammaticality} is what a community has entrenched, then members with different exposure histories will have different thresholds. Chapter~\ref{ch:words-wont-hold-still}'s \mention{think myself through that argument}~-- interpretable by analogy to \mention{talk oneself through} and \mention{work oneself through}, but rejected by some speakers~-- is exactly this: the detector\ixsq{detector} fires or doesn't depending on how richly the analogical base has been entrenched. The same logic applies to age expressions: a French-educated English speaker may accept \ungram\mention{I have twenty years} where a monolingual wouldn't, because her prior for the \mention{have}-age pairing isn't zero~-- the \mention{avoir} coupling already fills the value niche.

This raises the question of grain\ixsq{community!grain of}. At what scale is \enquote{community} projectible? The answer is empirical, but it isn't empty. Convergent entrenchment has operational signatures: speakers converge on the same repairs across contexts, produce similar distributions given similar opportunities, and share would-bes\ixsq{would-be} that fire for both parties in interaction. Where those diagnostics align, the coupling is stable and the category projects. Where they diverge, you have either nested communities or genuinely different kinds~-- and the maintenance structure tells you which. This is the same individuation problem every HPC domain faces: species concepts, psychiatric nosology, and ecological communities all draw boundaries wherever the mechanisms produce convergent outcomes, and all tolerate nesting and overlap. The boundaries of a grammatical community are wherever the maintenance mechanisms produce convergent entrenchment, which means they can nest and overlap. Double modals illustrate the nesting: \mention{might could} is flatly ungrammatical for non-Southern US speakers~-- they can't even parse it~-- but grammatical and productive for millions of Southern speakers. Within Southern English, the boundaries nest further: some speakers have \mention{might could} but not \mention{might should}; urban speakers accept fewer combinations than rural ones. The coupling is maintained at each level by a different population with different exposure histories.

\subsection{Relevance and maintenance}
\label{sec:15:relevance}

\ixnq{Scott-Phillips, Thom}\textcite{scottphillips2024communication} offers a complementary account. On his view, grammaticality\ixsq{grammaticality} intuitions are byproducts of relevance-tracking: we detect when utterances violate basic presumptions about communicative efficiency. Unacceptability arises not from mere inefficiency, but from an inherent impossibility of interpreting an utterance consistently with those presumptions. \textcite{scottphillips2025intuitions} develops the point further, arguing that acceptability intuitions specifically are byproducts of the capacity for ostensive-inferential communication.

The maintenance framework shares several premises: both reject an innate grammar faculty; both locate intuitions within general cognitive systems; both recognize that communicative pressures shape conventions. And relevance theory supplies something the maintenance account lacks on its own: an explanation of \emph{why} grammaticality intuitions exist~-- they're recruited from a general capacity for tracking communicative fit, not purpose-built for syntax. The frameworks differ in granularity. Relevance theory predicts broader failures when relevance presumptions collapse; the maintenance account predicts construction\ixsq{construction}-specific failures where entrenched pairings misalign. On the maintenance view, \ungram\mention{I have twenty years} fails as an age statement because the English age construction is \mention{be} + \mention{years old}, not because relevance-seeking fails in general. The three predictions above~-- construction-specific satiation, coupling-indexed productivity, pre-emptive repair~-- are projectibility predictions, not just mechanism descriptions. The satiation case is where the frameworks diverge most sharply. Relevance theory predicts that repeated exposure to \emph{any} relevance violation should produce habituation, since the system recalibrates to the lower relevance baseline. The maintenance account predicts construction-specific satiation: some violations satiate readily, others resist entirely, depending on entrenchment depth and competitor density (§\ref{sec:15:satiation}). The satiation data bear on this: if satiation is construction-specific rather than a general relevance-recalibration effect, the maintenance account has the finer grain.


\subsection{The HPC audit}
\ixsq{homeostatic property cluster!two-diagnostic test}
\label{sec:15:hpc-audit}
§\ref{sec:8:diagnostics} introduced the Two-Diagnostic Test\ixsq{two-diagnostic test}, and grammaticality\ixsq{grammaticality} passes both: it sustains a would-be\ixsq{would-be}, since a mastered form--value\ixsq{value} pairing projects to novel instances (§\ref{sec:15:projectibility}), and it's homeostatic, since the feeling of ungrammaticality\ixsq{feeling of ungrammaticality} stabilizes the coupling through hesitation, repair, and community-level norm enforcement (§\ref{sec:15:detector}).

Unpacking the cluster sharpens the claim. Three properties co-occur systematically while remaining partially independent: the zipper\ixsq{zipper} itself (morphosyntactic form, structural value\ixsq{value}, and the interpretant\ixsq{interpretant} that binds them), compatibility between those pairings and contextual interpretation, and community entrenchment\ixsq{entrenchment!defined}. The co-occurrence isn't accidental; it's explained by the maintenance loop of §\ref{sec:15:detector}. Entrenchment keeps form--value\ixsq{value} pairs stable for individual speakers; interactive alignment keeps them shared across interlocutors; transmission through acquisition keeps them available to new cohorts. Each mechanism reinforces the others: stable pairings are easier to align on, aligned pairings are easier to transmit, and transmitted pairings deepen entrenchment. Where one leg weakens, the others follow~-- which is why the partial-independence cases below show structured degradation, not random loss.

Independent relative \mention{whose}\ixlq{whose} (§\ref{sec:6:independent-whose}) has intact pairings and clear compatibility~-- the form marks genitive relative, the interpretation is unambiguous~-- but lacks entrenchment. Corpora show vanishing frequency; workarounds dominate; the detector\ixsq{detector} flags the form as marginal despite structural integrity. Two of three present.

\ungram\mention{I've finished it yesterday} has the mirror profile~-- but the two properties present are community-relative in different ways. The form--value\ixsq{value} pairing (present perfect + past-time adverbial) is well-entrenched in learner English and many contact varieties, where it functions as a normal temporal construction. The incompatibility, however, is evaluated relative to standard English, where \mention{yesterday} selects simple past, not present perfect; that community rejects the combination even though the form--value coupling is internally coherent. The entrenchment and the incompatibility belong to different communities, and the \mention{ungrammatical} verdict is indexed to the standard-English one. Again, two of three present~-- and that partial independence is the signature of an HPC\ixsq{homeostatic property cluster!grammaticality} kind.

Cross-linguistic neural evidence is consistent with the projectibility\ixsq{projectibility!induction from} claim, though it does not by itself establish maintenance. Agreement violations elicit robust late positivity/P600-type responses not only in Germanic and Romance languages but also in Hindi\ixgq{Hindi} \parencite{Nevins2007}, Basque\ixgq{Basque} \parencite{Zawiszewski2009}, and Arabic\ixgq{Arabic} \parencite{MuralikrishnanIdrissi2021}. Earlier negativities also appear in some of this work, but the later positivity is the more stable cross-linguistic signature. The convergent detection signature across genetically unrelated languages is consistent with grammaticality being a projectable kind~-- it establishes a common processing profile rather than the full maintenance-plus-projection claim, but it rules out the possibility that the pattern is an artefact of English-specific entrenchment.

On current evidence, then, grammaticality\ixsq{grammaticality} is an HPC kind~-- maintained by the same mechanism \emph{types} that Part~III documented for countability\ixsq{countability}, definiteness, gender, and lexical categories (entrenchment, alignment, transmission), though with its own specific instances, and serving as stabilizer for all of them. The claim is falsifiable. Three patterns would undermine it: if structure-based predictions failed on new construction types while construction-specific labels succeeded; if cross-construction interactions turned out to be simple sums of independent factors, eliminating the system-level claim of §\ref{sec:15:dual-role}; or if phonetic or discourse manipulations repaired structural bans as effectively as structural interventions, collapsing the regime boundaries of §\ref{sec:15:negative-space}. Chapter~\ref{ch:what-changes} develops the full gauntlet.


\section{Looking forward}
\label{sec:15:looking-forward}

The asterisk, then, is a measuring instrument~-- a signal from a noisy detector\ixsq{detector} that tracks the maintained coupling between morphosyntactic form and structural meaning. The coupling is homeostatic: perturbations trigger repair, repair restores the coupling, and the restored coupling supports further projection. The dual role~-- maintainer and maintained~-- is characteristic of homeostatic systems rather than paradoxical. But the equilibrium is also \term{contingent\ixsq{contingency}}. On strong constraint-based readings, subordinate interrogatives \emph{can't} invert in English~-- a grammatical principle reserves subject--auxiliary inversion for main-clause questions. The HPC\ixsq{homeostatic property cluster!vs.\ essentialism} account says only that the non-inverted coupling \emph{hasn't} been perturbed in most varieties under current conditions.

It has been perturbed in others. In Irish English\ixgq{Irish English}, \mention{I don't know what is it} is grammatical\ixsq{grammaticality} and productive \parencite{henry1995belfast}. The same pattern appears in African American, Appalachian, Chicano, and Newfoundland English~-- across several major dialect regions~-- and \textcite{Wolfram2015} describe it as increasingly routine in informal spoken American English. Lewis Carroll had it in 1865: \mention{Alice looked very anxiously into its face to see what was the matter with it}. What an essentialist account treats as a categorical constraint turns out to be a maintained coupling that holds in some communities\ixsq{discourse community} and not others, and the boundary is moving. If social conditions shift, priors shift, so what resists at a given moment is entrenched convention, not a timeless prohibition. The mechanism is maintenance, not law~-- but that raises the question of whether the maintenance itself is contingent.

But contingency is not arbitrariness: the claim is about the \emph{contents} of the coupling, not its \emph{existence}. That every known language maintains some form--value coupling~-- that grammaticality exists everywhere~-- is itself a convergence fact, explained by shared cognitive and communicative pressures that make projection useful. Eyes evolved independently dozens of times, not because optics is a law of biology, but because light is informative and photosensitive tissue is accessible. Grammaticality recurs for the same reason: real-time production under memory constraints, compositional communication that requires predictable form--meaning mapping, and iterated transmission through acquisition bottlenecks all make \emph{some} form--value coupling convergent. They don't dictate \emph{which} couplings. That's the distinction from parametric essentialism, which predicts universal contents (the same constraints, the same parameters, everywhere); HPC predicts universal existence with contingent contents~-- and further predicts that tight couplings (subject--verb agreement) should be more cross-linguistically stable than loose ones (island constraints), because tighter maintenance resists more perturbation. The analogy is limited: biological homeostasis operates through reproduction-bound selection, while grammatical homeostasis operates through classification-sensitive feedback (§\ref{sec:15:detector}). What carries over is the convergence logic~-- shared functional pressures producing structurally similar solutions~-- not the maintenance mechanism.

The contingency is the point, not the problem. What a community has entrenched is what it can project from. Grammaticality\ixsq{grammaticality} is the maintained coupling that makes the good bet\ixsq{good bet} possible~-- encounter this form--value\ixsq{value} pairing and you \emph{would} get this structural meaning\ixsq{structural meaning}, within a community and across novel instances. The asterisk probes that coupling~-- which is what makes grammaticality\ixsq{grammaticality} more than a feeling, and why Chapter~\ref{ch:social-stabilization} asks what keeps the coupling aligned well enough for shared projection.
