\chapter{How categories travel}
\label{ch:how-categories-travel}
\label{ch:what-changes}
% New citations needed in references-local.bib: borsboom2017, kendler2011, meehl1995
\epigraph{It is a result of the theory I am urging that the use of metaphor in the expression of our theoretical beliefs is not a sign of the infancy of the discipline in question. It is rather a natural and appropriate response to certain features of the world.}{~-- Richard Boyd, \enquote{Metaphor and Theory Change} (1979)}

If categories are maintained clusters, and two maintained clusters share enough structure, then treating one as the other generates predictions. Some of those predictions hold. Some don't. The boundary between productive and unproductive extension tracks the boundary between shared and unshared mechanisms.

That claim has been implicit since Chapter~\ref{ch:kinds-without-essences}. This chapter makes it explicit, and in doing so addresses the tradition closest to HPC\ixsq{homeostatic property cluster!vs.\ cognitive linguistics} in spirit: cognitive linguistics\ixsq{cognitive linguistics}.

\section{The nearest tradition}
\label{sec:16:nearest-tradition}

Cognitive linguistics got several things right, and it got them right early.

\ixnq{Rosch, Eleanor H.}Rosch showed that categories have internal structure: typicality gradients, prototype effects, membership that comes in degrees of centrality \citep{rosch1975}. Chapter~\ref{ch:words-wont-hold-still} took that evidence seriously. \ixnq{Lakoff, George}Lakoff extended the insight to grammar, arguing that radial categories~-- the senses of \mention{over}\ixlq{over}, the extensions of \mention{mother}\ixlq{mother}, Dixon's Dyirbal\ixgq{Dyirbal} classifiers~-- show genuine clustering rather than definitional tidiness \citep{lakoff1987}. \ixnq{Lakoff, George}\ixnq{Johnson, Mark}\textcite{lakoff1980} demonstrated that cross-domain mappings are systematic, not ornamental: we don't just \emph{talk} about time in spatial terms; we \emph{reason} about it that way, and the mappings are productive enough to generate novel inferences. \ixnq{Bybee, Joan}\ixnq{Goldberg, Adele E.}\ixnq{Croft, William}Usage-based work grounded categories in actual use rather than innate specification \citep{bybee2010,goldberg1995constructions,croft2001}. And \ixnq{Langacker, Ronald W.}Langacker's Cognitive Grammar proposed a non-modular architecture in which usage events are the primitives and abstractions are emergent \citep{langacker1987foundations}.

These are genuine contributions. The programme documented clustering that essentialism couldn't see and took seriously the mechanisms~-- frequency, analogy, embodiment~-- that produce it. On both counts it anticipated the maintenance view this book defends.

But there's a gap, and it isn't empirical. The mechanisms are there~-- and in the usage-based wing, the ontological instinct is close to right. What's missing is the apparatus for saying when a maintained cluster actually \emph{projects}~-- when knowing the category lets you predict novel instances.

Prototype theory\ixsq{prototype} made real discoveries. \ixnq{Rosch, Eleanor}\ixnq{Mervis, Carolyn B.}\textcite{roschmervis1975} showed that prototypical members have the highest cue validity~-- their features best predict category membership~-- and that family resemblance structure (overlapping features, no single feature shared by all; \ixnq{Wittgenstein, Ludwig}\citealp{wittgenstein1953}) can be quantified. Basic-level effects~-- the finding that there's a privileged grain of categorization (\textsc{dog}, not \textsc{animal} or \textsc{poodle}) where informativeness peaks \citep{rosch1976}~-- identified where category structure is richest. These are substantial achievements, and they come with real explanatory content: cue validity tells you \emph{why} some members are more central than others. But it's distributional. It tells you which features co-occur; it doesn't tell you why they \emph{keep} co-occurring across speakers and centuries. Basic-level effects identify where the privileged level sits; they don't explain what maintains it there. The gap isn't in the data or the analysis. It's in the maintenance-and-projection story: what keeps the clustering going, and what inductive work does the clustering underwrite?

If gradience were all there were, categories should drift: each generation learns from slightly different data, and small perturbations should accumulate. Chapter~\ref{ch:words-wont-hold-still} posed this as the stability problem. Cognitive linguistics proposed mechanisms that produce the gradience~-- entrenchment, frequency, analogical extension~-- but didn't develop a framework for predicting when those mechanisms keep a category stable, or when the resulting cluster projects~-- when knowing the category lets you predict things about novel instances that you couldn't predict without it.

Conceptual metaphor theory describes cross-domain mapping but doesn't explain what makes mapping \emph{possible}~-- what property of categories enables them to travel. \textcite{lakoff1980} showed that we systematically map spatial structure onto temporal reasoning. But not every domain maps onto every other. Why do some mappings work and others don't? What distinguishes a productive mapping from a dead one? Image schemas\ixsq{image schema} and experiential gestalts were serious attempts at an answer: experientially basic structures that serve as source domains. They identify what gets mapped. But they don't predict which new mappings should be productive, or explain why structural similarity licences inferential transfer in some cases and not others.

The HPC framework doesn't replace cognitive linguistics. It grounds it. Prototype effects are real because mechanisms produce and sustain them (Chapters~\ref{ch:stabilizers}--\ref{ch:failure-modes}). Radial categories are genuine clusters because the same maintenance processes~-- entrenchment, alignment, functional pressure~-- keep the extensions anchored to the core. And cross-domain mapping works when it works because of something specific about the categories involved: they share mechanism-sustained property overlap. Where the mechanisms diverge, the mapping breaks down. That's a testable claim, not a restatement.

What the cognitive linguists lacked wasn't data, ambition, or even the right ontological instinct. Bybee's entrenchment, Goldberg's constructional schemas, Tomasello's usage events all point toward a constitutive role for mechanisms~-- categories \emph{are} their usage patterns, not merely described by them. What was missing was the diagnostic toolkit: projectibility criteria that say when a cluster supports induction (Chapter~\ref{ch:projectibility}), failure modes that diagnose when it doesn't (Chapter~\ref{ch:failure-modes}), and field-relativity that explains why the same phenomena yield different categories for different purposes. You can hold the right ontology without having the means to test it.

This chapter closes it, and the closing has a payoff: cross-domain extension falls out of the framework for free~-- not as a separate cognitive capacity requiring its own explanation, but as a consequence of what categories are.

\section{Theory-constitutive metaphor and cross-domain extension}
\label{sec:16:tcm}

Richard Boyd made an argument in 1979 that the philosophy of science absorbed and linguistics largely ignored.

Metaphors in science, Boyd argued, aren't pedagogical aids~-- scaffolding that can be replaced by literal paraphrase once the theory matures. Some metaphors are \term{theory-constitutive\ixsq{theory-constitutive metaphor}}: they express causal relationships between natural kinds that no literal description yet captures \citep{boyd1979}. When a biologist calls natural selection a \enquote{pressure}, the metaphor isn't ornamental. It constitutes a claim about the dynamics of fitness landscapes that no available literal vocabulary can state as precisely. The metaphor succeeds not because source and target share surface features but because they share \emph{mechanism-sustained property overlap}. Pressure in physics and selection in biology are both maintained processes that produce directional change; the shared structure licences inferential transfer.

This is more principled than Lakoff and Johnson's account of mapping. \textcite{lakoff1980} demonstrated that cross-domain mappings are systematic~-- we don't map randomly from source to target. Their framework tracks which mappings are conventional and which are novel, and it distinguishes dead metaphors from live ones. But it doesn't predict which \emph{new} mappings should be productive, or explain why some domains map onto each other and others don't. Image schemas provide a partial answer (spatial structure is experientially basic and therefore available as a source), but they don't explain why TIME IS MONEY works while TIME IS FURNITURE doesn't, even though both furniture and money are culturally familiar domains with rich internal structure.

Boyd's answer: the mapping works to the extent that source and target are HPC kinds sharing partially overlapping mechanisms. TIME IS MONEY works because both time and money are maintained resources~-- budgeted, spent, saved, wasted~-- and the resource-management mechanisms overlap. TIME IS FURNITURE fails because furniture and time share no mechanism-sustained structure. The overlap isn't in the features per se; it's in the causal architecture that maintains the features.

Define \term{cross-domain extension}\ixsq{cross-domain extension} as follows: when two HPC kinds share partial mechanism-sustained property overlap, inferences licensed by one can be extended to the other. The extension is productive to the extent that the shared properties are maintained by shared or analogous mechanisms. This is what metaphor \emph{is}~-- not a decorative rhetorical device, not a cognitively primitive mapping, but a consequence of what categories are. If categories are maintained clusters, and two maintained clusters share enough structure, then treating one as the other generates predictions. Some hold (the metaphor works); some don't (the metaphor breaks down). The boundary between productive and unproductive extension tracks the boundary between shared and unshared mechanisms.

This yields a prediction: wherever you have HPC kinds, you should find cross-domain extension. It isn't a uniquely human cognitive capacity bolted on top of categorization; it falls out of categorization itself. The prediction is testable. Cross-domain extension should be more productive between kinds that share more mechanism-sustained structure, and it should degrade when the shared mechanisms are disrupted.

\ixnq{Noyes, Alexander}\ixnq{Keil, Frank C.}\textcite{noyes2019} provide indirect support. They show that generic statements (\mention{Dogs follow trails}\ixlq{dog}) designate kinds rather than attributing properties to individuals. In HPC terms, generics are would-be\ixsq{would-be} generators~-- they licence induction precisely because the kind is maintained. The generic projects because the mechanisms sustain the cluster. When generics cross domains (\mention{Time flies}\ixlq{time}), they carry inferential habits from the source kind, and those habits are productive to the extent that the target preserves the relevant causal structure.

\ixnq{Gentner, Dedre}\textcite{gentner1983} formalized a closely related insight with her structure-mapping theory: analogies are productive when they preserve relational structure, not when they match surface features. HPC adds the missing piece: what makes relational structure \emph{available} for mapping is that it's maintained by mechanisms. Dead metaphors die when the mechanisms diverge; live metaphors live because shared maintenance keeps the relational structure aligned.

\section{How categories travel: the trail}
\label{sec:16:trail}

To see cross-domain extension at work, consider a category that predates language by tens of millions of years.

Dogs navigate primarily by olfaction. When a person walks across a field, they deposit a scent trail: volatile organic compounds from skin, sweat glands, and the soles of shoes, supplemented by crushed vegetation and disturbed soil. The trail has structure. \ixnq{Thesen, A.}\textcite{thesen1993} showed that trained dogs can determine the direction of a human trail from as few as five footprints, using the concentration gradient~-- newer prints carry stronger scent. \ixnq{Hepper, Peter G.}\ixnq{Wells, Deborah L.}\textcite{hepper2005} replicated the finding and added that dogs discriminate trail direction even when the prints are only seconds apart, relying on decay asymmetries too subtle for human detection.

\textsc{Trail}\ixsq{trail} has HPC structure. Multiple co-occurring properties~-- directionality, recency, source identity, substrate type, concentration profile~-- cluster because physical and behavioural mechanisms maintain them. Scent marking reinforces the trail. Route reuse deepens it. Environmental persistence (moisture, temperature, substrate porosity) modulates how long the cluster holds together. The properties aren't logically connected; they co-occur because the mechanisms sustain their co-occurrence. Remove the mechanisms (dry the substrate, eliminate reuse, suppress scent marking) and the cluster degrades.

Now watch what dogs do with the category. A tracking dog trained on grass can follow the same person's trail across concrete, carpet, and leaf litter. The substrate changes; the scent chemistry changes; the concentration dynamics change. But the dog extends its trail-following strategies to the novel substrate, adjusting gait and sniffing behaviour while preserving the inferential core: directionality, source identity, recency. This is cross-domain extension on partial property overlap. The dog doesn't need a new category for each substrate. It extends an existing one because enough mechanism-sustained structure transfers.

\ixnq{Fagan, William F.}\textcite{fagan2025} push the point further. Studying GPS-tracked movement data across species, they find that wild canids show significantly stronger route reuse than felids. Wolves, coyotes, and African wild dogs return to established trails; big cats don't. The difference tracks mechanism density\ixsq{mechanism!density}: canids are social scent-markers that reinforce trails through collective use, while felids are largely solitary hunters whose movement patterns leave shallower traces. The \textsc{trail} category is more deeply entrenched in canid cognition~-- more properties cluster, more mechanisms maintain the clustering, and the category supports richer extension.

The phylogenetic\ixsq{phylogenetic depth} point follows. If cross-domain extension follows from HPC kinds, and non-human animals form HPC kinds (they do~-- maintained by the same physical and behavioural mechanisms, just without the semiotic amplification that language provides), then the capacity for something like metaphor predates language. Not the claim that dogs \enquote{think metaphorically}~-- that would be semiotic overreach. But the \emph{kind-formation capacity} that makes metaphor possible in humans is evolutionarily old. Dogs extend trail-following across substrates because the category is maintained richly enough to survive the transfer. Humans extend conceptual categories across domains because language adds a layer of semiotic maintenance~-- signs, interpretants\ixsq{interpretant}, habit-generating inference chains~-- that amplifies the same basic capacity. Language doesn't create cross-domain extension. It turbocharges it.

\section{The Boyd circle}
\label{sec:16:boyd-circle}

The argument is now reflexive, and the reflexivity is a feature.

Start the circle:

\begin{enumerate}
\item Boyd (1979) argued that theory-constitutive metaphors work because source and target are both HPC kinds sharing partial property overlap \citep{boyd1979}.
\item Boyd's HPC framework was developed for species in biology: organisms cluster because reproductive, developmental, and ecological mechanisms keep properties co-occurring \citep{boyd1991}.
\item This book applies HPC~-- originally from species~-- to grammatical categories. That application is itself a cross-domain extension on partial property overlap: species and grammatical categories share mechanism-sustained clustering, projectibility, and boundary behaviour, even though the specific mechanisms differ.
\item The book now predicts that HPC kinds should support cross-domain extension wherever the mechanism-sustained overlap is rich enough.
\item Cross-domain extension is what metaphor \emph{is}.
\item Which is what Boyd was explaining in step 1.
\end{enumerate}

This isn't vicious circularity. The theory doesn't presuppose its own conclusion. It predicts a capacity~-- cross-domain extension~-- that the theory itself exemplifies. The reflexivity is a test. If the framework couldn't account for its own method of explanation, that would be a problem: a theory of categories that can't explain why it was able to travel from biology to linguistics would be undermining its own explanatory claims. It can. The self-application passes because the property overlap between biological species and grammatical categories is genuine and mechanism-sustained~-- not a surface analogy but a structural one, grounded in shared causal architecture (maintained clustering, projectible properties, homeostatic boundaries).

The Peircean\ixsq{Peirce, Charles Sanders} vocabulary from Chapter~\ref{ch:projectibility} sharpens the point. The cable metaphor \citep[CP~5.265]{peirce1931cp} is itself a theory-constitutive metaphor in Boyd's sense. Peirce used the physical properties of cables~-- multiple partially independent strands whose collective strength exceeds any single strand~-- to constitute a claim about epistemic justification. The metaphor works because cables and arguments share HPC-like structure: maintained by distributed redundancy, resistant to local failure, degrading gracefully rather than catastrophically. This book's cable metaphor for category maintenance is a TCM in exactly Boyd's sense. And it works because the shared structure is real, not because someone stipulated the analogy.

\ixnq{Hofstadter, Douglas R.}\ixnq{Sander, Emmanuel}\textcite{hofstadter2013} argue that analogy is the \enquote{fuel and fire of thinking}~-- that every act of categorization is an act of analogy, mapping new instances onto existing categories by structural similarity. The HPC framework specifies what that structural similarity consists of: shared mechanism-sustained properties. Hofstadter and Sander are right that analogy is pervasive. What they leave unspecified is what makes one analogy productive and another empty. The answer is maintenance: when the shared structure is maintained by overlapping mechanisms, the analogy generates true predictions. When it isn't, it generates false ones. The quality of an analogy tracks the density of shared causal architecture, not the vividness of the surface match.

The explanatory coherence runs deeper than self-application. The framework that explains what grammatical categories are also explains how theories about grammatical categories get built~-- through the same cross-domain extension it predicts. This isn't a post hoc observation bolted on for elegance. It's a structural consequence of the theory's content: if categories are maintained clusters that support inductive extension, and if theories are themselves maintained clusters of concepts and commitments, then the capacity for theory-building is a special case of the capacity for categorization. The circle closes because it was never really a circle. It's a single capacity~-- kind-formation under maintenance~-- operating at different scales.

\section{What this means}
\label{sec:16:what-this-means}

For cognitive linguistics, three consequences.

First, prototype effects are mechanistically grounded, not primitive. When Rosch found that robins are better birds than penguins, she identified a real gradient and proposed cue validity as its explanation~-- a genuine one. The HPC framework adds the maintenance story: the same processes~-- entrenchment, acquisition bias, functional pressure~-- keep regenerating the cue-validity structure in each speaker and across generations. Prototype theory quantified the gradient and explained its shape. Homeostasis explains why the shape persists.

Second, cross-domain mappings work because of shared mechanism-sustained structure. Image schemas\ixsq{image schema} may well be real psychological structures, and nothing here denies their existence. But they describe the \emph{shape} of the overlap without explaining its \emph{source}. The HPC account says what source and target have to share for the mapping to be productive: not just structural resemblance but maintained structural resemblance. A mapping between two HPC kinds with rich overlapping mechanisms will be productive, generating novel inferences that hold. A mapping where the overlap is accidental~-- shared features without shared maintenance~-- will be shallow, generating surface analogies that break under load. The methodological upshot: stop looking for the mapping and start looking for the shared maintenance. Which mechanisms does the source domain share with the target? Where do they diverge? That's where the metaphor breaks down, and that's where the most informative data lives.

Third, the framework is self-applying: it predicts the very capacity it uses. This adds a falsification condition that cognitive linguistics lacks. If cross-domain extension between HPC kinds doesn't track shared mechanism density~-- if rich metaphors arise between domains with no overlapping causal architecture, or if impoverished metaphors persist between domains with substantial shared maintenance~-- the framework loses explanatory scope. The prediction is on the table.

For the book's programme, the consequences are equally direct. The constructive argument (Parts II and III) showed that grammatical categories are maintained clusters, projectible relative to purposes. The gauntlet that follows demands predictions and defeat conditions. This section adds one more prediction: cross-domain extension should track mechanism density. And it provides one more piece of evidence: the theory's own history~-- its journey from biology to linguistics to a theory of metaphor~-- instantiates the capacity it describes. If that's a coincidence, it's a remarkably well-structured one.

That history also explains why the gauntlet that follows can demand the predictions it demands. The application of HPC from biological species to grammatical categories isn't an analogy that motivated the book and can now be discarded. It's a theory-constitutive metaphor in Boyd's sense: the shared causal architecture between species and grammatical categories~-- maintained clustering, projectible properties, patterned boundary behaviour, degradation when mechanisms weaken~-- is what licences the inferential transfer. The predictions tested below (asymmetric fraying, sticky-lock drift, phase transition in genesis) are generated by that transfer: they hold for biological species, the property overlap is mechanism-sustained, and so the framework predicts they should hold for grammatical categories too. If the overlap were merely surface resemblance~-- if species and grammatical categories shared features without sharing maintenance architecture~-- the predictions would be decorative, not load-bearing. They carry weight because the metaphor is constitutive.

The remaining sections take these predictions~-- alongside all the others this book has accumulated~-- and turn them into a gauntlet with observables, baselines, and defeat conditions.

\section{What the book commits to}
\label{sec:16:synthesis}

The constructive argument is behind us. Categories are property clusters maintained by mechanisms and projectible for a purpose; the case studies have put that claim under load across countability, definiteness, lexical categories, and gender. What remains is to say what the framework predicts, what would break it, and what questions it opens that weren't askable before. The rest of this chapter takes on all three. Chapter~\ref{ch:failure-modes}'s audit asked what a single proposed category owes; this chapter asks what the framework itself owes across domains. Figure~\ref{fig:16:gauntlet} maps the evaluation pathway, and English \mention{data}\ixlq{data} will serve as a running case throughout.

\begin{figure}[htbp]
\centering
\resizebox{0.85\textwidth}{!}{%
\begin{tikzpicture}[
    node distance=0.55cm and 1.8cm,
    entrybox/.style={
        rectangle, draw, rounded corners=3pt, thick,
        minimum width=4.2cm, minimum height=0.8cm,
        align=center, font=\small\bfseries
    },
    gatebox/.style={
        rectangle, draw, rounded corners=3pt, fill=black!5,
        minimum width=4.2cm, minimum height=0.8cm,
        align=center, font=\small
    },
    gauntletbox/.style={
        rectangle, draw, rounded corners=3pt, fill=black!10,
        minimum width=5.6cm, align=left,
        font=\small, text width=5.4cm, inner sep=8pt
    },
    exitbox/.style={
        rectangle, draw=black!50, dashed, rounded corners=2pt,
        minimum height=0.6cm,
        align=center, font=\footnotesize
    },
    verdictbox/.style={
        rectangle, draw, rounded corners=3pt, thick,
        minimum width=2.2cm, minimum height=0.7cm,
        align=center, font=\small\scshape
    },
    pass/.style={->, >={Stealth[length=3mm]}, thick},
    fail/.style={->, >={Stealth[length=2.5mm]}, thick, dashed, draw=black!50}
]

% === ENTRY ===
\node[entrybox] (entry) {Proposed category};

% === THREE DEBTS ===
\node[gatebox, below=0.9cm of entry] (d1) {1.\ Show clustering};
\node[gatebox, below=0.5cm of d1] (d2) {2.\ Identify stabilisers};
\node[gatebox, below=0.5cm of d2] (d3) {3.\ Demonstrate projection};

% === EXIT BRANCHES ===
\node[exitbox, right=1.8cm of d1] (e1) {Not a candidate};
\node[exitbox, right=1.8cm of d2] (e2) {Essentialist challenge};
\node[exitbox, right=1.8cm of d3] (e3) {Nominalist challenge};

% === GAUNTLET BOX ===
\node[gauntletbox, below=1cm of d3] (gauntlet) {%
    \textbf{Predictions}\\[2pt]
    \hspace{0.3cm}Asymmetric fraying?\\
    \hspace{0.3cm}Sticky-lock drift order?\\
    \hspace{0.3cm}Phase transition in genesis?\\[6pt]
    \textbf{Stress tests}\\[2pt]
    \hspace{0.3cm}Mechanism-light projection?\\
    \hspace{0.3cm}Baseline parity after controls?%
};

% === VERDICTS ===
\node[verdictbox, below left=1.2cm and 1.2cm of gauntlet] (confirm) {confirm};
\node[verdictbox, below=1.2cm of gauntlet] (narrow) {narrow};
\node[verdictbox, below right=1.2cm and 1.2cm of gauntlet] (downgrade) {downgrade};

% === MAIN SPINE ===
\draw[pass] (entry) -- (d1);
\draw[pass] (d1) -- (d2);
\draw[pass] (d2) -- (d3);
\draw[pass] (d3) -- (gauntlet);

% === FAIL EXITS ===
\draw[fail] (d1.east) -- (e1.west);
\draw[fail] (d2.east) -- (e2.west);
\draw[fail] (d3.east) -- (e3.west);

% === VERDICT ARROWS ===
\draw[pass] (gauntlet.south) -- ++(0,-0.35) -| (confirm.north);
\draw[pass] (gauntlet.south) -- (narrow.north);
\draw[pass] (gauntlet.south) -- ++(0,-0.35) -| (downgrade.north);

% === GROUP LABELS ===
\node[font=\footnotesize, text=black!60, rotate=90, anchor=south]
    at ([xshift=-0.6cm]d2.west) {Three debts};
\node[font=\footnotesize, text=black!60, rotate=90, anchor=south]
    at ([xshift=-0.6cm]gauntlet.west) {Gauntlet};

\end{tikzpicture}}%
\caption{The evaluation pathway. Every proposed category owes three debts (solid path); failure at any gate exits to the named challenge (dashed). Categories that clear all three enter the gauntlet: three predictions the framework bets on, and two stress tests that could break it. Cumulative evidence determines the verdict.}
\label{fig:16:gauntlet}
\end{figure}

At the quantum scale there are no cats, and at the scale of the isolated token there are no categories. A category\ixsq{category!as HPC kind} can be real at any grain~-- phonological, morphosyntactic, semantic, discourse, social~-- as long as it's maintained and projectible there. The framework doesn't privilege any level of linguistic organisation, and it doesn't exempt any level from scrutiny.

Every proposed category\ixsq{category!as HPC kind} owes three debts: show that properties cluster, identify the stabilisers that maintain the clustering, and demonstrate projection~-- what learning one case buys you for the next. The three debts form a chain. Clustering alone isn't enough, because accidental clusters don't persist; the clustering has to be maintained by identifiable stabilisers. And maintenance alone doesn't earn the label, because a maintained cluster that doesn't project~-- that doesn't let you predict anything about unexamined cases~-- hasn't earned its keep as a category.

One commitment runs through all the case studies: boundaries are ontologically sharp but epistemically graded. What varies within a category is stability and typicality, not degree of membership. What looks like gradience across categories is either measurement noise near a sharp boundary or dual citizenship in overlapping basins (Chapter~\ref{ch:dynamic-discreteness}). The variance signature is the test: if boundary items show no elevated variance relative to basin cores, the sharp-boundary architecture is wrong and Khalidi-style fuzzy realism deserves the field.

Table~\ref{tab:16:synthesis-ledger} collects the book's outstanding debts. Each row names what the maintained cluster should project, what that projection looks like in observables, and what would force a downgrade. These aren't chapter summaries; they're commitments with due dates. All five rows face the same inferential logic that Chapter~\ref{ch:social-stabilization} established (conditioned models against pooled baselines), operationalised differently for each domain.

\begin{table}[htbp]
\centering
\caption{Synthesis ledger: what each maintained cluster projects, and what would break it.}
\label{tab:16:synthesis-ledger}
\small
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.16\textwidth}>{\raggedright\arraybackslash}p{0.27\textwidth}>{\raggedright\arraybackslash}p{0.27\textwidth}>{\raggedright\arraybackslash}p{0.24\textwidth}@{}}
\toprule
Domain & What the cluster projects & Observable signature & Downgrade signal \\
\midrule
Countability (Ch.~\ref{ch:countability}) & Knowing a noun is count predicts its behaviour across tight and loose frames; knowing its position in the individuation hierarchy predicts which diagnostics hold. & Structured boundary behaviour: core diagnostics resist while edge diagnostics drift. & Near-uniform drift across diagnostics and speakers with no stable conditioned structure across repeated datasets. \\
\addlinespace
Definiteness/\newline deitality (Ch.~\ref{ch:definiteness-and-deitality}) & Two independent inductive profiles: semantic identifiability predicts discourse behaviour; formal deitality predicts distributional behaviour. & Stable decouplings (form without semantics; semantics without form). & Durable lockstep where decoupling is predicted. \\
\addlinespace
Lexical categories (Ch.~\ref{ch:lexical-categories}) & Nouns and verbs project robustly (learn the category, predict morphology and distribution); adjectives project thinly; \mention{adverb} doesn't project at all. & Robust noun/verb projection, thinner adjective regimes, wastebasket failures. & Equal projection across classes after matched diagnostics and conditioning controls. \\
\addlinespace
Pro-form gender (Ch.~\ref{ch:proform-gender}) & Knowing a referent's personhood construal predicts pro-form selection across the entire inventory and across reference chains. & Personhood-conditioned selection and chain-coherence effects with graded edges. & No reliable designatum conditioning once context is controlled. \\
\addlinespace
Grammaticality + social stabilisation (Chs.~\ref{ch:grammaticality-itself}, \ref{ch:social-stabilization}) & Mastering a form--meaning coupling lets you deploy it in novel sentences; knowing the social conditioning structure predicts where judgments will converge and where they'll split. & Detector misfires are patterned; conditioned models recover structure from pooled noise. & Conditioning yields no predictive gain over pooled baselines. \\
\bottomrule
\end{tabular}
\end{table}

Where those patterns appear, the framework's credibility\ixsq{projectibility!field-relative} rises. Where they don't, it should drop. Table~\ref{tab:16:gauntlet} handles the operational side~-- the specific tests that could confirm, narrow, or break the claims.

\section{What HPC opens}
\label{sec:16:what-hpc-opens}

When HPC thinking entered other fields, it didn't just relabel existing debates. It made new questions askable. The pattern is worth tracing, because linguistics is about to repeat it.

In biology, the species problem was stuck between essentialism (what is the essence of a species?) and nominalism (species are just convenient labels). HPC dissolved the impasse by asking a different question: what mechanisms maintain species cohesion? That question generated a research programme. Gene flow, shared selection pressures, and developmental constraints turned out to be the relevant mechanisms, and their relative strength explained why the species concept works better for some taxa than others. Speciation became a question about mechanism decoupling: one cluster becomes two when gene flow breaks. Hybrid zones turned from embarrassments into informative data, because the mechanisms weaken there in a structured gradient \citep{boyd1999}.

In psychiatry, the pattern is closer to home. The DSM's necessary-and-sufficient criteria for mental disorders produced categories that didn't carve nature at its joints, while dimensional models captured symptom severity but couldn't explain why symptoms cluster at all~-- exactly the essentialism-versus-prototype impasse that linguistics knows well. Mechanistic approaches dissolved this by asking what maintains the symptom cluster \citep{kendler2011}. Network theory gave a specific answer: symptoms causally sustain each other (insomnia produces fatigue, which produces social withdrawal, which deepens low mood, which worsens insomnia). That reframing opened questions the old framework couldn't even formulate. Which symptom is the \emph{hub}~-- the one whose removal most destabilises the rest? Why do disorders co-occur (is comorbidity a matter of shared mechanisms rather than coincidence)? Is depression categorical or dimensional (Meehl's taxometric methods test this directly rather than assuming an answer)? And is there a general factor of psychopathological tightness~-- some individuals clustering more tightly across the board?

The parallel with linguistics is almost exact, and it's not a coincidence: the same intellectual impasse invites the same dissolution. We've been stuck between essentialist definitions (a noun is a word that names a person, place, or thing) and nominalist scepticism (categories are just convenient labels for gradient distributions). The case studies dissolve the impasse the same way Boyd dissolved the species problem. But we haven't yet imported the questions that the other fields started asking once the impasse was behind them.

The hub-node question is the most conspicuous gap. This book describes coupling between properties (Chapter~\ref{ch:the-category-zipper}) but doesn't ask which property in any given cluster is load-bearing~-- the one whose removal would most destabilise the rest. For countability: is it individuation, plural morphology, or determiner selection? The psychiatry analogue suggests the answer matters for both theory and intervention. Intervene on the hub and the cluster unravels; intervene on a peripheral node and it holds. The case studies have the data to begin answering this question but haven't asked it.

The comorbidity question is equally unasked. Why do certain categories co-occur in grammars? Languages with robust count/mass systems tend to have articles. If countability and definiteness share a maintenance mechanism (referential anchoring, say), their typological co-occurrence isn't coincidental~-- it's category-level comorbidity. Perturb the shared mechanism and both categories should weaken. The prediction is testable against typological databases, and it's a question that neither prototype theory nor essentialism can formulate.

The taxometric question may be the sharpest methodological import. Psychiatry developed specific statistical methods (MAXCOV, MAMBAC, L-Mode) to detect whether a latent variable is categorical or continuous, and has applied them to dozens of diagnostic constructs. These methods have never been applied to linguistic categories. Running MAXCOV on countability judgments across a large speaker pool would be a direct test of the sharp-boundary commitment. If a latent taxon appears, the sharp-boundary claim gets empirical support it currently lacks. If the distribution is continuous, Khalidi's fuzzy kinds deserve the field. The tools exist; nobody has used them.

Finally, the constructive question. The predictions in Section~\ref{sec:16:prediction-fraying} are all about degradation~-- what happens when maintenance weakens. The book has said little about what happens when new maintenance emerges, and that asymmetry matters, because category birth is as theoretically important as category death. The definiteness/deitality case from Chapter~\ref{ch:definiteness-and-deitality} provides the template, and the process isn't fission but budding. Definiteness existed as a semantic cluster; grammaticalization of the demonstrative into an article created a \emph{new} mechanism (the article system) that built a second cluster with its own maintenance regime. The formal system developed independent stabilisers and can now decouple from its parent~-- not a broken connection but a new one. That reframing opens specific questions for grammaticalization theory that Section~\ref{sec:16:research-agenda} picks up.

\section{The gauntlet}
\label{sec:16:prediction-fraying}

The three debts established that categories have to project. The gauntlet asks whether the maintenance account correctly predicts \emph{how} they project~-- where projection holds, where it thins, and what degradation looks like when stabilisers weaken. Projection is the point; maintenance is the claimed explanation. If the explanation is right, maintained clusters should project in ways that generate expectations specific enough to fail. Three predictions and two stress tests follow.

The first prediction concerns what happens when a stabiliser weakens. Take \mention{data}\ixlq{data} from Chapter~\ref{ch:countability}. As \mention{datum}\ixlq{datum} recedes from use, the framework predicts that diagnostics won't erode uniformly: tight frames should go first, loose frames should hold longer, and agreement should lag behind both. The variance should partition between speakers more than within them, with individual commitments staying sharp even while the population distribution looks graded.

The principle generalises: whenever stabilisers uncouple, decay should be patterned, keyed to which mechanism\ixsq{mechanism!perturbation of} weakened first rather than spreading uniformly across diagnostics. The case studies are consistent with this, but consistency isn't risk. The risky version of the prediction is that categories undergoing change in other languages~-- ones not examined in this book~-- should show the same asymmetric fraying. If they don't, the account has overfit to its training data. And even before fraying begins, the pattern of absences at the boundary~-- which properties a borderline case lacks~-- should map the mechanism landscape (Chapter~\ref{ch:failure-modes}). Random absences mean accidental clustering; structured absences mean real mechanisms.

The second prediction concerns not which diagnostics fray but the order in which they do. Usage frequency and semantic construal respond directly to communicative pressure and shift first. Morphosyntactic frames resist longer because they're anchored by entrenchment and by the distributional expectations that other constructions impose on them. When those frames do give way, they tend to fail in bursts rather than eroding smoothly~-- the \term{sticky-lock effect}\ixsq{sticky-lock effect}. The frame persists after the pressure profile that built it has already shifted.

The prediction divides by scope. Language-particular predictions target concrete diagnostics inside one grammar. Comparative predictions target recurring shapes of drift across unrelated languages under similar pressures, expecting convergence on similar local fixes even when the material details differ. For \mention{data}\ixlq{data}: staged erosion (tight before loose) confirms the sticky lag; lockstep refutes it. In cohorts where \mention{datum}\ixlq{datum} is absent from child-directed input, loose packaging should stabilise before tight singulative\ixsq{singulative} packaging. Age-graded corpora are the natural test bed.

The third prediction is constructive rather than degradative. If the budding account from Section~\ref{sec:16:what-hpc-opens} is right, grammaticalization should show a phase transition rather than the continuous clines that standard grammaticalization theory assumes \citep{hopper1993}. There should be a point where an emerging cluster acquires enough independent maintenance to project on its own. Before that point, knowing a form is \mention{the} predicts only \enquote{demonstrative for identifiable referents.} After that point, knowing it's an article predicts distributional behaviour, obligatoriness, phonological reduction, and complementary distribution with other determinatives.

The prediction is testable: plot the correlation between emergent properties over historical time. The maintenance account predicts a nonlinear increase~-- a phase transition. Cline theory predicts a linear one. This is a genuine empirical clash, and diachronic corpus data can adjudicate it.

Those are the predictions. Two stress tests could break the framework decisively.

The first is \term{projection without maintenance}\ixsq{projection without maintenance}: a robust, cross-linguistically stable category\ixsq{category!cross-linguistic} that projects strongly while showing no identifiable causal structure holding it together~-- no acquisition bias, no alignment\ixsq{alignment!convergence effects} pressure, no functional anchoring\ixsq{functional anchoring}, no institutional reinforcement, no attractor-like distributional profile. If that pattern replicates across datasets and preregistered controls, the maintenance story is unnecessary.

The second is \term{structure without lift}\ixsq{structure without lift}: usage variation that turns out to be a random walk once sampling noise is controlled, with no reliable conditioning by frequency, no alignment\ixsq{alignment!convergence effects} effects, no stable register or community attractors, no basin\ixsq{basin}-like recovery after perturbation. If categories don't predict beyond convenience labels on out-of-sample data, the account needs scope reduction.

To limit post hoc rescue, preregister the proxy family for \enquote{maintenance structure} and report how conclusions change under alternative specifications. The evaluation should be cumulative: if major domains repeatedly show no mechanism\ixsq{mechanism!perturbation of}-linked lift, downgrade strongly; mixed results call for scope narrowing; robust, replicated lift raises confidence.

Preregistration guards against one threat~-- the garden of forking paths, where researcher degrees of freedom inflate apparent support \citep{GelmanLoken2013}. But a subtler threat deserves attention: treating the gauntlet as hypothesis testing when its logic is really model criticism. A hypothesis test asks a binary question (did asymmetric fraying appear?). Model criticism asks a distributional one: if the maintenance story is right, what should the data look like~-- what variance structure, what fraying shape, what lag signature~-- and does the real data look like that \citep[ch.~6]{GelmanBDA3}? A framework can pass individual predictions while generating distributions that look nothing like the observed data. It can also fail a specific prediction while getting the overall shape right, and the latter is more informative than a clean pass on a weak test.

Under model criticism, the gauntlet produces a failure map rather than a verdict: a record of where the maintenance model's predictions match the data and where they diverge, keyed to specific mechanisms. A failure concentrated in one domain (definiteness decoupling shows no predicted lag, say, while countability and gender hold) tells you which mechanism is carrying less weight than claimed. A diffuse failure~-- no domain shows the predicted variance structure, convenience labels predict as well as mechanistic models across the board~-- is catastrophic. The pattern of failure carries more information than any single test.

The framework makes a further prediction that's easy to miss: the statistical model itself should be hierarchical. Categories are maintained at multiple levels (individual entrenchment, community alignment, cross-linguistic functional pressure), and those levels should show up as variance components. Speaker-level effects should dominate item-level effects for boundary cases; community-level conditioning should recover structure that pooled analyses miss. This is what Eckert-style variationist work already demonstrates ethnographically (Chapter~\ref{ch:social-stabilization}): condition on the right social structure and apparent noise resolves into tight coupling. If a flat model with no hierarchical structure fits the data just as well, the multi-timescale maintenance story is overclaiming.

One mechanism\ixsq{mechanism!perturbation of} claim can fail in one domain while others carry weight elsewhere; that's correction, not collapse. But if repeated studies show weak support for patterned fraying, no reliable lag structure, and no stable lift over convenience labels across major domains, the framework should be treated as provisional heuristic vocabulary rather than settled explanatory ontology.

\section{Two cases under load}
\label{sec:16:operationalizing}

To keep the gauntlet empirical, each claim needs observables and predeclared comparison criteria.

\begin{table}[htbp]
\centering
\caption{Gauntlet claims, observables, and predeclared stress-test criteria.}
\label{tab:16:gauntlet}
\small
\begin{tabular}{@{}p{0.20\textwidth}p{0.36\textwidth}p{0.36\textwidth}@{}}
\toprule
Claim & Observable signature & Outcome that strongly downgrades maintenance \\
\midrule
Asymmetric fraying & Boundary diagnostics drift in structured, non-synchronous ways; variance partitions by speaker/context more than pure item noise. & Diagnostics move as unstructured free noise; no stable variance structure across replications. \\
\addlinespace
Sticky-lock drift & Diachronic lag between usage/semantic shift and morphosyntactic lock shift; local burst dynamics near boundary transitions. & No lag and no burst structure: all dimensions move in lockstep or diffuse linearly without staged decoupling. \\
\addlinespace
Phase transition in genesis & Emergent properties of a grammaticalizing form show nonlinear correlation increase; projection appears abruptly once cluster becomes self-maintaining. & Property correlations increase linearly or show no inflection; grammaticalization proceeds as a featureless cline. \\
\addlinespace
Projection without maintenance & Category projects strongly under perturbation despite no identifiable maintenance architecture. & Persistent mechanism-light projection remains after preregistered proxy controls and robustness checks. \\
\addlinespace
Structure without lift & Baseline random-walk or convenience-label models match mechanistic models on out-of-sample prediction and perturbation recovery. & Mechanistic models show no consistent out-of-sample advantage once uncertainty and calibration are accounted for across datasets. \\
\bottomrule
\end{tabular}
\end{table}

For \mention{data}\ixlq{data}, the variables are agreement choice, tight-frame acceptance (\mention{a datum}\ixlq{datum}, low cardinals), and loose-frame acceptance (\mention{many}\ixlq{many} vs \mention{much}\ixlq{much}), conditioned by register, community, cohort, and constructional context. The statistical details belong in a preregistration, not a book chapter. And as Chapters~\ref{ch:projectibility} and \ref{ch:countability} stress, didactic ABMs\ixsq{agent-based model} are mechanism\ixsq{mechanism!defined} sketches, not alibis; status judgments rest on replicated empirical datasets, not on one toy trajectory.

\subsection{A second case: pro-form gender\ixsq{gender!pro-form}}
\label{sec:16:gender-gauntlet}

The \mention{data}\ixlq{data} case tests what happens to projection when a stabiliser weakens. Pro-form gender\ixsq{gender!pro-form} tests the opposite: what sustains projection in a tight system. Together, the two cases put the gauntlet under load in both directions.

A family's dog has died. \enquote{She was a good dog,} says the owner. \enquote{It was old,} says the vet. The referent is the same, but the construals differ and so do the pro-forms. The choice tracks the speaker's construal of the designatum along the personhood\ixsq{personhood hierarchy} hierarchy that Chapter~\ref{ch:proform-gender} established: personal or non-personal, and if personal, epicene or sex-specific.

That hierarchy is maintained by three converging stabilisers. \term{Cognitive grounding}\ixsq{cognitive grounding}: the person/non-person distinction maps onto a pre-linguistic conceptual split tied to Theory-of-Mind\ixsq{Theory of Mind} development. Children don't learn to distinguish persons from objects through grammar; grammar rides a distinction already in place. \term{Semantic transparency}\ixsq{semantic transparency}: the mapping from personhood to pro-form is largely predictable, unlike French or German noun-class gender. High learnability means low acquisition error, which means the system reconstructs itself cheaply in every generation. \term{Alignment and repair}\ixsq{alignment!in gender}: misgendering triggers real-time correction. When a speaker calls a person \mention{it}\ixlq{it}, the violation is immediate, social, and repaired on the spot. This triad~-- cognitive scaffold, transparent mapping, conversational enforcement~-- keeps English pro-form gender\ixsq{gender!pro-form} tightly coupled despite having no morphological agreement to lean on.

Each stabiliser generates a testable prediction. Cognitive grounding predicts that children's pro-form assignment for novel entities (robots, AI assistants, animated characters) should track Theory-of-Mind\ixsq{Theory of Mind} attribution rather than animacy per se: a child who attributes intentions to a robot should produce \mention{he}\ixlq{he} or \mention{she}\ixlq{she}, while one who treats the same robot as a tool should produce \mention{it}\ixlq{it}. Semantic transparency predicts that when a novel entity enters a speech community, pro-form assignment should converge quickly once the community's personhood construal settles, with no extended period of gender-agreement errors comparable to those in opaque systems. Alignment and repair predict that gender corrections should be faster and more socially clustered than other grammatical repairs.

The chain-coherence constraint adds a system-level prediction. Chapter~\ref{ch:proform-gender} showed that once a referent is construed as personal within a reference chain, switching to a non-personal pro-form is disfavoured. If this constraint is real, mixed-gender chains should incur measurable processing costs: longer reading times, higher error rates in recall, and more frequent spontaneous repairs. If mixed chains process as smoothly as consistent ones under controlled conditions, the coupling claim needs narrowing.

What would falsify the account? Three findings, any one of which would force serious revision. If pro-form selection shows no reliable sensitivity to personhood construal once syntactic context is controlled, the cluster doesn't project along the dimension claimed. If gender mismatches are repaired no more readily than other agreement errors, the social-enforcement stabiliser isn't doing the work claimed for it. And if novel-entity assignment shows no convergence pattern but instead looks like noise, the transparency mechanism isn't generating the inductive reliability the account requires.

\section{The research agenda}
\label{sec:16:research-agenda}

Throughout this book I've made predictions I haven't tested. Individually, each promissory note is defensible~-- a theory book earns its keep by generating testable claims, not by running every experiment. But the notes accumulate, and the reader deserves an honest ledger.

\textsc{Countability} (Ch.~\ref{ch:countability}). The two-cluster architecture predicts that classifier languages should show the same tight/loose asymmetry as English: diagnostics closest to individuation should resist change while packaging diagnostics should drift. A systematic comparison of Mandarin, Thai, and Japanese classifier systems against the implicational hierarchy would be a direct test.

\textsc{Definiteness} (Ch.~\ref{ch:definiteness-and-deitality}). Two predictions need experimental confirmation. Prosodic rescue should selectively affect \mention{there}-resistance (a form-cluster property) but not partitive licensing (a deeper distributional fact)~-- a dissociation that would confirm the two-cluster architecture independently of the acquisition data. And children should master the distributional restrictions (the form cluster) before the semantic ones (identifiability, uniqueness), though critically, this ordering needs to be shown as a prediction of the architecture rather than merely consistent with it.

\textsc{Lexical categories} (Ch.~\ref{ch:lexical-categories}). The adjective-as-plumage diagnosis has an acknowledged falsifier: a language with a small, closed adjective class that nevertheless has its own non-borrowed morphological paradigm would be skeletal, not plumage. Whether such a case exists is an empirical question the typological literature should be able to answer. Separately, the claim that lexical category is itself an HPC~-- a category of categories~-- is sketched but not tested.

\textsc{Pro-form gender} (Ch.~\ref{ch:proform-gender}). The three maintenance mechanisms generate predictions that remain promissory: children should assign pro-forms to robots based on Theory-of-Mind tests, not animacy alone; novel entities should be rapidly assigned to the personhood hierarchy; gender repairs should be faster than number repairs. The chain-coherence effects, which carry substantial theoretical weight, rest on constructed examples and introspective judgments rather than corpus or experimental evidence.

\textsc{The category zipper} (Ch.~\ref{ch:the-category-zipper}). The discourse level appears in the stabiliser-weighting table but received no evidence. If the framework claims scale-invariance, skipping discourse is conspicuous. Even a negative case~-- showing why discourse categories resist the two-diagnostic test~-- would strengthen the argument.

\textsc{Grammaticality} (Ch.~\ref{ch:grammaticality-itself}). The three-register phenomenology is testable: if the qualitative, reactive, and normative readouts collapse to a single factor under controlled measurement, the tripartite structure is mnemonic rather than mechanistic. Schema drift (Vagnino and Walker) predicts specific satiation asymmetries that can be tested with current methods.

\textsc{Social stabilisation} (Ch.~\ref{ch:social-stabilization}). The mixed-bin diagnostic predicts that conditioning should recover tight structure from apparently loose distributions. The methodological commitment~-- no conditioning gain, no maintenance claim~-- needs worked demonstrations beyond the double copula and existential \mention{there's}. Register-based predictions (academic writing, clinical discourse) are the most accessible testing grounds.

Those are the chapter-specific debts. The diff-in-diff analysis in Section~\ref{sec:16:what-hpc-opens} adds five more that cross chapter boundaries.

\textsc{Hub-node empirics}. For each case study, identify which property in the cluster is load-bearing. The method is perturbation: remove or degrade one property and measure how much the rest destabilises. If individuation is the hub for countability, degrading individuation should unravel more of the cluster than degrading, say, plural morphology. If every property contributes equally, the hub-node model is wrong and the cluster is better described as a flat network.

\textsc{Category comorbidity}. Test whether categories that share maintenance mechanisms co-occur typologically more than chance predicts. Countability and definiteness are the first candidates: if they share a referential-anchoring mechanism, they should co-occur across languages more than a feature-independence baseline would predict. Grambank and WALS provide the data; the prediction is testable now.

\textsc{Taxometrics}. Apply Meehl's taxometric methods (MAXCOV, MAMBAC, L-Mode) to linguistic category judgments. Run them on countability judgments across a large speaker pool. If a latent taxon appears, the sharp-boundary commitment gets direct support. If the distribution is continuous, Khalidi's fuzzy kinds win. This would be the first application of taxometric methods to linguistic categories.

\textsc{Competitive exclusion}. When two forms grammaticalize into the same functional niche (\mention{will}/\mention{going to}; \mention{must}/\mention{have to}), can they coexist as distinct categories? The framework predicts the answer depends on whether they recruit distinct mechanisms. Same mechanisms should produce competitive exclusion; different mechanisms should produce differentiation into distinct categories with different projection profiles. \mention{Will} and \mention{gonna} differentiated. The question is whether the framework can predict which competitions result in differentiation and which in death, based on mechanism overlap, before the outcome is known. Grammaticalization theory can describe these outcomes but can't predict them from mechanism landscape.

\textsc{Developmental cluster tightening}. Children's diagnostics should become more \emph{correlated} over time, not just more accurate. P(uses \mention{a(n)} $|$ uses plural) should increase from age 2 to 4. If diagnostics improve independently rather than tightening as a cluster, the coupling mechanism isn't operating in acquisition.

That's twelve open fronts. Some will confirm the framework, others will force scope narrowing, and a few may require substantive revision. A framework that can't point to its own outstanding debts has stopped being empirical.

To see what this changes in practice, consider a concrete case. A syntactician studies \mention{otherwise}\ixlq{otherwise}~-- the word that opened Chapter~\ref{ch:words-wont-hold-still}. The old question: is it an adverb or a conjunction? The HPC question: what maintains its distributional profile, and where does maintenance thin out? Instead of eliciting categorical judgments on constructed sentences, the researcher tracks which syntactic frames resist perturbation across registers, which vary between speakers, and which have changed diachronically. The gauntlet adds a further demand: preregister the variance structure you expect, name the outcome that would force you to treat \mention{otherwise}\ixlq{otherwise} as a convenience label rather than a maintained category, and report what you find either way.

\section{Conclusion}
\label{sec:16:zipper-at-scale}

\textcite{ladyman2007}\ixnq{Ladyman, James}\ixnq{Ross, Don} call this \term{scale-relative ontology}\ixsq{scale-relative ontology}: what counts as a real pattern depends on the grain at which you measure. Their target is physics, but the principle transfers directly. A phoneme\ixsq{phoneme!as HPC kind} is real at the grain of the syllable; a noun\ixsq{noun!as HPC kind} is real at the grain where distributional, morphosyntactic, and semantic properties cluster and project; a register is real at the grain where conditioning recovers predictability from apparent noise. The \term{meta-Occam\ixsq{meta-Occam}} principle from Chapter~\ref{ch:kinds-without-essences} follows: be parsimonious about mechanisms, not about labels. If a traditional label names a maintained profile, keep it; if it doesn't, replace it without guilt.

The comparative consequence is direct. Comparative concepts can stay useful without guaranteeing a language-particular category\ixsq{category!cross-linguistic} in every grammar. Linguists can compare adjective\ixsq{adjective}-like behaviour across languages even where no dedicated adjective\ixsq{adjective} category\ixsq{category!cross-linguistic} is maintained inside a given language~-- that's a null case rather than a framework failure. The absence of a category at one grain, in one language, doesn't threaten the reality of that category where it is maintained elsewhere.

What changes in practice is the question, not the answer. Fieldwork stops asking \enquote{is X a Y?} and starts asking \enquote{what maintains X as a Y, and where does that maintenance thin out?}\ixnq{Huddleston, Rodney} Boundary cases become the most informative data in the sample rather than embarrassments to explain away. Study design shifts from hunting for clean centres to building gauntlets~-- predictions that can fail, stress tests that can bite, baselines that can win.

The framework is young and the evidence is partial. Some of the predictions in this chapter will fail. That isn't a caveat bolted on for modesty~-- it's the whole point. Categories live at the scale where maintained clusters project, where acquisition, entrenchment, alignment\ixsq{alignment!convergence effects}, and gatekeeping converge on a profile that supports induction\ixsq{projectable}. The book's claim isn't that categories are fictions sustained by habit. It's that the sustaining is the reality. Remove the maintenance and you don't reveal the category\ixsq{category!reality of} underneath; you dissolve it.

If the maintenance story keeps showing predictive lift under this gauntlet, confidence in its realism rises. Where it doesn't, narrow the scope and revise the mechanisms. At the scales where categories project, they're as real as anything else we measure.
