\chapter{Grammaticality itself}
\label{ch:grammaticality-itself}
\epigraph{it's impossible to get rid of the illusion in spite of our better knowledge.}{— Hermann von Helmholtz, \textit{Treatise on Physiological Optics} (trans. J. P. C. Southall).}

Most analyses of constructional data presuppose a distinction between grammatical and ungrammatical\ixsq{ungrammatical}. The asterisk's the most common symbol in syntax, and one of the least likely to come with operating instructions. But what does the asterisk diagnose?

Chapter~\ref{ch:what-we-havent-been-asking} asked a question almost no one asks: is grammaticality\ixsq{grammaticality} itself a natural kind\ixsq{kind}? Not whether this sentence's grammatical, but what kind\ixsq{kind} of thing we're probing when we ask. This chapter provides the answer.

If categories are homeostatic property clusters, then grammaticality isn't merely a binary switch. It's itself an HPC\ixsq{HPC}~-- a maintained coupling between form and value\ixsq{value}. And the feeling of ungrammaticality\ixsq{feeling of ungrammaticality} that guides our judgments isn't an oracle; it's a noisy detector\ixsq{detector}, shaped by entrenchment and processing ease, capable of illusions.

Chapter~\ref{ch:dynamic-discreteness} introduced a two-layer picture: discrete grammaticality\ixsq{grammaticality} filtered through processing noise to yield gradient acceptability\ixsq{acceptability}. What we're adding here's the question of what \term{kind} of thing grammaticality\ixsq{grammaticality} is, and what maintains it.


\section{The HPC claim}
\label{sec:14:hpc-claim}

Grammaticality\ixsq{grammaticality}'s a form--value\ixsq{value} pairing. But the \enquote{value\ixsq{value}} side isn't compositional semantics~-- not truth conditions, not the full propositional content. It's something more basic: \term{structural meaning\ixsq{structural meaning}}.

Structural meaning\ixsq{structural meaning}'s what grammar\ixsq{grammar} contributes before lexical content fills the slots. It includes dependency relations, like the bracketing that distinguishes [\mention{the dogs'}] \mention{house} (a house belonging to the dogs) from \mention{the} [\mention{dogs' house}] (a type of house). It includes argument structure, distinguishing \mention{the dog bit the man} from \mention{the man bit the dog}. 

It provides temporal anchoring through tense and aspect, distinguishing whether the biting's past, ongoing, or hypothetical. It sets the clause type, distinguishing statements (\mention{he was bit}) from questions (\mention{was he bit?}) or commands (\mention{bite him.}). And it manages information structure\ixsq{information structure}~-- the difference between \mention{Him, he was bit} and \mention{it was him who was bit}.

This is \term{subcompositional\ixsq{subcompositional}}~-- it's the instruction manual, not the assembled product. Grammar\ixsq{grammar} tells you how to wire the meanings; the lexicon provides what to wire. \mention{Colorless green ideas sleep furiously} has perfect wiring instructions and absurd components. That's why it feels strange but not \term{ungrammatical\ixsq{ungrammatical}}.

A note on what structural meaning\ixsq{structural meaning} isn't. It isn't semantic coherence (\mention{Colorless green ideas} shows grammaticality\ixsq{grammaticality} survives semantic absurdity). It isn't contextual felicity (a sentence can be grammatical and pragmatically infelicitous). It isn't whatever the hearer infers (inference runs on many channels; grammaticality\ixsq{grammaticality}'s specifically the morphosyntactic one). Structural meaning\ixsq{structural meaning}'s the contribution of form to interpretation~-- the \term{wiring diagram\ixsq{wiring diagram}}, not the current flowing through it.

The HPC\ixsq{HPC} claim, then, is that grammaticality\ixsq{grammaticality}'s the coupling between morphosyntactic form and this structural semantics. When the coupling holds~-- when the form reliably signals the structural meaning\ixsq{structural meaning}~-- the sentence sits in the \term{grammatical basin\ixsq{grammatical basin}}. When it doesn't, the sentence's outside.

This is the \term{zipper\ixsq{zipper}}: morphosyntax provides one set of teeth, structural meaning\ixsq{structural meaning} the other. Grammaticality\ixsq{grammaticality}'s what it feels like when the zipper\ixsq{zipper} closes. An ungrammatical\ixsq{ungrammatical} sentence's a zipper\ixsq{zipper} that won't mesh.

Two metaphors will recur. The zipper\ixsq{zipper} describes the relation~-- the form--value\ixsq{value} alignment\ixsq{alignment} itself. The \term{immune system\ixsq{immune system}} (below) describes the \term{maintenance mechanism\ixsq{mechanism}}~-- what pushes the system back toward stable couplings when they slip. One's structure; the other's dynamics.


\section{The stabilizer\ixsq{stabilizer}: a feeling}
\label{sec:14:stabilizer}

What maintains this coupling? Partly the mechanisms we've discussed throughout: acquisition, entrenchment, alignment, transmission. But the proximate \term{stabilizer\ixsq{stabilizer}}~-- the thing that corrects deviations in real time~-- is a \term{feeling of ungrammaticality\ixsq{feeling of ungrammaticality}}.

When something's wrong with the form--value\ixsq{value} coupling, speakers notice. They hesitate, repair, rephrase. The feeling's what triggers the correction. It's the immune response of the grammar\ixsq{grammar}. Like immune systems, it can develop allergies: harmless processing hiccups get treated as pathogens.

The feeling's a \term{detector\ixsq{detector}}~-- and like all detectors, it's noisy. It's not a direct readout of the underlying structure. Instead, it's informed by probability and processing ease. First, the detector\ixsq{detector}'s Bayesian: a novel sentence's weighted by its likelihood under the distribution of attested forms~-- not just distance from the nearest exemplar, but probability given the entire pattern. Low-probability forms feel suspect even when grammatically licensed. The distribution itself is built by entrenchment\ixsq{entrenchment}: exposure tunes the prior. 

Second, sentences that parse smoothly feel grammatical; sentences that induce garden-paths or reanalysis feel suspect. Processing difficulty isn't the same as ungrammaticality, but the detector\ixsq{detector} often conflates them. This is why \term{acceptability\ixsq{acceptability}} and \term{grammaticality\ixsq{grammaticality}} come apart. Acceptability\ixsq{acceptability}'s what the detector\ixsq{detector} reports. Grammaticality\ixsq{grammaticality}'s the category\ixsq{category}~-- the HPC\ixsq{HPC} whose core's the form--value\ixsq{value} coupling. The two usually align, but they don't always.

Coupling shapes processing; processing shapes detector\ixsq{detector} output; detector\ixsq{detector} output crosses threshold; threshold triggers repair; repair reshapes coupling. To keep this concrete, we can write the loop with five moving parts:
\begin{quote}
\textbf{Mechanism\ixsq{mechanism} sketch (detector\ixsq{detector} and maintenance loop).}
\begin{enumerate}
    \item \textbf{Coupling state}: how strongly a construction\ixsq{construction} currently links morphosyntactic form to structural meaning\ixsq{structural meaning} for a speaker in a community.
    \item \textbf{Processing channel}: token-time load~-- surprisal, memory pressure, reanalysis cost.
    \item \textbf{Detector\ixsq{detector} output}: a graded mismatch signal combining coupling state and processing load, plus noise.
    \item \textbf{Decision threshold}: above some level, speakers report \enquote{ungrammatical\ixsq{ungrammatical}} and/or initiate repair.
    \item \textbf{Entrenchment\ixsq{entrenchment} update}: exposure and correction histories move both coupling strength and threshold over time.
\end{enumerate}
This keeps the levels separate: token-time parsing (\textasciitilde100 ms--seconds), individual entrenchment (days--years), and community norm dynamics (years--generations).
\end{quote}

This sketch commits us to an identification strategy. If processing load rises while coupling is stable, we should see transient detector\ixsq{detector} spikes without durable shifts in repair norms. If coupling weakens at matched load, we should see durable shifts in repair preferences and cohort trajectories. If thresholds move while coupling remains stable, ratings should move before production and repair distributions do. So ratings alone can't settle the ontology: the evidential standard is convergence across online processing, repair behavior, and community-distributional data (Chapters~\ref{ch:failure-modes} and~\ref{ch:what-changes}).

\subsection{Entrenchment\ixsq{entrenchment} in action}

But here's the puzzle: if grammaticality\ixsq{grammaticality} \emph{is} the maintained coupling, and entrenchment\ixsq{entrenchment}'s what maintains it, then \enquote{tuning the detector\ixsq{detector}} and \enquote{shaping what counts as grammatical} can't be fully separated over historical time. At a single time-slice, they \emph{are} separable~-- detector\ixsq{detector} output can misfire against a currently entrenched state, as illusions show. Across time, though, detector\ixsq{detector}-guided repair and uptake reshape the coupling itself. The coupling tunes the detector\ixsq{detector}; the detector\ixsq{detector}, over time, tunes the coupling. That's a reflexive equilibrium dynamic, not a vicious circle.

This is language-specific, construction-specific, even register-specific.

Consider age expressions: in French and Spanish, age is expressed with \mention{avoir}/\mention{tener}: \mention{J'ai vingt ans} (\enquote{I have twenty years}). In English, *\mention{I have twenty years}'s ungrammatical\ixsq{ungrammatical} as a statement of age~-- though perfectly fine as a statement of time (until retirement, on the job, etc.). The structural slot's the same; the construction\ixsq{construction}-specific entrenchment\ixsq{entrenchment} differs. 

The same applies to progressive aspect. If you're describing an ongoing event while your fingers are moving across a keyboard, \mention{I write} mis-signals aspect\ixsq{aspect} in English~-- many speakers experience this as \enquote{wrong} in the specific sense of a form--value\ixsq{value} mismatch. In mainstream present-day English, the progressive (\mention{I'm writing}) is often required for this ongoing-action construal. In French, the simple present would be grammatical for the same construal; the progressive's marked or absent. What counts as a violation of the form--value\ixsq{value} coupling depends on what the community has entrenched.

Countability\ixsq{countability} follows the same pattern. Whether \mention{furniture} takes count morphosyntax depends on entrenchment\ixsq{entrenchment}, not metaphysics. Furniture's as individuable as anything~-- you can count chairs and tables. But English has entrenched \mention{furniture} as noncount, so *\mention{I bought two furnitures}'s ungrammatical\ixsq{ungrammatical}. Other languages make different choices. The coupling's real; its contents are conventional.

And consider deitality\ixsq{deitality}: what requires the definite article in English~-- \mention{go to the hospital} vs.~\mention{go to hospital} (British)~-- is shaped by entrenchment\ixsq{entrenchment} within each dialect\ixsq{dialect}. The structural meaning\ixsq{structural meaning} (institutionalized activity frame)'s the same; the morphosyntactic realization differs.

These examples show grammaticality\ixsq{grammaticality} isn't a window onto universal logic. It's a window onto what a community has entrenched as the couplings between form and structural meaning\ixsq{structural meaning}. The feeling of ungrammaticality\ixsq{feeling of ungrammaticality} is tracking those couplings~-- but the couplings are historical, contingent, and maintained.

And here's where the entrenchment\ixsq{entrenchment} story connects to the social: entrenchment\ixsq{entrenchment}'s always entrenchment\ixsq{entrenchment} \enquote{within a discourse community\ixsq{discourse community}}. The grammar\ixsq{grammar} isn't floating free; it's tethered to the people who use it. When the community shifts~-- through contact, migration, register change~-- the entrenchment\ixsq{entrenchment} shifts with it. What feels ungrammatical\ixsq{ungrammatical} in one community may be unremarkable in another.

But speakers flow between communities~-- sometimes instantaneously~-- and communities can be created on the fly. \mention{Do you lift?} licenses intransitive \mention{lift} for weightlifters but not for warehouse workers; the grammaticality\ixsq{grammaticality} of the construction depends on which community you're invoking, and you can switch in mid-conversation. Code-switching within a clause~-- \mention{Vamos a hacer shopping}~-- is governed grammar\ixsq{grammar}, not error, but only within the discourse community\ixsq{discourse community} of the bilinguals who maintain it. The detector\ixsq{detector} isn't calibrated once; it's calibrated to whoever you're talking to, right now.

One more piece. We couldn't correct grammatical errors if we couldn't derive the intended meaning without the grammar. When someone says \ungram\mention{I have twenty years} for their age, we understand what they mean~-- and then we notice the mismatch. The feeling of ungrammaticality\ixsq{feeling of ungrammaticality} arises from comparison: the grammar's value\ixsq{value} versus the intended value\ixsq{value}. We need both to detect the error. This means comprehension's partly independent of grammaticality\ixsq{grammaticality}~-- and grammaticality\ixsq{grammaticality}'s partly a matter of checking the grammar's signal against other inference channels. When they align, we don't notice; when they mismatch, we do.



\section{The phenomenology\ixsq{phenomenology} of coupling: What it's like}
\label{sec:14:phenomenology}

If grammaticality\ixsq{grammaticality} is a maintained coupling, what is the \term{phenomenology\ixsq{phenomenology}} of that state? What is it \emph{like} to be in a state of grammaticality\ixsq{grammaticality}?

This section keeps a narrow claim. The phenomenology\ixsq{phenomenology} is a readout of the mechanism\ixsq{mechanism}, not evidence independent of it. The architecture was argued earlier (form, value\ixsq{value}, maintenance); what matters here is whether first-person reports line up with that architecture. They often do, but only at a coarse level. The analogies to balance, perception, and fluency are useful mnemonics, not explanatory load-bearing beams.

\subsection{Three registers}

The core phenomenological datum is pre-reflective disruption. Around that core, we can distinguish two downstream layers that matter for behavior and stabilization. Together, these three registers map onto the triadic architecture of Chapter~\ref{ch:the-category-zipper}.

First, a qualitative jolt: a pre-reflective sense that something's off, the way a sour note feels sour before you know which note it is. This is the sign-vehicle doing its work~-- the perceivable form registers as wrong before you can say why.

Second, reactive resistance: the hesitation, the repair, the sense that you \emph{can't} say it that way. This is the object\ixsq{object} pushing back~-- the brute constraint on what the form can mean. You feel blocked.

Third, normative correction: the stabilized habit\ixsq{habit} that makes the correction count as a correction. The fact that \mention{I'm twenty years old} is the right repair for \ungram\mention{I have twenty years} isn't arbitrary; it's the interpretant\ixsq{interpretant}~-- the would-be\ixsq{would-be}~-- reasserting itself. Encounter this construction and you \emph{would} get this structural meaning\ixsq{structural meaning}. When the would-be\ixsq{would-be} misfires, the habit\ixsq{habit} demands its due.

When the coupling fails, these registers often co-activate, but they needn't. The readout can be partial and dissociable: jolt without stable correction, correction without strong jolt, or persistent blockage with weak explicit norm talk.

This layering fits the reflexive-equilibrium story in Section~\ref{sec:14:stabilizer}: fast pre-reflective disruption at token-time, repair dynamics in interaction, and slower norm stabilization across communities.

This triadic mapping predicts a specific dissociation already visible in the illusion literature: a sentence can trigger a jolt without a structural violation (garden paths), or pass local fluency checks without global well-formedness (comparative illusions). That's why the next section matters. If the three-register readout is right, we should expect patterned misfires rather than random noise.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}p{0.13\textwidth}p{0.12\textwidth}p{0.21\textwidth}p{0.17\textwidth}p{0.20\textwidth}@{}}
\toprule
\textbf{Register} & \textbf{Timescale} & \textbf{Observable marker} & \textbf{Main confound} & \textbf{Falsifier} \\
\midrule
Qualitative jolt & 100--800 ms & Early disruption signatures (RT, first-pass regressions, EEG perturbation) & High surprisal without structural mismatch & No early disruption above lexical baseline \\
Reactive resistance & 0.5--5 s & Hesitation, restart, and self-repair under mismatch manipulations & General disfluency and task pressure & Repair and hesitation do not increase under controlled mismatch manipulations \\
Normative correction & Turns / years & Preferred repair choices (interactional) and cohort- or register-structured correction norms (sociohistorical) & Schooling or prescriptive monitoring independent of entrenchment & No stable cohort or register structure once exposure is controlled \\
\bottomrule
\end{tabular}
\caption{Operational readout for the three-register proposal.}
\label{tab:14:three-register-operational}
\end{table}

The third register\ixsq{register} therefore names two coupled processes: interactional correction on turn-time scales, and sociohistorical norm stabilization across months and years.

If these measures collapse to one undifferentiated factor, the triadic mapping should be treated as mnemonic rather than mechanistic.

Peirce called the felt disequilibrium that forces inquiry \enquote{doubt}~-- not sceptical doubt, but an irritation that demands repair \citep[CP~5.374]{peirce1931cp}. Grammatical doubt works the same way. The would-be\ixsq{would-be} has misfired; the habit\ixsq{habit} is unsatisfied; the feeling is what unsatisfied habits feel like from the inside. It forces repair not because speakers follow rules consciously, but because the system won't settle until the coupling closes. The immune system\ixsq{immune system}'s first responder.

\bigskip

In sum: Grammaticality\ixsq{grammaticality} itself is the structural property of the form--value\ixsq{value} coupling. What it's \emph{like} is the metacognitive readout of that coupling: usually silent, occasionally jolting, and often corrected in real time. The phenomenology\ixsq{phenomenology} claim stays modest: one core pre-reflective register\ixsq{register} plus reactive and normative downstream layers, tracking sign, object\ixsq{object}, and interpretant\ixsq{interpretant} without replacing the mechanistic account.


\section{Grammaticality illusions}
\label{sec:14:illusions}

The strongest evidence that the feeling is dissociable from the structure comes from \term{grammaticality\ixsq{grammaticality} illusions}~-- cases where the detector\ixsq{detector} misfires.

\subsection{Feels ungrammatical\ixsq{ungrammatical}, is grammatical}

The classic case is the garden-path sentence\ixsq{garden-path sentence}:

\ea[]{\label{ex:14:garden-path}
\mention{The horse raced past the barn fell.}}
\z

This is grammatical~-- a reduced relative clause (\mention{the horse [that was] raced past the barn fell}). But the processing system commits to the main-clause parse (\mention{the horse raced past the barn}) and crashes when \mention{fell} appears. The detector\ixsq{detector} reports ungrammaticality; the structure's fine.

\subsection{Feels grammatical, is ungrammatical\ixsq{ungrammatical}}

The converse is the \term{comparative illusion\ixsq{comparative illusion}}~-- sometimes called the \term{Escher sentence\ixsq{Escher sentence}}:

\ea[]{\label{ex:14:comparative-illusion}
\mention{More people have been to Russia than I have.}}
\z

Each local chunk parses. \mention{More people have been to Russia}~-- fine. \mention{Than I have}~-- fine as a comparative clause. But the whole doesn't compose cleanly: what's being compared? For many speakers, the sentence lacks a stable global interpretation, yet it feels acceptable because each piece satisfies local constraints.

These illusions are exactly analogous to visual illusions. The Müller-Lyer lines are the same length, but the perceptual system reports otherwise. The illusion doesn't show perception's broken; it shows perception is a mechanism with characteristic failure modes. Grammaticality\ixsq{grammaticality} illusions show the same point: detector\ixsq{detector}, not oracle.

\textcite{phillips2011grammatical} develop this point systematically, showing comprehension exhibits \term{selective fallibility\ixsq{selective fallibility}}~-- impressive accuracy for some constraints, and systematic vulnerability for others. The pattern reveals the architecture: a retrieval system that's structure-sensitive when structural cues have temporal priority, and error-prone when they don't.


\section{What doesn't count}
\label{sec:14:negative-space}

The form--value\ixsq{value} coupling's specifically morphosyntactic, and this predicts what won't trigger ungrammaticality. A phonetic error, like \mention{[\ipa{fɪlɔzəfi}]} for \mention{philosophy}, can misfire without being ungrammatical\ixsq{ungrammatical}. A lexical error, like saying \mention{airport} when you meant \mention{train station}, can misfire without being ungrammatical\ixsq{ungrammatical}. A semantic anomaly, like \mention{Colorless green ideas sleep furiously}, can misfire without being ungrammatical\ixsq{ungrammatical}.

These violations trigger different responses. Phonetic errors trigger \mention{what?} because the phoneme\ixsq{phoneme}'s value\ixsq{value}'s the contrast\ixsq{contrast}~-- that's the entire form--value\ixsq{value} coupling at that level, and when it breaks, comprehension stalls. Lexical errors trigger correction or confusion. Semantic anomaly triggers puzzlement. But none of them trigger the specific feeling of ungrammaticality\ixsq{feeling of ungrammaticality}~-- the sense the wiring's broken.

A revealing contrast\ixsq{contrast} comes from the allomorphy of the indefinite article. We reject \ungram{*a apple} and \ungram{*an banana} with the same immune response we bring to syntax. This looks phonological, but the selection rule is entrenched as a grammatical requirement~-- part of the morphosyntactic zipper\ixsq{zipper}. A mispronunciation of \mention{banana} would trigger \mention{what?}; distributing the wrong allomorph of the article triggers the grammar\ixsq{grammar}'s immune response. This is exactly what we'd expect from an HPC\ixsq{HPC}: the edge case reveals the boundary of the morphosyntactic coupling, and the feeling of ungrammaticality\ixsq{feeling of ungrammaticality}~-- if somewhat muted~-- is the trace of the immune system\ixsq{immune system} at work.

\subsection{Value\ixsq{value} at every grain}

What's \enquote{value\ixsq{value}} at each level of the grammar? The zipper\ixsq{zipper} closes differently depending on what you're zipping.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Grain} & \textbf{Form} & \textbf{Value} \\
\midrule
Phoneme & Sound segment & Contrast (minimal-pair distinctiveness) \\
Morpheme & Affix, stem & Grammatical or derivational modification \\
Word & Phonological shape & Lexical meaning + category membership \\
Construction & Syntactic pattern & Argument structure, information structure, meaning template \\
Clause type & Structural configuration & Illocutionary force (statement, question, command) \\
\bottomrule
\end{tabular}
\caption{Form--value coupling at different grains of grammar.}
\label{tab:form-value-grains}
\end{table}

At the phonemic level, value\ixsq{value}'s nothing \emph{but} contrast\ixsq{contrast}. The /p/ in \mention{pat} contributes no semantic content; it just distinguishes the word from \mention{bat}, \mention{cat}, \mention{sat}. When contrast\ixsq{contrast} fails~-- when you can't tell which phoneme\ixsq{phoneme} was intended~-- you say \mention{what?}

At the morphemic level, value\ixsq{value}'s the contribution to word meaning or grammatical function. The \mention{-ed} in \mention{walked} signals past tense; the \mention{un-} in \mention{unhappy} reverses polarity. Break the coupling~-- use \mention{-ed} on a noun, or \mention{un-} on a verb\ixsq{verb} that doesn't license it~-- and ungrammaticality results.

At the word level, value\ixsq{value}'s lexical meaning plus category\ixsq{category} membership. \mention{Dog} contributes dog-ness \emph{and} nounhood\ixsq{nounhood}. Substitute the wrong word and you get confusion; use a word in the wrong category\ixsq{category} slot and you get ungrammaticality.

At the constructional level, value\ixsq{value}'s the structural meaning\ixsq{structural meaning} the pattern contributes. The double-object construction\ixsq{construction} (\mention{gave her the book}) signals transfer plus affected recipient. The \term{passive} (\mention{was eaten}) demotes the agent. Break the construction\ixsq{construction}'s slot-filling requirements and the zipper\ixsq{zipper} won't close.

At the clause-type level, value\ixsq{value}'s illocutionary force. \term{Declarative} structure signals assertion; \term{interrogative} structure signals question; \term{imperative} structure signals command. Mismatch the structure and the force~-- use rising intonation on a declarative to fake a question~-- and you're exploiting the zipper\ixsq{zipper}, not breaking it.

This table helps explain why grammaticality\ixsq{grammaticality} is primarily diagnosed at the morphosyntactic grain~-- the level where form--value\ixsq{value} couplings are tight, obligatory, and enforced.

Phoneticians have their own coupling (\term{contrast\ixsq{contrast}}), and when it breaks, the response's \mention{what?}, not ungrammaticality. Semanticists have theirs (\term{compositionality}, \term{inference}), and when it breaks, the response's puzzlement or absurdity. Grammaticality\ixsq{grammaticality} names the failure mode where the morphosyntactic zipper\ixsq{zipper} won't close. That's why syntax gets first pass on it in this explanatory frame, not sole authority over all appropriateness\ixsq{appropriateness} judgments.

\subsection{Grammaticality\ixsq{grammaticality} as a species of appropriateness\ixsq{appropriateness}}

This grain-specific view suggests a larger picture. Grammaticality\ixsq{grammaticality} isn't sui generis; it's a \emph{species} of appropriateness\ixsq{appropriateness}, narrowed to the morphosyntactic channel.

\term{Appropriateness} in general is the coupling between a form and \emph{all} the values it signals: structural meaning, social indexicality, register stance, identity marking. Grammaticality\ixsq{grammaticality} is what you get when you restrict attention to one channel~-- morphosyntactic form and structural meaning\ixsq{structural meaning}. But morphosyntax isn't the only grain at which form--value couplings are maintained and enforced.

Operationally, a failure counts as grammaticality\ixsq{grammaticality} only if two tests pass. First, the minimal repair that preserves intended message requires a morphosyntactic change, not just lexical substitution, accent shift, or register adjustment. Second, speakers in the relevant discourse community\ixsq{discourse community} show convergent repair preferences across contexts. If either test fails, classify the case under another appropriateness\ixsq{appropriateness} channel.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Domain} & \textbf{Form} & \textbf{Value} \\
\midrule
Grammaticality & Morphosyntax & Structural meaning \\
Register fit & Lexical/syntactic choice & Situational stance \\
Accent & Phonetic realization & Social identity \\
Lexical precision & Word choice & Domain membership, exactness \\
Pragmatic felicity & Utterance in context & Communicative fit \\
\bottomrule
\end{tabular}
\caption{Appropriateness channels: form--value couplings at different grains.}
\label{tab:appropriateness-channels}
\end{table}

Each channel has its own detector\ixsq{detector}, its own failure mode, and its own phenomenology. Grammaticality\ixsq{grammaticality} failure triggers the immune response we've been describing: hesitation, repair, the jolt of \enquote{wrongness}. Register\ixsq{register} mismatch triggers a different response: the raised eyebrow, the red ink on contractions, the sense that something's \enquote{off} without being \emph{ungrammatical\ixsq{ungrammatical}}. The raised eyebrow is a detector\ixsq{detector} too, but it rarely comes with calibration curves or inter-annotator agreement. Accent mismatch triggers identity attribution: \enquote{that's not how we say it here.} Lexical inappropriateness triggers precision failure: \enquote{that's not quite the right word.}

The channels aren't orthogonal. A single utterance is evaluated simultaneously on all of them. And they interact: a register\ixsq{register}-inappropriate form can \emph{feel} ungrammatical\ixsq{ungrammatical} (through prescriptive training or hypercorrection) even when the morphosyntax is impeccable. But they're semi-independent: you can have perfect grammaticality\ixsq{grammaticality} with terrible register\ixsq{register} fit (\mention{I would like to utilize the restroom facilities} is grammatical but register\ixsq{register}-weird in casual speech), and native speakers can have strong grammaticality\ixsq{grammaticality} intuitions with weak register\ixsq{register} intuitions in unfamiliar domains.

Chapter~\ref{ch:social-stabilization} develops this picture further, showing how dialect\ixsq{dialect}, register\ixsq{register}, and discourse community\ixsq{discourse community} act as conditioning variables that select which appropriateness\ixsq{appropriateness} couplings apply in a given context. The academic register\ixsq{register} activates its own appropriateness\ixsq{appropriateness} detector\ixsq{detector} \emph{alongside} the grammaticality\ixsq{grammaticality} detector\ixsq{detector}, not instead of it. They're layered, not competing.

The upshot: grammaticality\ixsq{grammaticality} is appropriateness\ixsq{appropriateness} at the morphosyntactic grain. The asterisk probes one detector\ixsq{detector}; the raised eyebrow probes another. Both are form--value couplings; both are maintained by social mechanisms; both can trigger repair. Grammaticality\ixsq{grammaticality} is the detector\ixsq{detector} syntacticians usually interrogate most directly.


\section{The gradient at the boundary}
\label{sec:14:gradient}

When we judge a sentence, we aren't checking a rulebook. We're interrogating the \term{standing wave\ixsq{standing wave}}~-- probing the system to see if the dynamic equilibrium holds or collapses under strain.

On this model, a prototypically grammatical sentence sits at the bottom of a deep basin\ixsq{basin}. Perturb it slightly~-- change word order, shift stress, substitute a near-synonym~-- and the system tends to correct. A clearly ungrammatical\ixsq{ungrammatical} sentence sits well outside the basin\ixsq{basin}: small perturbations typically don't roll it back to stable ground.

The marginal cases~-- the 3-star and 4-star judgments, the sentences that split informants~-- are poised on the rim. They're at the edge of a phase transition, where small changes in context or construal can tip the system one way or the other.

This reframing has consequences. If grammaticality\ixsq{grammaticality} is basin\ixsq{basin}-stability, then at least part of the apparent sharp boundary between well-formed and ill-formed can be an artifact of how we test. Gradient judgments aren't noise to be smoothed away; they're signal. The gradient isn't a defect in the evidence; it's the evidence under strain.


\section{Grammaticality\ixsq{grammaticality} as mechanism\ixsq{mechanism} and category\ixsq{category}}
\label{sec:14:dual-role}

Grammaticality\ixsq{grammaticality} plays a dual role in the framework. As a \term{category\ixsq{category}}, it's an HPC\ixsq{HPC}~-- the maintained coupling between morphosyntactic form and structural meaning\ixsq{structural meaning}. It sustains a would-be\ixsq{would-be} (one form--value pairing lets you predict others) and it's homeostatic (the feeling of ungrammaticality\ixsq{feeling of ungrammaticality} stabilizes the coupling).

As a \term{mechanism\ixsq{mechanism}}, it's a stabilizer\ixsq{stabilizer} for other HPCs. The category \term{noun} is maintained partly because nounhood\ixsq{nounhood} is grammatical~-- because using a word in a noun slot, with noun morphology, in noun constructions, triggers no mismatch signal. Grammaticality\ixsq{grammaticality}'s the enforcement mechanism\ixsq{mechanism} for the form--value couplings that constitute all other categories. Grammaticality\ixsq{grammaticality} is maintained because it maintains those other couplings.

This dual role~-- maintainer and maintained~-- isn't paradoxical. It's characteristic of homeostatic systems. The immune system\ixsq{immune system}'s both a system (a biological category) and a mechanism that maintains other systems. Grammaticality\ixsq{grammaticality}'s the immune system\ixsq{immune system} of the grammar.

Or think of the \term{spinning top\ixsq{spinning top}} metaphor. A top's both an object and the spin that keeps it upright. Remove the spin and the top falls. Grammaticality\ixsq{grammaticality}'s the spin. It's both the pattern and the force that maintains it. The top keeps spinning because something keeps pushing.

If grammaticality\ixsq{grammaticality}'s a maintained coupling and acceptability\ixsq{acceptability}'s a noisy detector\ixsq{detector}, then the evidential status of the asterisk changes. It's a signal from a measuring instrument~-- informative, but fallible, and calibrated to the community in which it was trained.


\section{Empirical signatures}
\label{sec:14:empirical}

If the feeling of ungrammaticality\ixsq{feeling of ungrammaticality}'s a noisy detector\ixsq{detector}~-- detector, not oracle~-- then laboratory studies should reveal its characteristic failure modes. Three bodies of evidence are consistent with this prediction.

\subsection{Satiation}

\term{Syntactic satiation\ixsq{syntactic satiation}}'s the finding that repeated exposure to a marginally acceptable sentence can improve its perceived acceptability\ixsq{acceptability} \parencite{snyder2000grammaticality}. But the effect's selective. Some constraints are susceptible to frequency-based updating: repeated exposure makes them feel better. Others resist satiation entirely. In current English judgment datasets, left-branch extraction (\ungram\mention{Which did you buy car?}) remains near-categorically rejected even under exposure manipulations \parencite{Snyder2022}. The intended meaning's easily grasped~-- \mention{Which car did you buy?}~-- and the structure isn't unusually complex. The resistance has appeared across multiple tasks and speaker samples, though not uniformly in every paradigm.

Why the split? If the feeling of ungrammaticality\ixsq{feeling of ungrammaticality} arises from comparing input against entrenched form--value pairings, then repeated exposure should matter~-- but only for constraints that are frequency-sensitive in the first place. When you hear a marginal sentence ten times, you're adding exemplars to the distribution; the pairing becomes more familiar, the mismatch signal weakens, and the sentence feels better. 

But why does the mismatch signal weaken? It's not just that the existing category gets wider; the category itself moves. \textcite{vagninoWalker2026} term this \term{schema drift\ixsq{schema drift}}. In their study of relational concepts, they found that repeated exposure to diverse exemplars caused the abstract schema itself to drift, accommodating new forms. They explicitly contrast this with the standard view of \term{rerepresentation\ixsq{rerepresentation}} (changing the input to fit the category\ixsq{category}). As they put it: \enquote{instead of changing the square peg to better fit the round hole... the hole might also be altered to better fit the peg.}

Syntactic satiation\ixsq{syntactic satiation} may be one route by which schema drift\ixsq{schema drift} appears in real time. The \enquote{hole} (the grammatical category\ixsq{category}) deforms to accommodate the repeated \enquote{peg} (the marginal sentence).

But when a constraint behaves as near-categorical in current English evidence~-- as left-branch extraction appears to~-- repeated brief exposure has shown little or no shift. The resistance doesn't imply the structure is unimaginable; it implies the learner has detected a \term{conspicuous absence\ixsq{conspicuous absence}}. Fronting is productive in English; the *potential* for extraction is visible. But in this specific configuration, the community systematically avoids it (a conspicuous low-base-rate gap), using entrenched \term{workarounds\ixsq{workarounds}} instead: pied-piping (\mention{whose book}), fused-head constructions (\mention{whose is this}), and the \mention{big mess} construction (\mention{how good a teacher}). \textcite{reynolds2026lbe} argues that this density of alternatives can make the gap resistant to short-run satiation because competitor pathways are strongly activated.

The gap persists because the potential is there but the practice is absent. Learners infer that the community has tacitly agreed *not* to use the available mechanism in this way. The detector\ixsq{detector} isn't just saying \enquote{this doesn't happen}; it's saying \enquote{we could do this, but we don't.} The feeling of ungrammaticality\ixsq{feeling of ungrammaticality} tracks this specific, entrenched abstinence.

\subsection{Individual differences}

Speakers differ systematically in their grammaticality\ixsq{grammaticality} thresholds. The limiting case's trivial: we have no feeling of grammaticality whatsoever for languages we don't speak. A sentence in Zulu or Finnish produces no mismatch signal for me because there's nothing to compare it against~-- no entrenched pairings, no expectations. Foreign language learners occupy the middle ground: their feelings are degraded, unreliable, slow to fire, and easily fooled by surface plausibility. Only with sufficient exposure does the feeling sharpen.

Within a single language community, the variation persists. \textcite{dabrowska2012} shows that grammatical knowledge isn't uniform across the adult population: speakers with less formal education, less exposure to complex syntactic constructions, and less metalinguistic training produce reliably different acceptability\ixsq{acceptability} judgments~-- not just noise, but systematic divergence.

That's the prediction from community-indexed entrenchment\ixsq{entrenchment}. If grammaticality\ixsq{grammaticality} is what a community has entrenched, then members with different exposure histories will have different thresholds. The variation isn't random; it tracks social and educational ecologies. The \enquote{grammar\ixsq{grammar}}'s not a Platonic object that speakers approximate; it's a statistical regularity that speakers instantiate to varying degrees.

\subsection{Processing difficulty and beyond}

Processing difficulty isn't grammaticality\ixsq{grammaticality}, but the two are often conflated. \textcite{phillips2011grammatical} show that comprehension exhibits \term{selective fallibility\ixsq{selective fallibility}}: it's exquisitely sensitive to some constraints and systematically vulnerable to others. The pattern reveals something about the architecture of the feeling, not just noise.

Consider \term{garden-path sentences}. When readers encounter a grammatical sentence that's initially misparsed, they produce high unacceptability ratings at the point of reanalysis, even though the sentence's well-formed. The feeling of ungrammaticality\ixsq{feeling of ungrammaticality} fires because processing cost correlates, in normal use, with form--value\ixsq{value} mismatch. 

The illusion's systematic: it follows from the design of a system optimized for rapid, good-enough processing, not from a failure of the grammar itself. These three signatures~-- satiation, individual differences, and processing-based illusions~-- converge on the same conclusion. The feeling of ungrammaticality\ixsq{feeling of ungrammaticality}'s noisy, not oracular. It tracks the coupling between form and value\ixsq{value}, but it does so imperfectly, shaped by exposure and processing constraints. The asterisk's informative, but it's not infallible.


\subsection{Relevance and the immune system\ixsq{immune system}}
\label{sec:14:relevance}

\textcite{scottphillips2024communication} offers a complementary account. On his view, grammaticality\ixsq{grammaticality} intuitions are byproducts of relevance-tracking: we detect when utterances violate basic presumptions about communicative efficiency, just as we detect when visual stimuli violate assumptions about physical objects. Unacceptability arises not from mere inefficiency, but from an inherent impossibility of interpreting an utterance consistently with those presumptions.

The MMMG framework shares several premises with this account. Both reject the need for an innate grammar faculty; both locate intuitions within general cognitive systems; both recognize that communicative pressures shape linguistic conventions. The frameworks differ in mechanism and prediction. Relevance theory locates failure in inferential impossibility; MMMG locates failure in morphosyntactic form--value\ixsq{value} mismatch~-- the zipper\ixsq{zipper} that won't close.

The difference has empirical consequences. Relevance theory predicts broader failures when relevance presumptions collapse; MMMG predicts construction\ixsq{construction}-specific failures where entrenched pairings misalign. On the MMMG view, \ungram\mention{I have twenty years} fails as an age statement because the English age construction's \mention{be} + \mention{years old}, not because relevance-seeking fails in general. Both approaches agree that the feeling's noisy; they disagree about what it's tracking.

The accounts may ultimately converge. If efficient interpretation requires stable form--meaning pairings, then relevance-tracking and form--value\ixsq{value} coupling are two descriptions of the same constraint. What MMMG adds is granularity: it identifies the specific level (morphosyntax) and the specific mechanism (entrenchment\ixsq{entrenchment}) that make the coupling possible.


\section{The HPC\ixsq{HPC} audit}
\label{sec:14:hpc-audit}

Chapter~\ref{ch:failure-modes} introduced the Two-Diagnostic Test\ixsq{two-diagnostic test} for genuine HPC kinds: high projectibility\ixsq{projectibility} and robust homeostasis\ixsq{homeostasis}. On the current evidence, grammaticality\ixsq{grammaticality} appears to pass both.

First, \term{projectibility\ixsq{projectibility}}. Grammaticality\ixsq{grammaticality} sustains a would-be\ixsq{would-be}: encounter a form--value\ixsq{value} pairing of this type and you \emph{would} expect similar pairings to hold. If \mention{the dog bit \emph{him}}'s grammatical, then \mention{the cat scratched \emph{her}} would be too. This isn't because speakers have memorized every possible sentence; it's because they've abstracted the construction\ixsq{construction} and the would-be extends to instances not yet encountered.

Second, \term{homeostasis\ixsq{homeostasis}}. The feeling of ungrammaticality\ixsq{feeling of ungrammaticality} stabilizes the coupling, correcting deviations. Speakers hesitate, repair, rephrase. Hearers ask for clarification, or offer corrections. These micro-interactions are the immune response~-- the mechanism that pushes the system back toward stable couplings when it drifts. Grammaticality\ixsq{grammaticality}'s homeostatic because the feeling's a stabilizer\ixsq{stabilizer}.

Grammaticality\ixsq{grammaticality} isn't merely a binary switch, nor a primitive of the theory. It's an HPC\ixsq{HPC} kind in its own right~-- maintained by the same mechanisms that maintain countability\ixsq{countability}, definiteness, gender, and lexical categories. On current evidence, it meets the two diagnostics provisionally enough to treat it as a category in this framework.


\bigskip
\noindent\textsc{Definiteness\ixsq{definiteness} thread.} This chapter adds the final filter to the definiteness diagnosis: the morphosyntactic \enquote{zipper\ixsq{zipper}} rejection of infelicitous forms. In the relevant English constructions, the coupling between semantic definiteness\ixsq{definiteness} and morphosyntactic deitality\ixsq{deitality} behaves as a grammatical requirement, not merely a style preference.
\ea[]{\label{ex:14:definiteness-thread}
\ungram\mention{a the book}}
\z
The immune response that rejects this determiner stacking is closely related to the one that rejects \ungram\mention{I have twenty years} for age.
\textit{Scoped prediction.} In populations where a novel determiner sequence becomes genuinely entrenched, mismatch judgments should shift in cohort- and register\ixsq{register}-structured ways: high-exposure speakers and informal registers should move first, with formal edited norms lagging.
\textit{Falsifier.} If durable entrenchment\ixsq{entrenchment} of a novel sequence produces no cohort- or register-structured change in mismatch judgments, the maintenance account is wrong for this domain.


\section{Looking forward}
\label{sec:14:looking-forward}

We began with an asterisk, and we end with the same question: what does the asterisk diagnose?

Grammaticality\ixsq{grammaticality}'s not a primitive. It's not a direct readout of structure. It's not possibility versus impossibility, but coupling versus mismatch: form and value\ixsq{value}, zipper\ixsq{zipper} teeth that mesh or don't. The feeling that signals mismatch is a noisy instrument~-- detector\ixsq{detector}, not oracle~-- shaped by processing constraints and what a community has entrenched.

This reframing has consequences. If grammaticality\ixsq{grammaticality} is a category, then we can ask: how projectible is it? how robust is its homeostasis\ixsq{homeostasis}? And if grammaticality\ixsq{grammaticality} is a mechanism~-- the immune system\ixsq{immune system} of the grammar~-- then we can ask: what maintains it? what degrades it? what illusions does it produce?

The dual role~-- maintainer and maintained~-- isn't paradoxical. It's characteristic of homeostatic systems. The spinning top\ixsq{spinning top} is both the object and the spin that keeps it upright. Grammaticality\ixsq{grammaticality} is the spin.

This means the equilibrium is \term{contingent\ixsq{contingent}}. On strong constraint-based readings, LBE \emph{can't} emerge in English~-- a grammatical principle blocks it. The HPC\ixsq{HPC} account says only that it \emph{hasn't} under current social and statistical conditions. But imagine a high-status speaker~-- or a low-status one with memetic cachet like Homer Simpson~-- introducing \mention{Whose did you eat sandwich?} to the language. Use the form, get the laugh, signal the group membership. Homer doesn't just provide new evidence; he spawns a new discourse community\ixsq{discourse community}. If social uptake shifts, priors shift. On this view, what resists at a given moment is entrenched convention, not a timeless prohibition. The mechanism is maintenance, not law.

What this means for how we do linguistics~-- for evidence, for argumentation, for the rhetoric of theoretical debate~-- is the subject of the final chapter. The asterisk is a probe into the standing wave\ixsq{standing wave}. Chapter~\ref{ch:what-changes} asks what happens when we take that probe seriously.
