% Back cover page for inclusion in main book
% Assumes CoverWhite already defined by coverpage.tex

\definecolor{CoverSlate}{HTML}{2A3F54}   % deep slate from river channels
\definecolor{CoverGold}{HTML}{B89A5A}    % muted gold from sediment banks
\providecolor{CoverWhite}{HTML}{F6F4EF}

\thispagestyle{empty}
\pagecolor{CoverSlate}
\color{CoverWhite}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0pt}

\vspace*{4mm}

\begin{center}
\begin{minipage}{0.84\textwidth}
\color{CoverWhite}

% --- Blurb ---
{\fontsize{9.6pt}{13.2pt}\selectfont
Linguists assume their categories have determinate boundaries.
They don't~-- and the usual retreat to gradience leaves stability
unexplained.\par
\vspace{4pt}
This book offers a third option. Many linguistic categories are
homeostatic property-cluster kinds, held together not by
definitions but by mechanisms. Acquisition, entrenchment,
alignment, and transmission keep properties bundled. No single
	property is necessary; the edges can be hard to locate exactly; but the
clustering is real because the mechanisms are real.\par
\vspace{4pt}
The framework applies from phoneme contrasts through grammatical
categories to constructions, resolving long-standing disputes
about definiteness, countability, word classes, and
grammaticality itself. Projectibility turns out to be
field-relative: different subfields track different profiles of
the same causal system. Getting the metaphysics right changes
what questions we ask and what answers we accept.\par}

\vspace{6pt}
{\centering\textcolor{CoverGold}{\rule{0.88\textwidth}{0.35pt}}\par}
\vspace{6pt}

% --- Endorsement 1 ---
{\centering\fontsize{8.8pt}{12pt}\selectfont
\textit{\enquote{Endorsement placeholder. A compelling quote about the
book's contribution to linguistic theory and its clarity of
argument.}}\par
\vspace{2pt}
{\fontsize{8pt}{10pt}\selectfont
 \addfontfeatures{LetterSpace=1.8}\scshape
 --- Reviewer Name, University}\par}

\vspace{7pt}

% --- Endorsement 2 ---
{\centering\fontsize{8.8pt}{12pt}\selectfont
\textit{\enquote{Endorsement placeholder. A second quote
emphasising the book's originality and its relevance to working
linguists.}}\par
\vspace{2pt}
{\fontsize{8pt}{10pt}\selectfont
 \addfontfeatures{LetterSpace=1.8}\scshape
 --- Reviewer Name, University}\par}

\vspace{6pt}
{\centering\textcolor{CoverGold}{\rule{0.88\textwidth}{0.35pt}}\par}
\vspace{6pt}

% --- Author bio ---
{\fontsize{8.2pt}{11.2pt}\selectfont\color{CoverWhite!72}%
\textsc{Brett Reynolds} is [author bio placeholder~-- position,
institution, research interests, and previous publications].\par}

\end{minipage}
\end{center}

\vfill

% --- Bottom bar: publisher left, barcode right ---
\noindent\hspace{10mm}%
\begin{minipage}[b]{0.40\textwidth}
{\fontsize{7.4pt}{9.6pt}\selectfont\color{CoverWhite!55}%
Publisher Name\\Publisher Location}
\end{minipage}%
\hfill
\begin{minipage}[b]{0.40\textwidth}
\raggedleft
\colorbox{white}{\parbox[c][19mm][c]{35mm}{%
  \centering\fontsize{7pt}{9pt}\selectfont\color{black}%
  ISBN 000-0-00-000000-0\\[6pt]
  \textsc{barcode}}}
\end{minipage}\hspace{10mm}%

\vspace*{5mm}
